---
title: "02 - Singapore Bus Passenger Travel Patterns"
author: "Cathy C."
date: "Oct 7, 2025"
date-modified: "last-modified" #auto render date
format: html
execute: #how YML page work
  echo: true #will show code
  eval: true #will run code every time and show output
  warning: false #dont show warning msg
  freeze: true #controls rendering - no render if no changes
---

## 1 Overview

## 2 Data sources & preparation

### 2.1 Load packages

The packages that we will use for this study are outlined below:

| R Package | Description |
|----------------------|--------------------------------------------------|
| **sf** | A package that provides [**simple features access**](https://en.wikipedia.org/wiki/Simple_Features) for R. |
| **spdep** | Creates spatial weights matrices and to test for and model spatial autocorrelation in geographical data, like Moran's I and Geary's C and Local Moran's I and Getis-Ord G. |
| **tidyverse** | Includes the core packages like ggplot2, dplyr, readr, purrr, tibble, lubridate, etc. |
| **sfdep** | Builds on the great shoulders of **spdep** package for spatial dependence. sfdep creates an sf and tidyverse friendly interface to the package as well as introduces new functionality that is not present in spdep. |
| **h3** | H3 is a geospatial indexing system using a hexagonal grid that can be (approximately) subdivided into finer and finer hexagonal grids, combining the benefits of a hexagonal grid with [S2](https://code.google.com/archive/p/s2-geometry-library/)'s hierarchical subdivisions. |
| **knitr** | Color maps designed to improve graph readability for readers with common forms of color blindness and/or color vision deficiency. |
| **tmap** | geographical maps in which spatial data distributions are visualized. This package offers a flexible, layer-based, and easy to use approach to create thematic maps, such as choropleths and bubble maps. |
| **patchwork** |  |
| **rvest**| Provides the HTML parsing and extraction functions |
| **rsvg** | |

Let's install and load the following packages: **tidyverse, sf, sfdep, tmap, knitr** in R environment.

```{r}
pacman::p_load(tidyverse, sf, spdep, sfdep, tmap, knitr, h3, patchwork, rvest, RColorBrewer)
```

### 2.2 Import data

We will import 3 data sets for this study:

1.  Import *Bus Stop Location* shapfile from LTA DataMall into R environment
2.  Import *Master Plan 2019 Subzone Boundary (No Sea)* from <https://data.gov.sg/>
3.  Import *Passenger Volume by Origin Destination Bus Stops* from LTA DataMall
4.  Import *LTA MRT Station Exit (KML)* from <https://data.gov.sg/>

When importing data, need to check first. Make sure busstop code is in character form rather than numeric. If it's latter, need to insert 0 because the stop code can start with zero.

Open Bustop.prj and check if the data has the information we need, or need to transform. Below we will start importing the data.

::: panel-tabset
## **Master Plan Subzone Boundary**

Below we will import *Master Plan 2019 Subzone Boundary (No Sea).*

```{r}
#| eval: false
mpsz <- st_read("data/MasterPlan2019SubzoneBoundaryNoSeaKML.kml")
```

```{r}
mpsz <- read_rds("data/rds/mpsz.rds")
st_crs(mpsz)
```

## **Bus Stop Location**

Below we will import *Bus Stop Location* shapfile from LTA DataMall.

```{r}
BusStop = st_read(dsn = "data/LTADataMall",
                  layer = "BusStop")
```

Check the data type:

```{r}
st_crs(BusStop)
```

Based on the above, we identify that the the EPSG is not aligned with mpsz, so we need to transform the data for EPSG:3414 Projected coordinate system for Singapore.

```{r}
BusStop <- BusStop %>%
  st_transform(crs = 3414)
```

Check again on BusStop:

```{r}
st_crs(BusStop)
```

Now, we see BusStop is aligned with mpsz.

## **Passenger Volume**

Below code is used to import *Passenger Volume by Origin Destination Bus Stops* from LTA DataMall.

```{r}
#| eval: false
odbus_08 <- read_csv("data/origin_destination_bus_202508.csv")
odbus_09 <- read_csv("data/origin_destination_bus_202509.csv")
odbus_07 <- read_csv("data/origin_destination_bus_202507.csv")
odbus <- rbind(odbus_08, odbus_09, odbus_07)
write_csv(odbus, "data/odbus.csv")
```

```{r}
odbus <- read_csv("data/odbus.csv")
```

```{r}
st_crs(odbus)
```

describe the coordinate system EPSG, SF

## **MRT Station**
Below we will import **Master Plan 2019 Subzone Boundary (No Sea).**
```{r}
mrt <- st_read("data/LTAMRTStationExitKML.kml") %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 3414)
```
Below code is to write MRT station data into a rsd file, so we can directly call it without processing the data set again.
```{r}
#| eval: false
extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)
  rows <- page %>% html_elements("tr")
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>%
    html_element("td") %>%
    html_text2()
  
  if (length(value) == 0) NA_character_ else value
}

mrt_sf <- mrt %>%
  mutate(
    STATION = map_chr(Description, extract_kml_field, "STATION_NA"),
    EXIT_CODE = map_chr(Description, extract_kml_field, "EXIT_CODE"),
    INC_CRC = map_chr(Description, extract_kml_field, "INC_CRC"),
    FMEL_UPD_D = map_chr(Description, extract_kml_field, "FMEL_UPD_D")
  ) %>%
  
  # Just one exit for each station to avoid multiple points.
  filter(EXIT_CODE == "Exit A") %>%
  
  select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())


write_rds(mrt_sf, "data/mrt_sf.rds")
```


```{r}
mrt_sf <- read_rds("data/mrt_sf.rds")
```


Plot the MRT station exits in a tmap. 
```{r}
tmap_mode("view")
tm_shape(mpsz) +
  tm_polygons(fill = "#A1C2BD", filll_alpha = 0.3) +
  tm_text("SUBZONE_N", size = 0.4) +
tm_shape(mrt_sf) +
  tm_symbols(
   fill = "#EDA35A", size = 0.4,
   col = "#D9E9CF",
#   popup = "STATION"
  ) +
  tm_layout(
    main.title = "MRT Station Exits",
    legend.show = FALSE    # Hide the legend since there are many unique stations
  )

```

:::

### 2.3 Data wrangling: Prepare hexagon data

#### 2.3.1 Extracting bus stops in Singapore

First we plot the bus stops in Singapore below:

```{r}
#| fig-width: 12
#| fig-height: 8

tm_shape(mpsz) +
  tm_polygons() +
  tm_shape(BusStop) +
  tm_dots(size = 0.3, fill = "#DA6C6C", col = "#F49BAB") +
  tm_layout(bg.color = "#D0DDD0",
            outer.margins = c(0.01, 0.01, 0.01, 0.01))
```

We see that the dots (bus stops) outside the boundaries are buses for JB, Malaysia. Therefore, **these have to be excluded.** One way to exclude them is to plot the mpsz layer on top of the BusStop layer.

If we have mpsz in front of the layer, it will capture all bus stops. When we plot first layer using Busstop before mpsz, these spots for JB will be covered.

```{r}
BusStop_in_SG <- st_join(
  BusStop, mpsz, #BusStop first
  join = st_within,
  left = FALSE
)
```

The, we plot again:

```{r}
#| fig-width: 12
#| fig-height: 8

tm_shape(mpsz) +
  tm_polygons() +
  tm_shape(BusStop_in_SG) +
  tm_dots(size = 0.3, fill = "#DA6C6C", col = "#F49BAB") +
  tm_layout(bg.color = "#D0DDD0",
            outer.margins = c(0.01, 0.01, 0.01, 0.01))
```

#### 2.3.2 Deriving Analytical Hexagon

We want to create regular shapes layers with hexagon grid (square and hexagon are 2 most used), and exclude those without bus stops. Create the hexagon layer covering the entire Singapore.

We will need to derive the hexagon data of **375m** (this distance is the perpendicular distance between the centre of the hexagon and its edges) to represent the [traffic analysis zone (TAZ)](https://tmg.utoronto.ca/files/Reports/Traffic-Zone-Guidance_March-2021_Final.pdf). `cellsize` is the distance between opposite edges of the hexagon. This distance is twice the perpendicular distance from the center to any edge.

```{r}
hexagon <- st_make_grid(mpsz, cellsize = 750,
                        what = "polygon",
                        square = FALSE) %>% 
# not a tibbler data frame
  st_sf() # so need to convert to sf
```

The output does not have a unique ID.

#### 2.3.3 Filtering hexagons with bus stops

We only want to retain hexagons with bus stops. Use `st_intersects()` to return 2 values: True, False (0 or 1). The hexagon with a bus stop would be 1, and without a bus stop would return 0.

```{r}
hexagon$n_collisions = 
  lengths(st_intersects(
    hexagon, BusStop_in_SG
  )) #n_collisions show the number of bus stops.

hexagon <- filter(hexagon, n_collisions >0) #retain all that have bus stops.
```

#### 2.3.4 Visualise the output

```{r}
#| fig-width: 12
#| fig-height: 8
tm_shape(mpsz) +
  tm_polygons() +
  tm_shape(hexagon) +
  tm_fill(col = "grey20", fill = "#FFE8CD", fill_alpha = 0.7) +
  tm_layout(bg.color = "#D0DDD0",
            outer.margins = c(0.01, 0.01, 0.01, 0.01))
```

%%We want to know the trips occurred in each hexagon in certain time periods.%%

The hexagons do not have an ID, so we will give each an ID.

#### 2.3.5 Assign IDs to each hexagon

```{r}
hexagon <- hexagon %>%
  select(, -n_collisions)

hexagon$HEX_ID <- sprintf(
  "H%04d", seq_len(nrow(hexagon))
) %>%
  as.factor() #they will in ordinal scale

kable(head(hexagon))
```

### 2.4 Prepare trip generation data

#### 2.4.1 Clearing the data

We don't need the destination code. Below we will select the columns we need, including `ORIGIN_PT_CODE`, `DAY_TYPE`, `TIME_PER_HOUR`, and `TOTAL_TRIPS`. We also simplify the code to join the datasets together later.

```{r}
trips <- odbus %>%
  select(c(YEAR_MONTH, ORIGIN_PT_CODE, DAY_TYPE, TIME_PER_HOUR, TOTAL_TRIPS)) %>%
  rename(BUS_STOP_N = ORIGIN_PT_CODE) %>% #simply the code for joining.
  rename(HOUR_OF_DAY = TIME_PER_HOUR) %>%
  rename(TRIPS = TOTAL_TRIPS)
trips$BUS_STOP_N <- as.factor(trips$BUS_STOP_N) #convert to factor so it's not numerical.
kable(head(trips))
```

**Combine bus stop with Hex ID into hexagon layer.**

```{r}
bs_hex <- st_intersection(
  BusStop, hexagon
) %>%
  st_drop_geometry() %>%
  select(c(BUS_STOP_N, HEX_ID)) #drop away the rest
kable(head(bs_hex))
```

#### 2.4.2 Add HEX_ID into bus trips data

**Use inner_join() to join trips and bus_hex. - Count trips per hexagon.**

```{r}
trips <- inner_join(trips, bs_hex)
kable(head(trips))
```

Let's plot the trips in a bar plot:

```{r}
trips_sort <- trips %>%
  arrange(desc(TRIPS)) %>%
  mutate(HEX_ID = factor(HEX_ID, levels = unique(HEX_ID)))
```

```{r}
ggplot(trips_sort,
  aes(x = HEX_ID, y = TRIPS, fill = HEX_ID)) +
  geom_col(fill = "blue") +
  labs(title = "Total Trips by HexID",
       x = "Hex ID", y = "Number of Trips") + 
  theme_minimal()
```

------------------------------------------------------------------------

When we sort the number of trips by Hex ID, we see \*\*\*

### 2.5 Time preriod filtering

```{r}
glimpse(trips_sort)

```


## 3 Exploratory spatial analysis

In section 2.3, we derived an analytical hexagon data of 375m to represent the traffic analysis zone (TAZ). We also processed time period filtering for 3 time intervals to compute passenger trips generated by origin at hexagon level. Below are the 3 time intervals.

| Peak hour period       | Bus tap on time |
|------------------------|-----------------|
| Weekday morning peak   | 6am - 9am       |
| Weekday afternoon peak | 5pm - 8pm       |
| Weekend/holiday        | 11am - 8pm      |

### 3.1 Peak Hour Trips by HexID
Below charts are plotted to have a full picture of the volumn of each Hex ID in a descending manner. 

::: panel-tabset

## Weekday AM

In this section, we want to filter the Bus tap on the time between 6am to 9am as morning peak hours during weekdays.

```{r}
AM_weekday_sort <- trips_sort %>%
  filter(HOUR_OF_DAY %in% c(6, 7, 8, 9),
         DAY_TYPE == "WEEKDAY") %>%
  group_by(HEX_ID) %>%
  summarise(AM_TRIPS = sum(TRIPS)) %>%
  arrange(desc(AM_TRIPS))
```

Next let's plot it out.

```{r}
ggplot(AM_weekday_sort, 
       aes(x = reorder(HEX_ID, -AM_TRIPS), y = AM_TRIPS)) +
  geom_col(fill = "#B8C4A9") +
  labs(title = "Weekday Morning Peak Hour Trips (6-9 AM) by HexID",
       x = "Hex ID", y = "Number of Trips") + 
  theme(panel.background = element_rect(fill = "#D0DDD0"))
```

## Weekday PM

In this section, we want to filter the Bus tap on the time between 5pm to 8pm as afternoon peak hours during weekdays.

```{r}
PM_weekday_sort <- trips_sort %>%
  filter(HOUR_OF_DAY %in% c(17, 18, 19, 20),
         DAY_TYPE == "WEEKDAY") %>%
  group_by(HEX_ID) %>%
  summarise(PM_TRIPS = sum(TRIPS)) %>%
  arrange(desc(PM_TRIPS))
```

Next let's plot it out.
```{r}
ggplot(PM_weekday_sort, 
       aes(x = reorder(HEX_ID, -PM_TRIPS), y = PM_TRIPS)) +
  geom_col(fill = "#F2AEBB") +
  labs(title = "Weekday Afternoon Peak Hour Trips (5-8 PM) by HexID",
       x = "Hex ID", y = "Number of Trips") + 
  theme(panel.background = element_rect(fill = "#D0DDD0"))

```

## Weekend/PH

In this section, we want to filter the Bus tap on the time between 11am to 8pm as morning peak hours during weekend or holidays.

```{r}
weekend_sort <- trips_sort %>%
  filter(HOUR_OF_DAY %in% c(11:20),
         DAY_TYPE == "WEEKENDS/HOLIDAY") %>%
  group_by(HEX_ID) %>%
  summarise(WKD_TRIPS = sum(TRIPS)) %>%
  arrange(desc(WKD_TRIPS))
```

Next let's plot it out.
```{r}
ggplot(weekend_sort, 
       aes(x = reorder(HEX_ID, -WKD_TRIPS), y = WKD_TRIPS)) +
  geom_col(fill = "#A376A2") +
  labs(title = "Weekend/Holiday Trips (5-8 PM) by HexID",
       x = "Hex ID", y = "Number of Trips") + 
  theme(panel.background = element_rect(fill = "#D0DDD0"))

```

:::

### 3.1 Peak Hour Trips by HexID
Below data are sorted using Hex ID (instead of sorted number of trips) to plot different peak periods. We will use this data to join back to hexagon, so we can plot the passenger volume in the thematic maps with hexagons.

```{r}
#| code-fold: true
# AM weekdays sorted by Hex ID
AM_weekday <- trips %>%
  filter(HOUR_OF_DAY %in% c(6, 7, 8, 9),
         DAY_TYPE == "WEEKDAY") %>%
  group_by(HEX_ID) %>%
  summarise(AM_TRIPS = sum(TRIPS)) %>%
  arrange(as.character(HEX_ID))

# PM weekdays sorted by Hex ID
PM_weekday <- trips %>%
  filter(HOUR_OF_DAY %in% c(17, 18, 19, 20),
         DAY_TYPE == "WEEKDAY") %>%
  group_by(HEX_ID) %>%
  summarise(PM_TRIPS = sum(TRIPS)) %>%
  arrange(as.character(HEX_ID))

# Weekend sorted by Hex ID
weekend <- trips %>%
  filter(HOUR_OF_DAY %in% c(11:20),
         DAY_TYPE == "WEEKENDS/HOLIDAY") %>%
  group_by(HEX_ID) %>%
  summarise(WKD_TRIPS = sum(TRIPS)) %>%
  arrange(as.character(HEX_ID))

```

Below we plot out all peak hours trips per their Hex ID to have a full picture of the volumne of the passengers.

```{r}
#| fig-width: 12
#| fig-height: 40
#| code-fold: true

max_trips <- max(c(max(weekend$WKD_TRIPS), 
                   max(AM_weekday$AM_TRIPS), 
                   max(PM_weekday$PM_TRIPS)))
#weekend plot - adjust scale
wknd_bar <- ggplot(weekend, 
       aes(x = HEX_ID, y = WKD_TRIPS)) +
  geom_col(fill = "#A376A2") +
  labs(title = "Weekend/Holiday Trips",
       x = "Hex ID", y = "Number of Trips") + 
  theme(panel.background = element_rect(fill = "#D0DDD0")) +
  scale_y_continuous(limits = c(0, max_trips)) +
  coord_flip()

#AM plot - adjust scale
am_bar <- ggplot(AM_weekday, 
       aes(x = HEX_ID, y = AM_TRIPS)) +
  geom_col(fill = "#B8C4A9") +
  labs(title = "AM Trips",
       x = "Hex ID", y = "Number of Trips") + 
  theme(panel.background = element_rect(fill = "#D0DDD0")) +
  scale_y_continuous(limits = c(0, max_trips)) +  
  coord_flip()

#PM plot - adjust scale
pm_bar <- ggplot(PM_weekday, 
       aes(x = HEX_ID, y = PM_TRIPS)) +
  geom_col(fill = "#F2AEBB") +
  labs(title = "PM Trips",
       x = "Hex ID", y = "Number of Trips") + 
  theme(panel.background = element_rect(fill = "#D0DDD0")) +
  scale_y_continuous(limits = c(0, max_trips)) +
  coord_flip()



combined_plot <- am_bar + pm_bar + wknd_bar
combined_plot
```
### 3.2 Interactive Maps for Peak Hour Trips
Now let's join the data grouped by HEX_ID with `hexagon`, so we can plot it in tmap later for the three different peak hours of the week.

```{r}
# Join all three datasets
hexagon_all <- hexagon %>%
  left_join(AM_weekday, by = "HEX_ID") %>%
  left_join(PM_weekday, by = "HEX_ID") %>%
  left_join(weekend, by = "HEX_ID") %>%
  mutate(across(c(AM_TRIPS, PM_TRIPS, WKD_TRIPS), 
                ~replace_na(., 0))) # Handle any NA values - we fill in zero for hexagons with no trips

# Check if join worked
names(hexagon_all)  
head(hexagon_all$PM_TRIPS)  
head(hexagon_all$AM_TRIPS)  
head(hexagon_all$WKD_TRIPS)  

```

Below plots demonstrate the geographical distribution of the passenger trips in the 3 peak hour periods. Each page plots one of the peak hour periods. Note that there are two maps, and one plots the actual number of trips while the other has the squar root number of the trips to normalise the data. This is because some hexagons are outliers, and using square root numbers could better help us observe geographical distribution of passenger trips.

::: panel-tabset
## AM weekdays

```{r}
#| fig-width: 12
#| align: center
#| code-fold: true

#tmap_mode("view")
am_tm <- tm_shape(mpsz) +
  tm_polygons(fill = "#FFF4EA", alpha = 0.3) +
  tm_text("SUBZONE_N", size = 0.5) +
tm_shape(hexagon_all) +
  tm_fill("AM_TRIPS", palette = "-Spectral", fill_alpha = 0.5) +
  tm_borders(col = "#B3C8CF", lwd = 0.1) +
  tm_layout(bg.color = "#D0DDD0",
            outer.margins = c(0.01, 0.01, 0.01, 0.01),
            main.title = "AM Peak Hour Trips by Hexagon",
            legend.position = c("right", "bottom"),
            frame = FALSE) +
    tm_view(set_zoom_limits = c(11,18))

# Add sqr-transformed column (not using)
hexagon_all <- hexagon_all %>%
  mutate(AM_TRIPS_SQRT = sqrt(AM_TRIPS))

#tmap_mode("view")
am_tm_qt <- tm_shape(mpsz) +
  tm_polygons(fill = "#FFF4EA", alpha = 0.3) +
  tm_text("SUBZONE_N", size = 0.5) +
  tm_shape(hexagon_all) +
  tm_fill("AM_TRIPS", palette = "-Spectral", fill_alpha = 0.6,
          style = "quantile") +
  tm_borders(col = "#B3C8CF", lwd = 0.1) +
  tm_layout(bg.color = "#D0DDD0",
            outer.margins = c(0.01, 0.01, 0.01, 0.01),
            main.title = "AM Peak Hour Trips (Quantile)",
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_shape(mrt_sf) +
  tm_symbols(   fill = "#CA8787", size = 0.5, col = "#D9E9CF"
  ) +
  tm_view(set_zoom_limits = c(11,18)) 

tmap_arrange(am_tm, am_tm_qt, 
             asp=1, ncol=1,
             sync = TRUE)
tmap_mode("plot")
```

## PM weekdays

```{r}
#| fig-width: 12
#| fig-height: 7
#| align: center
#| code-fold: true

#tmap_mode("view")
pm_tm <- tm_shape(mpsz) +
  tm_polygons(fill = "#FFF4EA", alpha = 0.3) +
  tm_text("SUBZONE_N", size = 0.5) +
tm_shape(hexagon_all) +
  tm_fill("PM_TRIPS", palette = "-Spectral", fill_alpha = 0.5) +
  tm_borders(col = "#B3C8CF", lwd = 0.1) +
  tm_layout(bg.color = "#D0DDD0",
            outer.margins = c(0.01, 0.01, 0.01, 0.01),
            main.title = "PM Peak Hour Trips by Hexagon",
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_view(set_zoom_limits = c(11,18))

# Add sqr-transformed column
hexagon_all <- hexagon_all %>%
  mutate(PM_TRIPS_SQRT = sqrt(PM_TRIPS))

#tmap_mode("view")

pm_tm_qt <- tm_shape(mpsz) +
  tm_polygons(fill = "#FFF4EA", alpha = 0.3) +
  tm_text("SUBZONE_N", size = 0.5) +
  tm_shape(hexagon_all) +
  tm_fill("PM_TRIPS", palette = "-Spectral", fill_alpha = 0.6,
          style = "quantile"
          ) +
  tm_borders(col = "#B3C8CF", lwd = 0.1) +
  tm_layout(bg.color = "#D0DDD0",
            outer.margins = c(0.01, 0.01, 0.01, 0.01),
            main.title = "PM Peak Hour Trips (Quantile)",
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_shape(mrt_sf) +
  tm_symbols(fill = "#CA8787", size = 0.5, col = "#D9E9CF") +
  tm_view(set_zoom_limits = c(11,18))

tmap_arrange(pm_tm, pm_tm_qt, 
             asp=1, ncol=1,
             sync = TRUE)

tmap_mode("plot")
```

## Weekends

```{r}
#| fig-width: 12
#| fig-height: 7
#| align: center
#| code-fold: true

#tmap_mode("view")
wkd_tm <- tm_shape(mpsz) +
  tm_polygons(fill = "#A1C2BD", alpha = 0.3) +
  tm_text("SUBZONE_N", size = 0.5) +
tm_shape(hexagon_all) +
  tm_fill("WKD_TRIPS", palette = "-Spectral", fill_alpha = 0.5) +
  tm_borders(col = "#B3C8CF", lwd = 0.1) +
  tm_layout(bg.color = "#D0DDD0",
            outer.margins = c(0.01, 0.01, 0.01, 0.01),
            main.title = "Weekend Peak Hour Trips by Hexagon",
            legend.position = c("right", "bottom"),
            frame = FALSE) +
    tm_view(set_zoom_limits = c(11,18))

# Add sqr-transformed column
hexagon_all <- hexagon_all %>%
  mutate(WKD_TRIPS_SQRT = sqrt(WKD_TRIPS))

#tmap_mode("view")
wkd_tm_qt <- tm_shape(mpsz) +
  tm_polygons(fill = "#A1C2BD", alpha = 0.3) +
  tm_text("SUBZONE_N", size = 0.5) +
  tm_shape(hexagon_all) +
  tm_fill("WKD_TRIPS", palette = "-Spectral", fill_alpha = 0.6,
          style = "quantile") +
  tm_borders(col = "#B3C8CF", lwd = 0.1) +
  tm_layout(bg.color = "#D0DDD0",
            outer.margins = c(0.01, 0.01, 0.01, 0.01),
            main.title = "Weekend Peak Hour Trips (Quantile)",
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_shape(mrt_sf) +
  tm_symbols(fill = "#CA8787", size = 0.5, col = "#D9E9CF"
  ) +
  tm_view(set_zoom_limits = c(11,18))

tmap_arrange(wkd_tm, wkd_tm_qt, 
             asp=1, ncol=1,
             sync = TRUE)
tmap_mode("plot")
```
:::

### 3.3 Comparative Maps for Peak Hour Trips (delete)

```{.r}
#| fig-width: 12
#| fig-height: 8
#| code-fold: true
#| eval: false

# Create three maps
am_map <- tm_shape(mpsz) +
  tm_text("SUBZONE_N", size = 0.5) +
  tm_borders(col = "grey70", lwd = 0.3) +
  tm_shape(hexagon_all) +
  tm_fill("AM_TRIPS", palette = "Oranges", title = "AM Trips", alpha = 0.5) +
  tm_borders(col = "grey20", lwd = 0.1) +
  tm_shape(mrt_sf) +
  tm_dots(fill = "#D97D55",
    size = 0.1,
    title = "MRT Stations") +
  tm_layout(legend.position = c("right", "bottom"))

pm_map <- tm_shape(mpsz) +
  tm_text("SUBZONE_N", size = 0.5) +
  tm_borders(col = "grey70", lwd = 0.3) +
  tm_shape(mrt_sf) +
  tm_dots(fill = "#D97D55",
    size = 0.1,
    title = "MRT Stations") +  
  tm_shape(hexagon_all) +
  tm_fill("PM_TRIPS", palette = "Blues", title = "PM Trips", alpha = 0.5) +
  tm_borders(col = "grey20", lwd = 0.1) +
  tm_layout(legend.position = c("right", "bottom"))

wkd_map <- tm_shape(mpsz) +
  tm_text("SUBZONE_N", size = 0.5) +
  tm_borders(col = "grey70", lwd = 0.3) +
  tm_shape(mrt_sf) +
  tm_dots(fill = "#D97D55",
    size = 0.1,
    title = "MRT Stations") +
  tm_shape(hexagon_all) +
  tm_fill("WKD_TRIPS", palette = "Purples", title = "Weekend Trips", alpha = 0.5) +
  tm_borders(col = "grey20", lwd = 0.1) +
  tm_layout(legend.position = c("right", "bottom"))

am_map
pm_map
wkd_map
#tmap_arrange(am_map, pm_map, wkd_map, ncol = 3)
```


## 4 Local Measures of Spatial Autocorrelation (LMSA) Analysis
For this study, we want to understand if passenger trips are even distributed geographically among the hexagons. From the previous plots, we can confidently yield the conclusion that they are not evenly distributed. In this section, we would like to use statistical methods to provide the evidence. 

If this proves our assumption, then our next question will be “is there sign of spatial clustering?”. If the answer is yes, then our next question will be “where are these clusters?”

### 4.1 Global Measures of Spatial Autocorrelation
In previous section 2, we have prepared the passender trips data per hexagon area in `hexagon_all`. The plots can be observed in section 3 to see equal interval classification and quantile classification.

Here we want to compute global spatial autocorrelation statistics and to perform spatial complete randomness test for global spatial autocorrelation.

#### **4.1.1 Compute contiguity spatial weights**

```{r}
wm_q <- poly2nb(hexagon_all, queen = TRUE)
summary(wm_q)
```
Summmary Report: There are 826 regions (hexagons) in the Singapore map. The most connected area unit has 6 neighbors (hexagon has 6 edages). There is only one unit with no neighbors.

#### **4.1.2 Row-standardised weights matrix**

Now we want to assign weights t =o each neighboring hexagon. In this study, each hexagon will be assigned equal weight (stylep = W). 
```{r}
rswm_q <- nb2listw(wm_q,
                   style = "W", zero.policy = TRUE)
rswm_q
```
#### **4.1.3 Moran's I Test**

Below we will perform Moran's I statistics testing by using `moran.test()` from spdep package. The test is being performed for the 3 different periods of time.

::: panel-tabset
## Weekday AM

**Moran's I Test**
```{r}
moran.test(hexagon_all$AM_TRIPS,
           listw = rswm_q,
           zero.policy = TRUE, na.action = na.omit)
```
Statisticl conclusion: 
- Given statistically significant spatial autocorrelation (p-value < 2.2e-16), we can confidently reject the null hypothesis of random spatial distribution.
- Moran I statistic is 0.3363, as a postive value that indicates similar values cluster together.
- The result indicates that hexagons with high passenger trips tend to be near other high-trip hexagons and hexagons with low passenger trips tend to be near other low passenger trip hexagons (cold spots). There are clear geographic clusters.

**Monte Carlo Moran's I**
We will run Monte Carlo Moran's I simulation for 1,000 times. 

```{r}
set.seed(1234)
bperm = moran.mc(hexagon_all$AM_TRIPS,
                 listw = rswm_q,
                 nsim = 999,
                 zero.policy = TRUE, na.action = na.omit)
bperm
```
Statistical conclusion:
- p-value = 0.001 indicates that it is highly significant (little chance that this pattern is reandom) and confirms the analytical test result from the Moran's I test.
- The result shows it is statistically significant by both analytical and empirical methods
- 

**Visualise Moran's I Test**

We plot the distribution of statistical values below. 
```{r}
mean <- mean(bperm$res[1:999])
print(paste0("Mean of permutations: ", mean))

var <- var(bperm$res[1:999])
print(paste0("Variance: ", var))

summary(bperm$res[1:999])
```
```{r}
perm_data <- data.frame(moran_i = bperm$res)

ggplot(perm_data, aes(x = moran_i)) +
  geom_histogram(bins = 20, fill = "#96A78D", color = "#DCDCDC") +
  geom_vline(xintercept = 0, color = "blue", linewidth = 0.5) +
  geom_vline(xintercept = 0.3362907435, color = "red", linewidth = 0.5) +
  labs(x = "Simulated Moran's I (Weekday AM)", y = "Frequency") +
  theme(plot.background = element_rect("#D0DDD0"),
        panel.background = element_rect(fill = "#D0DDD0"))
```


## Weekday PM
**Moran's I Test**
```{r}
moran.test(hexagon_all$PM_TRIPS,
           listw = rswm_q,
           zero.policy = TRUE, na.action = na.omit)
```
Statisticl conclusion:
- Though not as extreme as AM weekdays, PM weekdays' Moran's I test also shows highly significatn evidence to reject the null hypothesis, showing that clustering exists, just much weaker than AM weekdays.
- While the Moran's I = 0.140, PM trips show less than half the clustering of AM trips
- PM weekday clustering seems weaker. It may be because afternoon trips are more varied. People start trips from more varied locations (workplaces, shopping, schools, scattered throughout the city), so the trips origins are more spread out across commercialm business and retail areas. In addition, afternoon has more staggered departures across different times. For examples, some may leave office earlier than 5pm to avoid the traffic or fetch children from schools. 
- Chain trips like people may go for dinner or shopping from work before going back home. This would create more dispersed origin patterns as well. 

**Monte Carlo Moran's I**
```{r}
set.seed(1234)
bperm_pm = moran.mc(hexagon_all$PM_TRIPS,
                 listw = rswm_q,
                 nsim = 999,
                 zero.policy = TRUE, na.action = na.omit)
bperm_pm
```

Statistical conclusion:
- p-value = 0.001 shows that there is significant evidence to reject the null hypothesis, indicating that there is clustering in PM passenger trips.
- This confirms the analytical test and matches the analytical result.


**Visualise Moran's I Test**

We plot the distribution of statistical values below. 
```{r}
mean_pm <- mean(bperm_pm$res[1:999])
print(paste0("Mean of permutations: ", mean))

var_pm <- var(bperm_pm$res[1:999])
print(paste0("Variance: ", var))

summary(bperm_pm$res[1:999])
```


```{r}
perm_data_pm <- data.frame(moran_i = bperm_pm$res)

ggplot(perm_data_pm, aes(x = moran_i)) +
  geom_histogram(bins = 20, fill = "#F2AEBB", color = "#DCDCDC") +
  geom_vline(xintercept = 0, color = "blue", linewidth = 0.5) +
  geom_vline(xintercept = 0.1401003174, color = "red", linewidth = 0.5) +
  labs(x = "Simulated Moran's I (Weekday PM)", y = "Frequency") +
  theme(plot.background = element_rect("#D0DDD0"),
        panel.background = element_rect(fill = "#D0DDD0"))
```
## Weekend/Public holiday
**Moran's I Test**
```{r}
moran.test(hexagon_all$WKD_TRIPS,
           listw = rswm_q,
           zero.policy = TRUE, na.action = na.omit)
```
Statisticl conclusion:
- Weekend has p-value < 2.2e-16 showing extremely significant evidence for us to reject the null hypothesis. 
- Its Moran I statistic = 0.21684 is the middle among the 3 periods we study, stronger than PM weekdays.
- Weekend trip origins show moderate spatial clustering, suggesting that while people depart for leisure activities from concentrated residential areas, the pattern is less rigid than weekday morning commutes but more organized than weekday afternoon movements.

**Monte Carlo Moran's I**
```{r}
set.seed(1234)
bperm_wkd = moran.mc(hexagon_all$WKD_TRIPS,
                 listw = rswm_q,
                 nsim = 999,
                 zero.policy = TRUE, na.action = na.omit)
bperm_wkd
```

Statistical conclusion:
- p-value = 0.001 shows highly significant evidence for us to reject tje null hypothesis. This confirms the analytical test result (p < 2.2e-16)
- In 999 random permutations, none produced a Moran's I as high as 0.217. Therefore, weekend spatial clustering is definitely NOT due to chance. 


**Visualise Moran's I Test**

We plot the distribution of statistical values below. 
```{r}
mean_wkd <- mean(bperm_wkd$res[1:999])
print(paste0("Mean of permutations: ", mean))

var_wkd <- var(bperm_wkd$res[1:999])
print(paste0("Variance: ", var))

summary(bperm_wkd$res[1:999])
```


```{r}
perm_data_wkd <- data.frame(moran_i = bperm_wkd$res)

ggplot(perm_data_wkd, aes(x = moran_i)) +
  geom_histogram(bins = 20, fill = "#86B0BD", color = "#DCDCDC") +
  geom_vline(xintercept = 0, color = "blue", linewidth = 0.5) +
  geom_vline(xintercept = 0.2168371154, color = "red", linewidth = 0.5) +
  labs(x = "Simulated Moran's I (Weekend)", y = "Frequency") +
  theme(plot.background = element_rect(fill = "#D0DDD0"),
        panel.background = element_rect(fill = "#D0DDD0"))
```

:::

Conclusion:
- AM weekday passenger trips exhibit statistically significant positive spatial autocorrelation (Moran's I = 0.336, p < 0.001). Monte Carlo simulation with 999 permutations confirmed this finding. The observed statistic ranked highest among all simulations (p = 0.001), which indicates extremely strong evidence of spatial clustering.

- Spatial autocorrelation analysis reveals significant temporal variations in trip clustering patterns. Morning weekday trips exhibit the strongest spatial clustering (Moran's I = 0.336, p < 0.001), while afternoon trips show significantly weaker but still statistically significant clustering (Moran's I = 0.140, p < 0.001). Weekend trips display intermediate clustering strength (Moran's I = 0.217, p < 0.001). This suggests that trip patterns are **most concentrated during morning commutes** and **more dispersed during afternoon and weekend periods**.

- The result suggests a spatial pattern for leisure/weekend activities and people start weekend trips from clustered areas but not as rigidly as weekday commutes. It is distinctly different from both weekday morning and afternoon patterns. This validates that Singapore's weekend travel behavior follows predictable spatial patterns worth investigating further with LISA maps.

## 4.2 Local Measures of Spatial Autocorrelation (LMSA)
The Global measures of spatial autocorrelation suggests that all three study periods have clustering spatial patterns. We will now use Local Measures of Spatial Autocorrelation to compute the statistic of the passenger trips, and understand where the significant clusters are located by displaying the LMSA maps.

Below we will continue to use spdep to compute Local Moran's I statistic.

### 4.2.1 Calculate Local Moran's I

```{r}
hexagon_clean <- hexagon_all %>%
  mutate(nb = st_contiguity(geometry),
         wt = st_weights(
           nb, style = "W",
           allow_zero = TRUE),
         .before = 1) 
```

:::panel-tabset
## Weekday AM
```{r}
#AM
lisa_am <- hexagon_clean %>%
  mutate(local_moran_am = localmoran_perm(
  hexagon_clean$AM_TRIPS, 
  rswm_q, 
  nsim = 999,
  zero.policy = TRUE
))

lisa_am <- lisa_am %>%
  mutate(
    local_i_am = local_moran_am[,1],      # Local Moran's I
    expected_i = local_moran_am[,2],      # Expected value
    variance_i = local_moran_am[,3],      # Variance
    z_score_am = local_moran_am[,4],      # Standard deviation
    p_value_am = local_moran_am[,5],      # p-value
    significant_am = p_value_am < 0.05
  ) %>%
  select(-local_moran_am) 
```

```{r}
names(lisa_am)
```

## Weekday PM
```{r}
#AM
lisa_pm <- hexagon_clean %>%
  mutate(local_moran_pm = localmoran_perm(
  hexagon_clean$PM_TRIPS, 
  rswm_q, 
  nsim = 999,
  zero.policy = TRUE
))

lisa_pm <- lisa_pm %>%
  mutate(
    local_i_pm = local_moran_pm[,1],      # Local Moran's I
    expected_i = local_moran_pm[,2],      # Expected value
    variance_i = local_moran_pm[,3],      # Variance
    z_score_pm = local_moran_pm[,4],      # Standard deviation
    p_value_pm = local_moran_pm[,5],      # p-value
    significant_pm = p_value_pm < 0.05
  ) %>%
  select(-local_moran_pm) 
```

```{r}
names(lisa_pm)
```


## Weedkend/PH
```{r}
#AM
lisa_wkd <- hexagon_clean %>%
  mutate(local_moran_wkd = localmoran_perm(
  hexagon_clean$WKD_TRIPS, 
  rswm_q, 
  nsim = 999,
  zero.policy = TRUE
))

lisa_wkd <- lisa_wkd %>%
  mutate(
    local_i_wkd = local_moran_wkd[,1],      # Local Moran's I
    expected_i = local_moran_wkd[,2],      # Expected value
    variance_i = local_moran_wkd[,3],      # Variance
    z_score_wkd = local_moran_wkd[,4],      # Standard deviation
    p_value_wkd = local_moran_wkd[,5],      # p-value
    significant_wkd = p_value_wkd < 0.05
  ) %>%
  select(-local_moran_wkd) 
```

```{r}
names(lisa_wkd)
```


:::
### 4.2.2 Mapping Local Moran's I

```{r}


```

```{r}


```

```{r}
# AM


```



Compute LMSA statistic of the passengers trips generate by origin at hexagon level.

Display the LMSA maps of the passengers trips generate by origin at hexagon level. The maps should only display the significant (i.e. p-value < 0.05)

With reference to the analysis results, draw statistical conclusions (not more than 200 words per visual).

1.  Calculate Local Moran's I first
2.  Build complexity: Add Local Geary's C and Getis-Ord Gi\*
3.  Focus on interpretation: What do the patterns mean for urban planning?

## 5 Emerging Hot Spot Analysis(EHSA)
With reference to the passenger trips by origin at the hexagon level for the three time intervals given above:

Perform Mann-Kendall Test by using the spatio-temporal local Gi* values,
Prepared EHSA maps of the Gi* values of the passenger trips by origin at the hexagon level. The maps should only display the significant (i.e. p-value < 0.05).
With reference to the EHSA maps and data visualisation prepared, describe the spatial patterns reveled. (not more than 250 words per cluster).

1.  Create space-time cube: Stack the three time periods
2.  Apply Mann-Kendall test to detect trends
3.  Categorize patterns: New, intensifying, diminishing, sporadic

## 6 Space-time analysis

animated time-series maps

## 7 Key findings & Conclusions

**Visualisation** Create comparison layouts (e.g., side-by-side time periods) Highlight only significant results (p \< 0.05) Write concise, insightful captions focusing on: What the pattern shows, Why it matters for transportation planning, Potential causes or implications

**Extra data - MRT data**

## Reference
