---
title: "02 - Singapore Bus Passenger Travel Patterns"
author: "Cathy C."
date: "Oct 7, 2025"
date-modified: "last-modified" #auto render date
format: html
execute: #how YML page work
  echo: true #will show code
  eval: true #will run code every time and show output
  warning: false #dont show warning msg
  freeze: true #controls rendering - no render if no changes
---

## 1 Overview

## 2 Data sources & preparation

### 2.1 Load packages

The packages that we will use for this study are outlined below:

| R Package | Description |
|------------------|------------------------------------------------------|
| **sf** | A package that provides [**simple features access**](https://en.wikipedia.org/wiki/Simple_Features) for R. |
| **spdep** | Creates spatial weights matrices and to test for and model spatial autocorrelation in geographical data,  like Moran's I and Geary's C and Local Moran's I and Getis-Ord G. |
| **tidyverse** | Includes the core packages like ggplot2, dplyr, readr, purrr, tibble, lubridate, etc. |
| **sfdep** | Builds on the great shoulders of **spdep** package for spatial dependence. sfdep creates an sf and tidyverse friendly interface to the package as well as introduces new functionality that is not present in spdep. |
| **h3** | H3 is a geospatial indexing system using a hexagonal grid that can be (approximately) subdivided into finer and finer hexagonal grids, combining the benefits of a hexagonal grid with [S2](https://code.google.com/archive/p/s2-geometry-library/)'s hierarchical subdivisions. |
| **knitr** | Color maps designed to improve graph readability for readers with common forms of color blindness and/or color vision deficiency. |
| **tmap** | geographical maps in which spatial data distributions are visualized. This package offers a flexible, layer-based, and easy to use approach to create thematic maps, such as choropleths and bubble maps. |

Let's install and load the following packages: **tidyverse, sf, sfdep, tmap, knitr** in R environment.

```{r}
pacman::p_load(tidyverse, sf, sfdep, tmap, knitr, h3)
```

### 2.2 Import data

We will import 3 data sets for this study:

1.  Import *Bus Stop Location* shapfile from LTA DataMall into R environment
2.  Import *Master Plan 2019 Subzone Boundary (No Sea)* from <https://data.gov.sg/>
3.  Import *Passenger Volume by Origin Destination Bus Stops* from LTA DataMall

When importing data, need to check first. Make sure busstop code is in character form rather than numeric. If it's latter, need to insert 0 because the stop code can start with zero.

Open Bustop.prj and check if the data has the information we need, or need to transform. Below we will start importing the data.

::: panel-tabset

## **Master Plan Subzone Boundary**

Below we will import *Master Plan 2019 Subzone Boundary (No Sea).*

```{r}
mpsz <- st_read("data/MasterPlan2019SubzoneBoundaryNoSeaKML.kml")
```

```{r}
mpsz <- read_rds("data/rds/mpsz.rds")
st_crs(mpsz)
```

## **Bus Stop Location**

Below we will import *Bus Stop Location* shapfile from LTA DataMall.

```{r}
BusStop = st_read(dsn = "data/LTADataMall",
                  layer = "BusStop")
```

Check the data type:

```{r}
st_crs(BusStop)
```

Based on the above, we identify that the the EPSG is not aligned with mpsz, so we need to transform the data for EPSG:3414 Projected coordinate system for Singapore.

```{r}
BusStop <- BusStop %>%
  st_transform(crs = 3414)
```

Check again on BusStop:

```{r}
st_crs(BusStop)
```

Now, we see BusStop is aligned with mpsz.

## **Passenger Volume**

Below code is used to import *Passenger Volume by Origin Destination Bus Stops* from LTA DataMall.

```{r}
odbus <- read_csv("data/origin_destination_bus_202508.csv")
```

```{r}
st_crs(odbus)
```

describe the coordinate system EPSG, SF

:::

### 2.3 Data wrangling: Prepare hexagon data

#### 2.3.1 Extracting bus stops in Singapore

First we plot the bus stops in Singapore below:

```{r}
#| fig-width: 12
#| fig-height: 8

tm_shape(mpsz) +
  tm_polygons() +
  tm_shape(BusStop) +
  tm_dots(size = 0.3, fill = "#344F1F") +
  tm_layout(bg.color = "#D0DDD0",
            outer.margins = c(0.01, 0.01, 0.01, 0.01))
```

We see that the dots (bus stops) outside the boundaries are buses for JB, Malaysia. Therefore, **these have to be excluded.** One way to exclude them is to plot the mpsz layer on top of the BusStop layer.

If we have mpsz in front of the layer, it will capture all bus stops. When we plot first layer using Busstop before mpsz, these spots for JB will be covered.

```{r}
BusStop_in_SG <- st_join(
  BusStop, mpsz, #BusStop first
  join = st_within,
  left = FALSE
)
```

The, we plot again:

```{r}
#| fig-width: 12
#| fig-height: 8

tm_shape(mpsz) +
  tm_polygons() +
  tm_shape(BusStop_in_SG) +
  tm_dots(size = 0.3, fill = "#E45A92") +
  tm_layout(bg.color = "#D0DDD0",
            outer.margins = c(0.01, 0.01, 0.01, 0.01))
```

#### 2.3.2 Deriving Analytical Hexagon

We want to create regular shapes layers with hexagon grid (square and hexagon are 2 most used), and exclude those without bus stops. Create the hexagon layer covering the entire Singapore.

We will need to derive the hexagon data of **375m** (this distance is the perpendicular distance between the centre of the hexagon and its edges) to represent the [traffic analysis zone (TAZ)](https://tmg.utoronto.ca/files/Reports/Traffic-Zone-Guidance_March-2021_Final.pdf). `cellsize` is the distance between opposite edges of the hexagon. This distance is twice the perpendicular distance from the center to any edge.

```{r}
hexagon <- st_make_grid(mpsz, cellsize = 750,
                        what = "polygon",
                        square = FALSE) %>% 
# not a tibbler data frame
  st_sf() # so need to convert to sf
```

The output does not have a unique ID.

#### 2.3.3 Filtering hexagons with bus stops

We only want to retain hexagons with bus stops. Use `st_intersects()` to return 2 values: True, False (0 or 1). The hexagon with a bus stop would be 1, and without a bus stop would return 0.

```{r}
hexagon$n_collisions = 
  lengths(st_intersects(
    hexagon, BusStop_in_SG
  )) #n_collisions show the number of bus stops.

hexagon <- filter(hexagon, n_collisions >0) #retain all that have bus stops.
```

#### 2.3.4 Visualise the output

```{r}
#| fig-width: 12
#| fig-height: 8
tm_shape(mpsz) +
  tm_polygons() +
  tm_shape(hexagon) +
  tm_fill(col = "black", fill = "#FFD5D5") +
  tm_layout(bg.color = "#D0DDD0",
            outer.margins = c(0.01, 0.01, 0.01, 0.01))
```

We want to know the trips occurred in each hexagon in certain time periods.

The hexagons do not have an ID, so we will give each an ID.

#### 2.3.5 Assign IDs to each hexagon

```{r}
hexagon <- hexagon %>%
  select(, -n_collisions)

hexagon$HEX_ID <- sprintf(
  "H%04d", seq_len(nrow(hexagon))
) %>%
  as.factor() #they will in ordinal scale

kable(head(hexagon))
```

### 2.4 Prepare trip generation data

#### 2.4.1 Clearing the data

We dont need the destination code. Below we will select the columns we need, including `ORIGIN_PT_CODE`, `DAY_TYPE`, `TIME_PER_HOUR`, and `TOTAL_TRIPS`. We also simply the code to join the datasets together later.

```{r}
trips <- odbus %>%
  select(c(ORIGIN_PT_CODE, DAY_TYPE, TIME_PER_HOUR, TOTAL_TRIPS)) %>%
  rename(BUS_STOP_N = ORIGIN_PT_CODE) %>% #simply the code for joining.
  rename(HOUR_OF_DAY = TIME_PER_HOUR) %>%
  rename(TRIPS = TOTAL_TRIPS)
trips$BUS_STOP_N <- as.factor(trips$BUS_STOP_N) #convert to factor so it's not numerical.
kable(head(trips))
```

**Combine bus stop with Hex ID into hexagon layer.**

```{r}
bs_hex <- st_intersection(
  BusStop, hexagon
) %>%
  st_drop_geometry() %>%
  select(c(BUS_STOP_N, HEX_ID)) #drop away the rest
kable(head(bs_hex))
```

#### 2.4.2 Add HEX_ID into bus trips data

**Use inner_join() to join trips and bus_hex.**

```{r}
trips <- inner_join(trips, bs_hex)
kable(head(trips))
```

\*\* count trips per hexagon?

### 2.5 Time preiod filtering

Create separate datasets for each time period

## 3 Exploratory spatial analysis

Bus geographical distribution of passenger trips

## 4 Local Measures of Spatial Autocorrelation (LMSA) Analysis

1.  Calculate Local Moran's I first
2.  Build complexity: Add Local Geary's C and Getis-Ord Gi\*
3.  Focus on interpretation: What do the patterns mean for urban planning?

## 5 Emerging Hot Spot Analysis(EHSA)

1.  Create space-time cube: Stack the three time periods
2.  Apply Mann-Kendall test to detect trends
3.  Categorize patterns: New, intensifying, diminishing, sporadic

## 6 Space-time analysis

animated time-series maps

## 7 Key findings & Conclusions

**Visualisation** Create comparison layouts (e.g., side-by-side time periods) Highlight only significant results (p \< 0.05) Write concise, insightful captions focusing on: What the pattern shows, Why it matters for transportation planning, Potential causes or implications

**Extra data - MRT data**

## Reference
