---
title: "04 - Spatial Weights and Applications"
author: "Cathy C."
date: "Sep 17, 2025"
date-modified: "last-modified" #auto render date
format: html
execute: #how YML page work
  echo: true #will show code
  eval: true #will run code every time and show output
  warning: false #dont show warning msg
  freeze: true #controls rendering - no render if no changes
---

## Overview

Starting from this, we will work on spatial weights and applications. In this exercise, we will learn how to compute spatial weights using R.

## 1 Objectives

We will use the the ppropriate function(s) of below packages to achieve various tasks & goals.

| Package   | Task                                                           |
|------------------|------------------------------------------------------|
| **sf**    | Import geospatial data                                         |
| **readr** | Import csv file                                                |
| **dplyr** | Perform relational join                                        |
| **spdep** | Compute spatial weights & calculate spatially lagged variables |

## 2 The Study Area and Data

Two data sets will be used in this hands-on exercise, they are:

-   Hunan county boundary layer. This is a geospatial data set in ESRI shapefile format.

-   *Hunan_2012.csv*: This csv file contains selected Hunan’s local development indicators in 2012.

### 2.1 Getting started

First we will install the necessary packages or ensure they are installed.

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse, knitr)
```

### 2.2 Getting data into R environment

The geospatial data is in **ESRI shapefile** format and the attribute table is in **csv fomat**. We will bring the geospatial data and its associated attribute table into R environment.

#### [**2.2.1 Import shapefile into R environment**]{style="color: #7D8D86;"}

Below we use [*st_read()*](https://r-spatial.github.io/sf/reference/st_read.html) of **sf** package to import Hunan shapefile into R. The imported shapefile will be **simple features** Object of **sf**.

```{r}
hunan <- st_read(dsn = "data/geospatial", layer = "Hunan")
```

#### [**2.2.2 Import csv file into R environment & Performing relational join**]{style="color: #7D8D86;"}

Use *read_csv()* of **readr** package to import *Hunan_2012.csv*. The output will be R dataframe class.

Then, use *left_join* from **dplyr** package to update the attribute table ofHunan's SpatialPolygonsDataFrame with the attribute fields of *hunan2012* dataframe.

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")

hunan <- left_join(hunan, hunan2012) %>% #this match by county
  select(1:4, 7, 15) #keep county info and GDPPC
```

## 3 Visualising Regional Development Indicator

Prepare a basemap and a choropleth map that shows the distribution of GDPPC 2012 by using `qtm()` from **tmap** package.

```{r}
#| fig-width: 12
#| fig-align: center
basemap <- tm_shape(hunan) +
  tm_polygons() +
  tm_text("NAME_3", size = 0.5) +
  tm_layout(bg.color = "#BBDCE5", main.title = "Hunan Basemap")

gdppc <- qtm(hunan, "GDPPC") +
  tm_layout(
    legend.position = c("left", "bottom"),
    bg.color = "#BBDCE5", main.title = "GDP Per Capita"
  )
tmap_arrange(basemap, gdppc, asp = 1, ncol = 2)
```

## 4 Computing Contiguity Spatial Weights

We will use [*poly2nb()*](https://r-spatial.github.io/spdep/reference/poly2nb.html) of **spdep** package to compute **contiguity weight matrices** for the study area.

[*poly2nb()*](https://r-spatial.github.io/spdep/reference/poly2nb.html) builds a neighbours list based on regions with contiguous boundaries.

The documentation says that you can pass a “queen” argument that takes TRUE or FALSE as options where the default is set to TRUE. And then this function will return a list of first order neighbours using the Queen criteria.

### 4.1 **Computing (QUEEN) contiguity based neighbours**

Let's compute Queen contiguity weight matrix.

```{r}
wm_q <- poly2nb(hunan, queen = TRUE)
summary(wm_q)
```

::::: markgoals
::: markgoals-header
**Learning**
:::

::: markgoals-container
The summary report above shows that there are **88 area units** in Hunan. The most connected area unit has **11 neighbours**. There are two area units with only one heighbours.
:::
:::::

For each polygon in the polygon object, `wm_q` lists all neighboring polygons. For example, to see the neighbors for the first polygon in the object, type:

```{r}
wm_q[[1]]
```

::::: markgoals
::: markgoals-header
**Learning**
:::

::: markgoals-container
This means Polygon 1 has 5 neighbours. The numbers represent the polygon IDs stored in [hunan SpatialPolygonsDataFrame]{style="color: #BA487F;"} class.
:::
:::::

We can also retrieve the country name of Polygon ID = 1 below:

```{r}
hunan$County[1]
```

Above output shows that Polygon ID = 1 is Anxiang county.

To reveal the names of the 5 neighbours, we use `hunan$NAME_3`. [NAME_3]{style="color: #BA487F;"} is the County name column in [hunan]{style="color: #BA487F;"}.

```{r}
hunan$NAME_3[c(2,3,4,57,85)]
```

We can also retrieve the GDPPC of the 5 counties. Here is how:

1.  retreieve IDs of the neighbours
2.  retrieve their GDPPC from hunan using the IDs

```{r}
nb1 <- wm_q[[1]]
nb1 <- hunan$GDPPC[nb1]
nb1
```

The printed output shows the GDPPC of these 5 nearest neighbours based on Queen's method.

We can also display the complete weight matrix by using `str()`. As it shows there are 88 lists (counties).

```{r}
str(wm_q)
```

### 4.2 **Creating (ROOK) contiguity based neighbours**

Next we will compute Rook contiguity weight matrix.

```{r}
wm_r <- poly2nb(hunan, queen = FALSE)
summary(wm_r)
```

::::: markgoals
::: markgoals-header
**Learning**
:::

::: markgoals-container
-   The report shows there are 88 areas in Hunan.

-   The most connected area unit has 10 neighbours.

-   There are 2 area units with only one neighbour.
:::
:::::

### 4.3 **Visualising contiguity weights**

To visualise contiguity weights, we use connectivity graph. We will get the longitude and latitude of the Polygon Centroids. We will use a mapping function to get the coordinates in a separate data frame.

The mapping function applies a given function to each element of a vector and returns a vector of the same length. The input is the geometry column of `us.bound` and the function is `st_centroid`.

We will also use `map_dbl` variation of map from the **purrr** package.

1.  Get longitude values: map the st_centroid function over the geometry column of us.bound and access the longitude value through double bracket notation [\[\[\]\] and 1]{style="color: #BA487F;"}. This allows us to get the longitude.
2.  Use the same method for latitude with \[\[2\]\].
3.  Use [cbind]{style="color: #BA487F;"} to put longitude and latitude into the same object.

```{r}
longitutde <- map_dbl(hunan$geometry, ~st_centroid(.x)[[1]])

latitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[2]])

coords <- cbind(longitutde, latitude)
```

4.  Check the first few observations to see if they were formatted correctly.

```{r}
head(coords)
```

### 4.3.1 Queen contiguity vs Rook contiguity

Hover over two tabs and click to see the differences.

::: panel-tabset
## Queen contiguity based neighbours map

```{r}
#| fig-width: 12
#| fig-height: 8
par(bg = "#D0DDD0")
plot(hunan$geometry, border = "grey")
plot(wm_q, coords, pch = 20, cex = 0.6, add = TRUE, col = "red")
```

## Rook contiguity based neighbours map

```{r}
#| fig-width: 12
#| fig-height: 8
par(bg = "#D0DDD0")
plot(hunan$geometry, border = "grey")
plot(wm_r, coords, pch = 20, cex = 0.6, add = TRUE, col = "red")
```
:::

### 4.3.1 Queen contiguity vs Rook contiguity Comparison

```{r}
#| fig-width: 12
#| fig-height: 8
#| fig-align: center
par(mfrow = c(1,2), bg = "#D0DDD0")
plot(hunan$geometry, border = "grey", main = "Queen Contiguity")
plot(wm_q, coords, pch = 18, cex = 0.6, add = TRUE, col = "red")
plot(hunan$geometry, border = "grey", main = "Rook Contiguity")
plot(wm_r, coords, pch = 18, cex = 0.6, add = TRUE, col = "red")

```

## 5 Computing distance based neighbours

In this section, we will derive distance-based weight matrices by using [*dnearneigh()*](https://r-spatial.github.io/spdep/reference/dnearneigh.html) of **spdep** package. The function creates **distance-based neighbor relationships** for spatial analysis by identifying which geographic areas are "neighbors" based on distance ranges.

-   It will set a distance band with **lower (d1)** and **upper (d2)** bounds using the `bounds=` argument. Areas within this distance range become neighbors.

-   For distance calculation, the **projected coordinates** uses Euclidean (straight-line) distance. The **geographic coordinates** (with `longlat=TRUE`) uses great circle distance in **km**, accounting for Earth's curvature via WGS84 ellipsoid.

-   An example is with d1=5 and d2=20, each region considers only areas between 5-20 km away as neighbors.

### 5.1 **Determine the cut-off distance**

This step finds the **maximum distance** for the distance band.

[**1: Find k-nearest neighbors**]{style="color: #84994F;"}\
`knearneigh()` finds the k closest neighbors for each area. Returns a matrix showing which areas are neighbors.

[**2: Convert to neighbor list**]{style="color: #84994F;"}\
`knn2nb()` converts the matrix into a more usable list format. Each area gets a list of its neighbor IDs.

[**3: Calculate actual distances**]{style="color: #84994F;"}\

`nbdists()` measures the real distances between each area and its neighbors. Units meters/km for projected coordinates, km for lat/lon coordinates.

[**4: Flatten the data**]{style="color: #84994F;"}\
`unlist()` converts the nested list structure into a simple vector of all distances.

```{r}
k1 <- knn2nb(knearneigh(coords))
k1dists <- unlist(nbdists(k1, coords, longlat = TRUE))
summary(k1dists)
```

::::: markgoals
::: markgoals-header
**Learning**
:::

::: markgoals-container
Above summary report shows that the largest first nearest neighbour distance is 61.79km, so using this as the upper threshold gives certainty that all units will have at least ONE neighbour.
:::
:::::

### 5.2 **Computing fixed distance weight matrix**

Next we use `dnearneigh()` to compute the distance weight matrix.

```{r}
wm_d62 <- dnearneigh(coords, 0, 62, longlat = TRUE)
wm_d62
```

::::: markgoals
::: markgoals-header
**Learning**
:::

::: markgoals-container
-   There are 88 polygons in the dataset.

-   There are 324 pairs of regions that are considered neighbors.

-   Out of all possible pairs of regions (88 × 88 = 7,744 total possible connections), only 4.18% are actually neighbors.

-   On average, each region has about 3.68 neighbors
:::
:::::

Next we will display the content of `wm_d62` weight matrix using [str()]{style="color: #BA487F;"} & [*table()*](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/table){style="color: #BA487F;"} and [*card()*](https://r-spatial.github.io/spdep/reference/card.html){style="color: #BA487F;"} of **spdep**.

::: panel-tabset
## str()

```{r}
str(wm_d62)
```

## table() + card()

```{r}
table(hunan$County, card(wm_d62))
```
:::

#### **Connected Components**

1.  `n.comp.nb()` analyzes the neighbor list (`wm_d62`) to identify connected components. The function returns information about how many separate groups/clusters exist in your spatial network.

```{r}
n_comp <- n.comp.nb(wm_d62)
n_comp
```

::::: markgoals
::: markgoals-header
**Learning**
:::

::: markgoals-container
-   The **outcome 1** shows that all 88 regions form **one connected network**. If 2 or more, it means the spatial data has disconnected clusters.

-   This is an important diagnostic check before proceeding with spatial econometric models or spatial autocorrelation analysis.
:::
:::::

2.  Below code uses `table()` to count how many regions are in each component, and show the size distribution of the connected components.

```{r}
table(n_comp$comp.id)
```

::::: markgoals
::: markgoals-header
**Learning**
:::

::: markgoals-container
-   All 88 regions belong to component 1.
-   There is one fully connected spatial network.
:::
:::::

#### **5.2.1 Plot fixed distance weight matrix**

::: panel-tabset
## Distance weight matrix

The red lines show the links of 1st nearest neighbours and the black lines show the links of neighbours within the cut-off distance of 62km.

```{r}
#| fig-width: 12
#| fig-height: 8

par(bg = "#D0DDD0")
plot(hunan$geometry, border = "grey60")
plot(wm_d62, coords, add = TRUE)
plot(k1, coords, add = TRUE, col = "red", length = 0.08)
```

## Links comparison

We can also plot both nearest neighbours and neighbours within 62km.

```{r}
#| fig-width: 12
#| fig-height: 8
par(mfrow = c(1, 2), bg = "#D0DDD0")
plot(hunan$geometry, border = "grey60", main = "1st nearest neighbours")
plot(k1, coords, add = TRUE, col = "red", length = 0.08)
plot(hunan$geometry, border = "grey60", main = "Distance link")
plot(wm_d62, coords, add = TRUE, pch = 19, cex = 0.6)
```
:::

### 5.3 **Computing adaptive distance weight matrix**

In fixed distance weight matrix, the more densely settled areas (usually the urban areas) tend to have more neighbours and the less densely settled areas (usually the rural counties) tend to have lesser neighbours. Having many neighbours smoothes the neighbour relationship across more neighbours.

To control the number of neighbours directly using k-nearest neighbours, use 1. accept asymmetric neighbours 2. impose symmatry in the code below

```{r}
knn6 <- knn2nb(knearneigh(coords, k = 6))
knn6
```

**No let's display the matrix:**

```{r}
str(knn6)
```

::::: markgoals
::: markgoals-header
**Learning**
:::

::: markgoals-container
From the outcome we can see each country has exactly **6 neighbours**!
:::
:::::

### 5.3.1 Plot distance based neighbours

```{r}
#| fig-width: 12
#| fig-height: 8
par(bg = "#D0DDD0")
plot(hunan$geometry, border = "grey60")
plot(knn6, coords, pch = 9, cex = 0.6, add = TRUE, col = "red")
```

## 6 Weights based on IDW

### 6.1 **Inverse distance weights** 

Here we use **Inversed Distance method** to derive a spatial weight matrix. In the first step, we compute the distance between areas by [*nbdists()*](https://r-spatial.github.io/spdep/reference/nbdists.html) of **spdep**.

`lapply()` applies a function to each element in the `dist` list. The function `function(x) 1/(x)` calculates the inverse of each distance. This transforms distances into weights where **closer neighbors get higher weights**

```{r}
dist <- nbdists(wm_q, coords, longlat = TRUE)
ids <- lapply(dist, function(x) 1/(x))
ids
```

### 6.2 Row-standardised Weights Matrix

We use `nb2listw(wm_q, style="W")`to create a spatial weights matrix where each neighbor gets equal weight calculated as 1 divided by the total number of neighbors. This creates row standardisation where each row sums to 1. If a region has 4 neighbors, each gets 0.25 weight. If another region has 2 neighbors, each gets 0.50 weight. The spatial lag becomes a weighted average of all neighboring values.

**Edge problem:** **Interior regions** have more neighbors which creates stable estimates, while **edge regions** have fewer neighbors and potentially biased estimates. This could lead to over or under-estimating spatial autocorrelation patterns.

This approach is simple and intuitive to use. We could explore `style="B"` which uses binary weights and has different trade-offs. IN a nutshell, style="W" gives us reliable spatial lag calculations, but we should be aware that edge regions might be less reliable in thefinal results.

```{r}
rswm_q <- nb2listw(wm_q, style="W", zero.policy = TRUE)
rswm_q
```
::::: markgoals
::: markgoals-header
**Note**
:::

::: markgoals-container
The `zero.policy=TRUE` option allows for lists of non-neighbors. This should be used with caution since the user may not be aware of missing neighbors in their dataset however, a `zero.policy` of FALSE would return an error.
:::
:::::

### Check the weight of the 8th - 10th polygons' neighbors type:
```{r}
rswm_q$weights[8:10]
```

## 8 Application of Spatial Weight Matrix

### 8.1 **Spatial lag with row-standardized weights**

### 8.2 **Spatial lag as a sum of neighboring values**

### 8.3 **Spatial window average**

### 8.4 **Spatial window sum**

## 9 References
