---
title: "03 - Spatio-Temporal Point Patterns Analysis"
author: "Cathy C."
date: "Sep 17, 2025"
date-modified: "last-modified" #auto render date
format: html
execute: #how YML page work
  echo: true #will show code
  eval: true #will run code every time and show output
  warning: false #dont show warning msg
  freeze: true #controls rendering - no render if no changes
---

## Overview

Starting from this, we will work on spatial weights and applications. In this exercise, we will learn how to compute spatial weights using R.

## 1 Objectives

We will use the the ppropriate function(s) of below packages to achieve various tasks & goals.

| Package   | Task                                                           |
|-----------|----------------------------------------------------------------|
| **sf**    | Import geospatial data                                         |
| **readr** | Import csv file                                                |
| **dplyr** | Perform relational join                                        |
| **spdep** | Compute spatial weights & calculate spatially lagged variables |

## 2 The Study Area and Data

Two data sets will be used in this hands-on exercise, they are:

-   Hunan county boundary layer. This is a geospatial data set in ESRI shapefile format.

-   *Hunan_2012.csv*: This csv file contains selected Hunan’s local development indicators in 2012.

### 2.1 Getting started

First we will install the necessary packages or ensure they are installed.

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse, knitr)
```

### 2.2 Getting data into R environment

The geospatial data is in **ESRI shapefile** format and the attribute table is in **csv fomat**. We will bring the geospatial data and its associated attribute table into R environment.

#### [**2.2.1 Import shapefile into R environment**]{style="color: #7D8D86;"}

Below we use [*st_read()*](https://r-spatial.github.io/sf/reference/st_read.html) of **sf** package to import Hunan shapefile into R. The imported shapefile will be **simple features** Object of **sf**.

```{r}
hunan <- st_read(dsn = "data/geospatial", layer = "Hunan")
```

#### [**2.2.2 Import csv file into R environment & Performing relational join**]{style="color: #7D8D86;"}

Use *read_csv()* of **readr** package to import *Hunan_2012.csv*. The output will be R dataframe class.

Then, use *left_join* from **dplyr** package to update the attribute table ofHunan's SpatialPolygonsDataFrame with the attribute fields of *hunan2012* dataframe.

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")

hunan <- left_join(hunan, hunan2012) %>% #this match by county
  select(1:4, 7, 15) #keep county info and GDPPC
```

## 3 Visualising Regional Development Indicator

Prepare a basemap and a choropleth map that shows the distribution of GDPPC 2012 by using `qtm()` from **tmap** package.

```{r}
#| fig-width: 12
#| fig-align: center
basemap <- tm_shape(hunan) +
  tm_polygons() +
  tm_text("NAME_3", size = 0.5) +
  tm_layout(bg.color = "#BBDCE5", main.title = "Hunan Basemap")

gdppc <- qtm(hunan, "GDPPC") +
  tm_layout(
    legend.position = c("left", "bottom"),
    bg.color = "#BBDCE5", main.title = "GDP Per Capita"
  )
tmap_arrange(basemap, gdppc, asp = 1, ncol = 2)
```

## 4 Computing Contiguity Spatial Weights

### 4.1 **Computing (QUEEN) contiguity based neighbours**

### 4.2 **Creating (ROOK) contiguity based neighbours**

### 4.3 **Visualising contiguity weights**

## 5 Computing distance based neighbours

### 5.1 **Determine the cut-off distance**

### 5.2 **Computing fixed distance weight matrix**

### 5.3 **Computing adaptive distance weight matrix**

## 6 Weights based on IDW

## 7 Row-standardised Weights Matrix

## 8 Application of Spatial Weight Matrix

### 8.1 **Spatial lag with row-standardized weights**

### 8.2 **Spatial lag as a sum of neighboring values**

### 8.3 **Spatial window average**

### 8.4 **Spatial window sum**

## 9 References
