---
title: "04 - Spatial Weights and Applications"
author: "Cathy C."
date: "Sep 17, 2025"
date-modified: "last-modified" #auto render date
format: html
execute: #how YML page work
  echo: true #will show code
  eval: true #will run code every time and show output
  warning: false #dont show warning msg
  freeze: true #controls rendering - no render if no changes
---

## Overview

Starting from this, we will work on spatial weights and applications. In this exercise, we will learn how to compute spatial weights using R.

## 1 Objectives

We will use the the ppropriate function(s) of below packages to achieve various tasks & goals.

| Package   | Task                                                           |
|-----------------------|-------------------------------------------------|
| **sf**    | Import geospatial data                                         |
| **readr** | Import csv file                                                |
| **dplyr** | Perform relational join                                        |
| **spdep** | Compute spatial weights & calculate spatially lagged variables |

## 2 The Study Area and Data

Two data sets will be used in this hands-on exercise, they are:

-   Hunan county boundary layer. This is a geospatial data set in ESRI shapefile format.

-   *Hunan_2012.csv*: This csv file contains selected Hunan’s local development indicators in 2012.

### 2.1 Getting started

First we will install the necessary packages or ensure they are installed.

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse, knitr)
```

### 2.2 Getting data into R environment

The geospatial data is in **ESRI shapefile** format and the attribute table is in **csv fomat**. We will bring the geospatial data and its associated attribute table into R environment.

#### [**2.2.1 Import shapefile into R environment**]{style="color: #7D8D86;"}

Below we use [*st_read()*](https://r-spatial.github.io/sf/reference/st_read.html) of **sf** package to import Hunan shapefile into R. The imported shapefile will be **simple features** Object of **sf**.

```{r}
hunan <- st_read(dsn = "data/geospatial", layer = "Hunan")
```

#### [**2.2.2 Import csv file into R environment & Performing relational join**]{style="color: #7D8D86;"}

Use *read_csv()* of **readr** package to import *Hunan_2012.csv*. The output will be R dataframe class.

Then, use *left_join* from **dplyr** package to update the attribute table ofHunan's SpatialPolygonsDataFrame with the attribute fields of *hunan2012* dataframe.

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")

hunan <- left_join(hunan, hunan2012) %>% #this match by county
  select(1:4, 7, 15) #keep county info and GDPPC
```

## 3 Visualising Regional Development Indicator

Prepare a basemap and a choropleth map that shows the distribution of GDPPC 2012 by using `qtm()` from **tmap** package.

```{r}
#| fig-width: 12
#| fig-align: center
basemap <- tm_shape(hunan) +
  tm_polygons() +
  tm_text("NAME_3", size = 0.5) +
  tm_layout(bg.color = "#BBDCE5", main.title = "Hunan Basemap")

gdppc <- qtm(hunan, "GDPPC") +
  tm_layout(
    legend.position = c("left", "bottom"),
    bg.color = "#BBDCE5", main.title = "GDP Per Capita"
  )
tmap_arrange(basemap, gdppc, asp = 1, ncol = 2)
```

## 4 Computing Contiguity Spatial Weights

We will use [*poly2nb()*](https://r-spatial.github.io/spdep/reference/poly2nb.html) of **spdep** package to compute **contiguity weight matrices** for the study area.

[*poly2nb()*](https://r-spatial.github.io/spdep/reference/poly2nb.html) builds a neighbours list based on regions with contiguous boundaries.

The documentation says that you can pass a “queen” argument that takes TRUE or FALSE as options where the default is set to TRUE. And then this function will return a list of first order neighbours using the Queen criteria.

### 4.1 **Computing (QUEEN) contiguity based neighbours**

Let's compute Queen contiguity weight matrix.

```{r}
wm_q <- poly2nb(hunan, queen = TRUE)
summary(wm_q)
```

::::: markgoals
::: markgoals-header
**Learning**
:::

::: markgoals-container
The summary report above shows that there are **88 area units** in Hunan. The most connected area unit has **11 neighbours**. There are two area units with only one heighbours.
:::
:::::

For each polygon in the polygon object, `wm_q` lists all neighboring polygons. For example, to see the neighbors for the first polygon in the object, type:

```{r}
wm_q[[1]]
```

::::: markgoals
::: markgoals-header
**Learning**
:::

::: markgoals-container
This means Polygon 1 has 5 neighbours. The numbers represent the polygon IDs stored in [hunan SpatialPolygonsDataFrame]{style="color: #BA487F;"} class.
:::
:::::

We can also retrieve the country name of Polygon ID = 1 below:

```{r}
hunan$County[1]
```

Above output shows that Polygon ID = 1 is Anxiang county.

To reveal the names of the 5 neighbours, we use `hunan$NAME_3`. [NAME_3]{style="color: #BA487F;"} is the County name column in [hunan]{style="color: #BA487F;"}.

```{r}
hunan$NAME_3[c(2,3,4,57,85)]
```

We can also retrieve the GDPPC of the 5 counties. Here is how:

1.  retreieve IDs of the neighbours
2.  retrieve their GDPPC from hunan using the IDs

```{r}
nb1 <- wm_q[[1]]
nb1 <- hunan$GDPPC[nb1]
nb1
```

The printed output shows the GDPPC of these 5 nearest neighbours based on Queen's method.

We can also display the complete weight matrix by using `str()`. As it shows there are 88 lists (counties).

```{r}
str(wm_q)
```

### 4.2 **Creating (ROOK) contiguity based neighbours**

Next we will compute Rook contiguity weight matrix.

```{r}
wm_r <- poly2nb(hunan, queen = FALSE)
summary(wm_r)
```

::::: markgoals
::: markgoals-header
**Learning**
:::

::: markgoals-container
-   The report shows there are 88 areas in Hunan.

-   The most connected area unit has 10 neighbours.

-   There are 2 area units with only one neighbour.
:::
:::::

### 4.3 **Visualising contiguity weights**

To visualise contiguity weights, we use connectivity graph. We will get the longitude and latitude of the Polygon Centroids. We will use a mapping function to get the coordinates in a separate data frame.

The mapping function applies a given function to each element of a vector and returns a vector of the same length. The input is the geometry column of `us.bound` and the function is `st_centroid`.

We will also use `map_dbl` variation of map from the **purrr** package.

1.  Get longitude values: map the st_centroid function over the geometry column of us.bound and access the longitude value through double bracket notation [\[\[\]\] and 1]{style="color: #BA487F;"}. 
This allows us to get the longitude. 
2. Use the same method for latitude with [[2]].
3. Use [cbind]{style="color: #BA487F;"} to put longitude and latitude into the same object.

```{r}
longitutde <- map_dbl(hunan$geometry, ~st_centroid(.x)[[1]])

latitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[2]])

coords <- cbind(longitutde, latitude)
```
4. Check the first few observations to see if they were formatted correctly.
```{r}
head(coords)
```
### 4.3.1 Queen contiguity vs Rook contiguity
Hover over two tabs and click to see the differences.

::: panel-tabset
## Queen contiguity based neighbours map
```{r}
#| fig-width: 12
#| fig-height: 8
par(bg = "#D0DDD0")
plot(hunan$geometry, border = "grey")
plot(wm_q, coords, pch = 20, cex = 0.6, add = TRUE, col = "red")
```

## Rook contiguity based neighbours map
```{r}
#| fig-width: 12
#| fig-height: 8
par(bg = "#D0DDD0")
plot(hunan$geometry, border = "grey")
plot(wm_r, coords, pch = 20, cex = 0.6, add = TRUE, col = "red")
```

:::

### 4.3.1 Queen contiguity vs Rook contiguity Comparison

```{r}
#| fig-width: 12
#| fig-height: 8
#| fig-align: center
par(mfrow = c(1,2), bg = "#D0DDD0")
plot(hunan$geometry, border = "grey", main = "Queen Contiguity")
plot(wm_q, coords, pch = 18, cex = 0.6, add = TRUE, col = "red")
plot(hunan$geometry, border = "grey", main = "Rook Contiguity")
plot(wm_r, coords, pch = 18, cex = 0.6, add = TRUE, col = "red")

```

## 5 Computing distance based neighbours

### 5.1 **Determine the cut-off distance**

### 5.2 **Computing fixed distance weight matrix**

### 5.3 **Computing adaptive distance weight matrix**

## 6 Weights based on IDW

## 7 Row-standardised Weights Matrix

## 8 Application of Spatial Weight Matrix

### 8.1 **Spatial lag with row-standardized weights**

### 8.2 **Spatial lag as a sum of neighboring values**

### 8.3 **Spatial window average**

### 8.4 **Spatial window sum**

## 9 References
