---
title: "03b - Data Wranggling"
author: "Cathy C."
date: "Sep 13, 2025"
date-modified: "last-modified" #auto render date
format: html
execute: #how YML page work
  echo: true #will show code
  eval: true #will run code every time and show output
  warning: false #dont show warning msg
  freeze: true #controls rendering - no render if no changes
---

In this exercise, we will go through the steps for preparing data for Take-home exercise one where we will investigate the spatial and spatio-temporal patterns of new businesses and firms established in Singapore.

## 1 Loading the R package

We will use **tidyverse, sf, tmap, tmap,** and **httr** packages for this exercise to process data sets.

```{r}
pacman::p_load(tidyverse, sf, tmap, httr)
```

## 2 Importing ACRA data

We will import the ACRA data and transform them as tibble frame

```{r}
folder_path <- "data/aspatial/acra"
file_list <- list.files(path = folder_path,
                        pattern = "^ACRA*.*\\.csv$",
                        full.names = TRUE)
acra_data <- file_list %>%
  map_dfr(read_csv) #import and append all of them together!!
  
```

Then, we will write the data into .rds to store it in the data folder. This is so no need to process data every time.

```{r}
write_rds(acra_data, "data/rds/acra_data.rds")
```

## 3 Tidying ACRA data

With this step, we will extract the business code 56111, rename the columns and transform the data type necessarily.

```{r}
biz_56111 <- acra_data %>%
  select(1:24) %>%
  filter(primary_ssic_code == 56111) %>%
  rename(date = registration_incorporation_date) %>%
  mutate(date = as.Date(date),
         YEAR = year(date),
         MONTH_NUM = month(date),
         MONTH_ABBR = month(date, label = TRUE, abbr = TRUE)) %>% #this gives Jan, Feb, Mar...
  mutate(postal_code = str_pad(postal_code,
                               width = 6, side = "left", pad = "0")) %>% #if missing digits, will replace with zero.
  filter(YEAR == 2025)
```

## 4 Geocoding

(Avoid doing so for \> 5K datasets) We will perform the following in this step:

-   Extract unique postal code from the poastal_code field of the tidied tibble data frame.
-   Perform geocode by submitting the postal code to SLA OneMap Reverse Geocode API.
-   Save the return results into 2 data frame namely *found* and *not_found*.

```{r}
postcodes <- unique(biz_56111$postal_code)

url <- "https://onemap.gov.sg/api/common/elastic/search"

found <- data.frame()
not_found <- data.frame(postcode = character())

for (pc in postcodes) {
  query <- list(
    searchVal = pc,
    returnGeom = "Y",
    getAddrDetails = "Y",
    pageNum = "1"
  )
  
  res <- GET(url, query = query)
  json <- content(res)
  
  if(json$found != 0) {
    df <- as.data.frame(json$results, stringsAsFactors = FALSE)
    df$input_postcode <- pc
    found <- bind_rows(found, df)
  } else{
    not_found <- bind_rows(not_found, data.frame(postcodes = pc))
  }
}
```

## 5 Appending the location information

```{r}
biz_56111 = biz_56111 %>%
  left_join(found,
            by = c('postal_code' = 'POSTAL'))
```

## 6 Converting into SF data frame

```{r}
biz_56111_sf <- st_as_sf(biz_56111,
                         coords = c("X", "Y"),
                         crs = 3414)
```

```{r}
head(biz_56111_sf)
```
