---
title: "In-class exerricse 04"
author: "Cathy C."
date: "Sep 20, 2025"
date-modified: "last-modified" #auto render date
format: html
execute: #how YML page work
  echo: true #will show code
  eval: true #will run code every time and show output
  warning: false #dont show warning msg
  freeze: true #controls rendering - no render if no changes
---

### 1 Load libraries
```{r}
pacman::p_load(sf, ggstatsplot, tmap, tidyverse, knitr, GWmodel)
```

GWmodel presently is build around the plder sp and not sf formats for handling spatial data in R.

### 2 Importing data
```{r}
hunan_sf <- st_read(dsn = "data/geospatial", layer = "Hunan")
```


```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

### 3 Join datasets Hunan and Hunan_2012

```{r}
hunan_sf <- left_join(hunan_sf, hunan2012) %>% #this match by county
  select(1:3, 7, 15, 16, 31,32)
```


```{r}
hunan_sp <- hunan_sf %>%
  as_Spatial()
```

## 4 Compute bandwidth
### 4.1 **Cross-validation (CV)**
Will generate a score. It goes through iterations, and then stops at 76.29123.
The return is 76.29126, in KM.
```{r}
bw_CV <- bw.gwr(GDPPC ~ 1, data = hunan_sp,
                approach = "CV", adaptive = FALSE,
                kernel = "bisquare", longlat = T) #input data is in latlong
bw_CV
```

### 4.2 AIC
AIC is the alternative method. Return is bigger: 160.5517, in KM. 
```{r}
bw_AIC <- bw.gwr(GDPPC ~ 1, data = hunan_sp,
                approach = "AIC", adaptive = FALSE,
                kernel = "bisquare", longlat = T) #input data is in latlong
bw_AIC
```

## Compute fixed bandwidth

```{r}
gwstat <- gwss(data = hunan_sp, vars = "GDPPC",
               bw = bw_AIC, kernel = "bisquare",
               adaptive = FALSE, longlat = T)
```

## Convert to a daraframe 
Extract **SDF** data table from **gwss** object output from `gwss()`. 
It will be converted into data.frame using `as.data.frame()`.

```{r}
gwstat_df <- as.data.frame(gwstat$SDF)
```


Then use `cbind()` to append the newly derived data.frame onto hunan_sf sf data.frame.

```{r}
hunan_gstat <- cbind(hunan_sf, gwstat_df)
```

```{r}
# Summary of LISA results
summary(hunan_gstat$GDPPC_LM)   # Local Moran's I values
summary(hunan_gstat$GDPPC_LSD)  # Standard deviation
summary(hunan_gstat$GDPPC_LVar) # Local variance
summary(hunan_gstat$GDPPC_LSKe)
summary(hunan_gstat$GDPPC_LCV)

```

## Visualise geographically weighted statistics
```{r}
#| fig-width: 12
#| fig-height: 20

LM <- tm_shape(hunan_gstat) +
  tm_polygons("GDPPC_LM") +
  tm_layout(legend.outside = TRUE) +
  tm_title("GDPPC Local Moran")

LSD <- tm_shape(hunan_gstat) + 
  tm_polygons("GDPPC_LSD") +
  tm_layout(legend.outside = TRUE) +
  tm_title("Local Standard Deviation")

LSKe <- tm_shape(hunan_gstat) +
  tm_polygons("GDPPC_LSKe") +
  tm_layout(legend.outside = TRUE) +
  tm_title("GDPPC Local Skewness")

LVar <- tm_shape(hunan_gstat) + 
  tm_polygons("GDPPC_LVar") +
  tm_layout(legend.outside = TRUE) +
  tm_title("Local Variance")

LCV <- tm_shape(hunan_gstat) + 
  tm_polygons("GDPPC_LCV") +
  tm_layout(legend.outside = TRUE) +
  tm_title("Local Coefficient of Variation")

tmap_arrange(LM, LSD, LSKe, LVar, LCV, asp = 1, ncol = 1)
  

```

