---
title: "In-class Exercise 05"
author: "Cathy C."
date: "Sep 27, 2025"
date-modified: "last-modified" #auto render date
format: html
execute: #how YML page work
  echo: true #will show code
  eval: true #will run code every time and show output
  warning: false #dont show warning msg
  freeze: true #controls rendering - no render if no changes
---

## 1 Getting started

We will install and load the R packages **sf, tmap, sfdep** and **tidyverse** packages in R environment.

```{r}
pacman::p_load(sf, tmap, sfdep, tidyverse)
```

## 2 The data

### 2.1 Import the data sets

We will be using 2 data sets in this use case.

-   Hunan: a geospatial data
-   Hunan_2012, an attribute data set in csv format

```{r}
hunan <- st_read(dsn = "data/geospatial", layer = "Hunan")

hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

### 2.2 Combine data into one.

```{r}
hunan_GDPPC <- left_join(hunan,hunan2012) %>%
  dplyr::select(1:4, 7, 15)
```

**Step 1**

Derive Queen's contiguity weights: **sfdep** methods

```{r}
wm_q <- hunan_GDPPC %>%
  mutate(nb = st_contiguity(geometry),
         wt = st_weights(nb, style = "W"), #calculate weight metrics
         .before = 1)
```

```{r}
class(wm_q) #it's a data frame.
```

st_weights() provides tree arguments:

-   nb: a neighbor list objects as created by st_neighbors()
-   style: default "W" for row standardised weights.
-   allow_zero: if TRUE, assigns zero as lagged value to zone without neighbors.

## 3 Perform global Moran's I permutation test
Use Mote Carlo to perform the statistical test. For sfdep, it is supported by global_moran_perm()

**Step 1 - set seed**

```{r}
set.seed(1234)
```

**Step 2 - Monte Carlo simulation**
```{r}
global_moran_perm(wm_q$GDPPC,
                  wm_q$nb,
                  wm_q$wt,
                  nsim = 99)
```
::::: markgoals
::: markgoals-header
**Analysis**
:::

::: markgoals-container
-   The statistical report on previous tab shows that p-value is maller than alpha value of 0.05. There is enough statistical evidence to reject the null hypothesis that the distribution is random at 95% confidence level.
-   The Hunan province at the county level has signs of clustering. We can then further investigate whther is HH or LL clusters.
:::
:::::


