---
title: "06 - Global Measures of Spatial Autocorrelation"
author: "Cathy C."
date: "Oct 1, 2025"
date-modified: "last-modified" #auto render date
format: html
execute: #how YML page work
  echo: true #will show code
  eval: true #will run code every time and show output
  warning: false #dont show warning msg
  freeze: true #controls rendering - no render if no changes
---

## 1 Overview

We will learn how to delineate homogeneous region by using geographically referenced multivariate data. There are two major analysis, namely:

-   hierarchical cluster analysis
-   spatially constrained cluster analysis

By the end of this hands-on exercise, we will be able to do the following:

-   convert GIS polygon data into R’s simple feature data.frame by using appropriate functions of **sf** package of R;
-   convert simple feature data.frame into R’s SpatialPolygonDataFrame object by using appropriate **sf** of package of R;
-   perform custer analysis by using *hclust()* of Base R;
-   perform spatially constrained cluster analysis using *skater()* of Base R; and
-   visualise the analysis output by using **ggplot2** and **tmap** package.

## 2 Getting started

In geobusiness and spatial policy, it is a common practice to delineate the market or planning area into homogeneous regions by using multivariate data. In this hands-on exercise, we are interested to delineate [Shan State](https://en.wikipedia.org/wiki/Shan_State), [Myanmar](https://en.wikipedia.org/wiki/Myanmar) into homogeneous regions by using multiple Information and Communication technology (ICT) measures, namely: Radio, Television, Land line phone, Mobile phone, Computer, and Internet at home.

## 3 The data

### 3.1 Data sets

Two data sets will be used in this study. They are:

-   Myanmar Township Boundary Data (i.e. *myanmar_township_boundaries*) : This is a GIS data in ESRI shapefile format. It consists of township boundary information of Myanmar. The spatial data are captured in polygon features.
-   *Shan-ICT.csv*: This is an extract of [**The 2014 Myanmar Population and Housing Census Myanmar**](https://myanmar.unfpa.org/en/publications/2014-population-and-housing-census-myanmar-data-sheet) at the township level.

Both data sets are download from [Myanmar Information Management Unit (MIMU)](http://themimu.info/)

### 3.2 Installing and loading R packages

| R package | Purpose |
|----|----|
| **sf, rgdal, spdep** | Spatial data handling |
| **tidyverse - readr, ggplot2 and dplyr** | Attribute data handling |
| **tmap** | Choropleth mapping |
| **coorplot, ggpubr, heatmaply** | Multivariate data visualisation and analysis |
| **cluster, ClustGeo** | Cluster analysis |

```{r}
pacman::p_load(sf, rgdal, spdep, tidyverse, tmap, coorplot, ggpubr, heatmaply, cluster, ClustGeo,
               factoextra, NbClust, psych, GGally)
```

## 4 Data import and preparation

### 4.1 Importing geospatial data into R environment

We will import Myanmar Township Boundary GIS data and its associated attrbiute table into R environment.

The Myanmar Township Boundary GIS data is in ESRI shapefile format. It will be imported into R environment by using the [*st_read()*](https://www.rdocumentation.org/packages/sf/versions/0.7-2/topics/st_read) function of **sf**.

```{r}
shan_sf <- st_read(dsn = "data/geospatial",
                   layer = "myanmar_township_boundaries") %>%
  filter(ST %in% c("Shan (East)", "Shan (North)", "Shan (South)")) %>%
  select(c(2:7))
```

The imported township boundary object is called *shan_sf*. It is saved in **simple feature data.frame** format. We can view the content of the newly created *shan_sf* simple features data.frame with the code below.

```{r}
shan_sf
```

Notice that sf.data.frame is conformed to Hardy Wickham’s [tidy](https://edzer.github.io/rstudio_conf/#1) framework.

Since *shan_sf* is conformed to tidy framework, we can also *glimpse()* to reveal the data type of it’s fields.

```{r}
glimpse(shan_sf)
```

### 4.2 **Importing aspatial data into R environment**

The csv file will be import using *read_csv* function of **readr** package.

The code chunks used are shown below:

```{r}
ict <- read_csv("data/aspatial/Shan-ICT.csv")
```

The imported InfoComm variables are extracted from **The 2014 Myanmar Population and Housing Census Myanmar**. The attribute data set is called *ict*. It is saved in R’s *tibble data.frame* format.

```{r}
summary(ict)
```

There are a total of **11 fields** and **55 observations** in the tibble data.frame.

### 4.3 **Derive new variables using dplyr package**

The unit of measurement of the values are number of household. Using these values directly will be bias by the underlying total number of households. In general, the townships with relatively higher total number of households will also have higher number of households owning radio, TV, etc.

In order to overcome this problem, we will derive the penetration rate of each ICT variable by using the code chunk below.

```{r}
ict_derived <- ict %>%
  mutate(`RADIO_PR` = `Radio`/`Total households`*1000) %>%
  mutate(`TV_PR` = `Television`/`Total households`*1000) %>%
  mutate(`LLPHONE_PR` = `Land line phone`/`Total households`*1000) %>%
  mutate(`MPHONE_PR` = `Mobile phone`/`Total households`*1000) %>%
  mutate(`COMPUTER_PR` = `Computer`/`Total households`*1000) %>%
  mutate(`INTERNET_PR` = `Internet at home`/`Total households`*1000) %>%
  rename(`DT_PCODE` =`District Pcode`,`DT`=`District Name`,
         `TS_PCODE`=`Township Pcode`, `TS`=`Township Name`,
         `TT_HOUSEHOLDS`=`Total households`,
         `RADIO`=`Radio`, `TV`=`Television`, 
         `LLPHONE`=`Land line phone`, `MPHONE`=`Mobile phone`,
         `COMPUTER`=`Computer`, `INTERNET`=`Internet at home`) 
```

Review the summary statistics of the newly derived penetration rates.

```{r}
summary(ict_derived)
```

Now six fields have been added to the data.frame. They are RADIO_PR, TV_PR, LLPHONE_PR, MPHONE_PR, COMPUTER_PR, and INTERNET_PR.

## 5 Exploratory data analysis (EDA)

### 5.1 **EDA using statistical graphics**

We can plot the distributiob of the variables (ie. number of households with radio) by using appropriate EDA. Below histogram is useful to identify the overall distrbution of the data values (ie. left skew, right skew or normal distribution)

```{r}
#| fig-width: 12
ggplot(data = ict_derived,
       aes(x = `RADIO`)) +
  geom_histogram(bins = 20, color = "grey", fill = "#5D866C") +
  theme(panel.background = element_rect(fill = "#D0DDD0"))
  
```

Boxplot is useful to detect outliers.

```{r}
#| fig-width: 12
ggplot(data = ict_derived,
       aes(x = `RADIO`)) +
  geom_boxplot(color = "black", fill = "#5D866C") +
  theme(panel.background = element_rect(fill = "#D0DDD0"))
```

Next we will plot the distribution of the newly derived variables (ie. Radio penetration rate) with the code below.

```{r}
#| fig-width: 12
ggplot(data = ict_derived,
       aes(x = `RADIO_PR`)) +
  geom_histogram(bins = 20, color = "grey", fill = "#5D866C")
```

```{r}
#| fig-width: 12
ggplot(data = ict_derived,
       aes(x = `RADIO_PR`)) +
  geom_boxplot(color = "black", fill = "#5D866C")
```

In the figure below, multiple histograms are plotted to reveal the distribution of the selected variables in the ict_derived data.frame.

::: panel-tabset
## Multiplot

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 6
radio <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="grey60", 
                 fill="#B8C4A9")

tv <- ggplot(data=ict_derived, 
             aes(x= `TV_PR`)) +
  geom_histogram(bins=20, 
                 color="grey60", 
                 fill="#B8C4A9")

llphone <- ggplot(data=ict_derived, 
             aes(x= `LLPHONE_PR`)) +
  geom_histogram(bins=20, 
                 color="grey60", 
                 fill="#B8C4A9")

mphone <- ggplot(data=ict_derived, 
             aes(x= `MPHONE_PR`)) +
  geom_histogram(bins=20, 
                 color="grey60", 
                 fill="#B8C4A9")

computer <- ggplot(data=ict_derived, 
             aes(x= `COMPUTER_PR`)) +
  geom_histogram(bins=20, 
                 color="grey60", 
                 fill="#B8C4A9")

internet <- ggplot(data=ict_derived, 
             aes(x= `INTERNET_PR`)) +
  geom_histogram(bins=20, 
                 color="grey60", 
                 fill="#B8C4A9")

# Use ggarrange() of ggpubr package to group these histograms.

ggarrange(radio, tv, llphone, mphone, computer, internet,
          ncol = 3, nrow = 2)
```

## Code

``` r
#| fig-width: 12
#| fig-height: 6
radio <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="grey60", 
                 fill="#B8C4A9")

tv <- ggplot(data=ict_derived, 
             aes(x= `TV_PR`)) +
  geom_histogram(bins=20, 
                 color="grey60", 
                 fill="#B8C4A9")

llphone <- ggplot(data=ict_derived, 
             aes(x= `LLPHONE_PR`)) +
  geom_histogram(bins=20, 
                 color="grey60", 
                 fill="#B8C4A9")

mphone <- ggplot(data=ict_derived, 
             aes(x= `MPHONE_PR`)) +
  geom_histogram(bins=20, 
                 color="grey60", 
                 fill="#B8C4A9")

computer <- ggplot(data=ict_derived, 
             aes(x= `COMPUTER_PR`)) +
  geom_histogram(bins=20, 
                 color="grey60", 
                 fill="#B8C4A9")

internet <- ggplot(data=ict_derived, 
             aes(x= `INTERNET_PR`)) +
  geom_histogram(bins=20, 
                 color="grey60", 
                 fill="#B8C4A9")

# Use ggarrange() of ggpubr package to group these histograms.

ggarrange(radio, tv, llphone, mphone, computer, internet,
          ncol = 3, nrow = 2)
```
:::

## 6 Correlation analysis

## 7 Hierarchy cluster analysis

## 8 Spatially constrained clustering: SKATER approach

## 9 Spatially constrained clustering: ClustGeo method

## 10 Visual interpretation of clusters

## 11 Reference
