---
title: "06 - Global Measures of Spatial Autocorrelation"
author: "Cathy C."
date: "Oct 1, 2025"
date-modified: "last-modified" #auto render date
format: html
execute: #how YML page work
  echo: true #will show code
  eval: true #will run code every time and show output
  warning: false #dont show warning msg
  freeze: true #controls rendering - no render if no changes
---

## 1 Overview

We will learn how to delineate homogeneous region by using geographically referenced multivariate data. There are two major analysis, namely:

-   hierarchical cluster analysis
-   spatially constrained cluster analysis

By the end of this hands-on exercise, we will be able to do the following:

-   convert GIS polygon data into R’s simple feature data.frame by using appropriate functions of **sf** package of R;
-   convert simple feature data.frame into R’s SpatialPolygonDataFrame object by using appropriate **sf** of package of R;
-   perform custer analysis by using *hclust()* of Base R;
-   perform spatially constrained cluster analysis using *skater()* of Base R; and
-   visualise the analysis output by using **ggplot2** and **tmap** package.

## 2 Getting started

In geobusiness and spatial policy, it is a common practice to delineate the market or planning area into homogeneous regions by using multivariate data. In this hands-on exercise, we are interested to delineate [Shan State](https://en.wikipedia.org/wiki/Shan_State), [Myanmar](https://en.wikipedia.org/wiki/Myanmar) into homogeneous regions by using multiple Information and Communication technology (ICT) measures, namely: Radio, Television, Land line phone, Mobile phone, Computer, and Internet at home.

## 3 The data

### 3.1 Data sets

Two data sets will be used in this study. They are:

-   Myanmar Township Boundary Data (i.e. *myanmar_township_boundaries*) : This is a GIS data in ESRI shapefile format. It consists of township boundary information of Myanmar. The spatial data are captured in polygon features.
-   *Shan-ICT.csv*: This is an extract of [**The 2014 Myanmar Population and Housing Census Myanmar**](https://myanmar.unfpa.org/en/publications/2014-population-and-housing-census-myanmar-data-sheet) at the township level.

Both data sets are download from [Myanmar Information Management Unit (MIMU)](http://themimu.info/)

### 3.2 Installing and loading R packages

| R package | Purpose |
|------------------------------------|------------------------------------|
| **sf, rgdal, spdep** | Spatial data handling |
| **tidyverse - readr, ggplot2 and dplyr** | Attribute data handling |
| **tmap** | Choropleth mapping |
| **corrplot, ggpubr, heatmaply** | Multivariate data visualisation and analysis |
| **cluster, ClustGeo** | Cluster analysis |

```{r}
pacman::p_load(sf, rgdal, spdep, tidyverse, tmap, corrplot, ggpubr, heatmaply, cluster, ClustGeo,
               factoextra, NbClust, psych, GGally)
```

## 4 Data import and preparation

### 4.1 Importing geospatial data into R environment

We will import Myanmar Township Boundary GIS data and its associated attrbiute table into R environment.

The Myanmar Township Boundary GIS data is in ESRI shapefile format. It will be imported into R environment by using the [*st_read()*](https://www.rdocumentation.org/packages/sf/versions/0.7-2/topics/st_read) function of **sf**.

```{r}
shan_sf <- st_read(dsn = "data/geospatial",
                   layer = "myanmar_township_boundaries") %>%
  filter(ST %in% c("Shan (East)", "Shan (North)", "Shan (South)")) %>%
  select(c(2:7))
```

The imported township boundary object is called *shan_sf*. It is saved in **simple feature data.frame** format. We can view the content of the newly created *shan_sf* simple features data.frame with the code below.

```{r}
shan_sf
```

Notice that sf.data.frame is conformed to Hardy Wickham’s [tidy](https://edzer.github.io/rstudio_conf/#1) framework.

Since *shan_sf* is conformed to tidy framework, we can also *glimpse()* to reveal the data type of it’s fields.

```{r}
glimpse(shan_sf)
```

### 4.2 **Importing aspatial data into R environment**

The csv file will be import using *read_csv* function of **readr** package.

The code chunks used are shown below:

```{r}
ict <- read_csv("data/aspatial/Shan-ICT.csv")
```

The imported InfoComm variables are extracted from **The 2014 Myanmar Population and Housing Census Myanmar**. The attribute data set is called *ict*. It is saved in R’s *tibble data.frame* format.

```{r}
summary(ict)
```

There are a total of **11 fields** and **55 observations** in the tibble data.frame.

### 4.3 **Derive new variables using dplyr package**

The unit of measurement of the values are number of household. Using these values directly will be bias by the underlying total number of households. In general, the townships with relatively higher total number of households will also have higher number of households owning radio, TV, etc.

In order to overcome this problem, we will derive the penetration rate of each ICT variable by using the code chunk below.

```{r}
ict_derived <- ict %>%
  mutate(`RADIO_PR` = `Radio`/`Total households`*1000) %>%
  mutate(`TV_PR` = `Television`/`Total households`*1000) %>%
  mutate(`LLPHONE_PR` = `Land line phone`/`Total households`*1000) %>%
  mutate(`MPHONE_PR` = `Mobile phone`/`Total households`*1000) %>%
  mutate(`COMPUTER_PR` = `Computer`/`Total households`*1000) %>%
  mutate(`INTERNET_PR` = `Internet at home`/`Total households`*1000) %>%
  rename(`DT_PCODE` =`District Pcode`,`DT`=`District Name`,
         `TS_PCODE`=`Township Pcode`, `TS`=`Township Name`,
         `TT_HOUSEHOLDS`=`Total households`,
         `RADIO`=`Radio`, `TV`=`Television`, 
         `LLPHONE`=`Land line phone`, `MPHONE`=`Mobile phone`,
         `COMPUTER`=`Computer`, `INTERNET`=`Internet at home`) 
```

Review the summary statistics of the newly derived penetration rates.

```{r}
summary(ict_derived)
```

Now six fields have been added to the data.frame. They are RADIO_PR, TV_PR, LLPHONE_PR, MPHONE_PR, COMPUTER_PR, and INTERNET_PR.

## 5 Exploratory data analysis (EDA)

### 5.1 **EDA using statistical graphics**

We can plot the distributiob of the variables (ie. number of households with radio) by using appropriate EDA. Below histogram is useful to identify the overall distrbution of the data values (ie. left skew, right skew or normal distribution)

```{r}
#| fig-width: 12
ggplot(data = ict_derived,
       aes(x = `RADIO`)) +
  geom_histogram(bins = 20, color = "grey", fill = "#5D866C") +
  theme(panel.background = element_rect(fill = "#D0DDD0"))
  
```

Boxplot is useful to detect outliers.

```{r}
#| fig-width: 12
ggplot(data = ict_derived,
       aes(x = `RADIO`)) +
  geom_boxplot(color = "black", fill = "#5D866C") +
  theme(panel.background = element_rect(fill = "#D0DDD0"))
```

Next we will plot the distribution of the newly derived variables (ie. Radio penetration rate) with the code below.

```{r}
#| fig-width: 12
ggplot(data = ict_derived,
       aes(x = `RADIO_PR`)) +
  geom_histogram(bins = 20, color = "grey", fill = "#5D866C")
```

```{r}
#| fig-width: 12
ggplot(data = ict_derived,
       aes(x = `RADIO_PR`)) +
  geom_boxplot(color = "black", fill = "#5D866C")
```

In the figure below, multiple histograms are plotted to reveal the distribution of the selected variables in the ict_derived data.frame.

::: panel-tabset
## Multiplot

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 6
radio <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="grey60", 
                 fill="#B8C4A9")

tv <- ggplot(data=ict_derived, 
             aes(x= `TV_PR`)) +
  geom_histogram(bins=20, 
                 color="grey60", 
                 fill="#B8C4A9")

llphone <- ggplot(data=ict_derived, 
             aes(x= `LLPHONE_PR`)) +
  geom_histogram(bins=20, 
                 color="grey60", 
                 fill="#B8C4A9")

mphone <- ggplot(data=ict_derived, 
             aes(x= `MPHONE_PR`)) +
  geom_histogram(bins=20, 
                 color="grey60", 
                 fill="#B8C4A9")

computer <- ggplot(data=ict_derived, 
             aes(x= `COMPUTER_PR`)) +
  geom_histogram(bins=20, 
                 color="grey60", 
                 fill="#B8C4A9")

internet <- ggplot(data=ict_derived, 
             aes(x= `INTERNET_PR`)) +
  geom_histogram(bins=20, 
                 color="grey60", 
                 fill="#B8C4A9")

# Use ggarrange() of ggpubr package to group these histograms.

ggarrange(radio, tv, llphone, mphone, computer, internet,
          ncol = 3, nrow = 2)
```

## Code

``` r
#| fig-width: 12
#| fig-height: 6
radio <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="grey60", 
                 fill="#B8C4A9")

tv <- ggplot(data=ict_derived, 
             aes(x= `TV_PR`)) +
  geom_histogram(bins=20, 
                 color="grey60", 
                 fill="#B8C4A9")

llphone <- ggplot(data=ict_derived, 
             aes(x= `LLPHONE_PR`)) +
  geom_histogram(bins=20, 
                 color="grey60", 
                 fill="#B8C4A9")

mphone <- ggplot(data=ict_derived, 
             aes(x= `MPHONE_PR`)) +
  geom_histogram(bins=20, 
                 color="grey60", 
                 fill="#B8C4A9")

computer <- ggplot(data=ict_derived, 
             aes(x= `COMPUTER_PR`)) +
  geom_histogram(bins=20, 
                 color="grey60", 
                 fill="#B8C4A9")

internet <- ggplot(data=ict_derived, 
             aes(x= `INTERNET_PR`)) +
  geom_histogram(bins=20, 
                 color="grey60", 
                 fill="#B8C4A9")

# Use ggarrange() of ggpubr package to group these histograms.

ggarrange(radio, tv, llphone, mphone, computer, internet,
          ncol = 3, nrow = 2)
```
:::

### 5.2 **EDA using choropleth map**

#### 5.2.1 Joining geospatial data with aspatial data

Before we can prepare the choropleth map, we need to combine both the geospatial data object (i.e. *shan_sf*) and aspatial data.frame object (i.e. *ict_derived*) into one. This will be performed by using the [*left_join*](https://dplyr.tidyverse.org/reference/join.tbl_df.html) function of **dplyr** package. The *shan_sf* simple feature data.frame will be used as the base data object and the *ict_derived* data.frame will be used as the join table.

Below is to perform the task. The unique identifier used to join both data objects is *TS_PCODE*.

```{r}
#| eval: false
shan_sf <- left_join(shan_sf, 
                     ict_derived, by=c("TS_PCODE"="TS_PCODE"))
  
write_rds(shan_sf, "data/rds/shan_sf.rds")
```

```{r}
shan_sf <- read_rds("data/rds/shan_sf.rds")
```

```{r}
summary(shan_sf)
```

#### 5.2.2 Preparing a choropleth map

To have a quick look at the distribution of Radio penetration rate of Shan State at township level, a choropleth map will be prepared.

The code chunks below are used to prepare the choroplethby using the *qtm()* function of **tmap** package.

```{r}
qtm(shan_sf, "RADIO_PR.y")
```

In order to reveal the distribution shown in the choropleth map above are bias to the underlying total number of households at the townships, we will create two choropleth maps.

One is for the **total number of households** (i.e. TT_HOUSEHOLDS.map) and the other is for the **total number of household with Radio (RADIO.map)** by using the code chunk below.

```{r}
#| fig-width: 12
#| fig-height: 8
TT_HOUSEHOLDS.map <- tm_shape(shan_sf) + 
  tm_fill(col = "TT_HOUSEHOLDS.y",
          n = 5,
          style = "jenks", 
          title = "Total households",
          palette = "Greens") + 
  tm_borders(alpha = 0.5) +
  tm_text("TS.y", size = 0.5) +
  tm_layout(legend.position = c("right", "top"))

RADIO.map <- tm_shape(shan_sf) + 
  tm_fill(col = "RADIO.y",
          n = 5,
          style = "jenks",
          title = "Number Radio ",
          palette = "Greens") + 
  tm_borders(alpha = 0.5) +
  tm_text("TS.y", size = 0.5) +
  tm_layout(legend.position = c("right", "top"))

tmap_arrange(TT_HOUSEHOLDS.map, RADIO.map,
             asp=NA, ncol=2)
```

::::: markgoals
::: markgoals-header
**Learning**
:::

::: markgoals-container
The choropleth maps above clearly show that townships with relatively larger number of households are also showing relatively higher number of radio ownership.
:::
:::::

Now let us plot the choropleth maps showing the dsitribution of total number of households and Radio penetration rate.

```{r}
#| fig-width: 12
#| fig-height: 8
tm_shape(shan_sf) +
    tm_polygons(c("TT_HOUSEHOLDS.y", "RADIO_PR.y"),
                style="jenks") +
    tm_text("TS.y", size = 0.5) +
    tm_facets(sync = TRUE, ncol = 2) +
#  tm_legend(legend.position = c("right", "bottom"))+
  tm_layout(legend.position = c("right", "bottom"), outer.margins=0.01, asp=0)
```

::::: markgoals
::: markgoals-header
**Analysis**
:::

::: markgoals-container
-   In the previous visualisation, we noticed that the Radio ownership is positively correlated, meaning that the more households in the township, the more radios there are in this area.
-   Now we compare the total household against the Radio penetration, we see the some township has apparently higher penetration rate while their population isn's as dense. These townships include Mabein in the northwest, Ywangan in the west, and Monghpyak in the southeast.
-   The higher Radio penetration areas concentrate on the townships next to Laos and other domestic states in the west. Townships on the east side next to China have the lowest Radio penetration rate. These townships next to China might be less developed as compared to others.
:::
:::::

## 6 Correlation analysis

Before we perform cluster analysis, it is important for us to ensure that the cluster variables are not highly correlated.

In this section, we will use [*corrplot.mixed()*](https://cran.r-project.org/web/packages/corrplot/corrplot.pdf) function of [**corrplot**](https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html) package to visualise and analyse the correlation of the input variables.

```{r}
#| fig-width: 12
#| fig-height: 8
par(bg = "#A7C1A8")
cluster_vars.cor = cor(ict_derived[,12:17])
corrplot.mixed(cluster_vars.cor,
         lower = "ellipse", 
         upper = "number",
         bg = "#DCDCDC",
         tl.pos = "lt",
         tl.srt = 45,
         diag = "l",
         tl.col = "black")
```

::::: markgoals
::: markgoals-header
**Analysis**
:::

::: markgoals-container
The correlation plot above shows that COMPUTER_PR and INTERNET_PR are highly correlated. This suggest that only one of them should be used in the cluster analysis instead of both.
:::
:::::

## 7 Hierarchy cluster analysis

In this section, we will perform hierarchical cluster analysis. The analysis consists of four major steps:

### 7.1 Extracting clustering variables

Below we will extract the clustering variables from the shan_sf simple feature object into data.frame. COMPUTER_PR is included, but not INTERNET_PR, because they are highly correlated.

```{r}

cluster_vars <- shan_sf %>%
  st_set_geometry(NULL) %>%
  select("TS.x", "RADIO_PR.x", "TV_PR.x", "LLPHONE_PR.x", "MPHONE_PR.x", "COMPUTER_PR.x")
head(cluster_vars,10)
```

Next we will change the rows by township name instead of row number:

```{r}

row.names(cluster_vars) <- cluster_vars$"TS.x"
head(cluster_vars,10)
```

We see that the numbers have been replaced with township names. Next we will delete TS.x field:

```{r}
shan_ict <- select(cluster_vars, c(2:6))
head(shan_ict, 10)
```

It's clearly TS.x is removed, and we will save this to a rds file.

```{r}
shan_ict <- write_rds(shan_ict, "data/rds/shan_ict.rds")
```

```{r}
shan_ict <- read_rds("data/rds/shan_ict.rds")
```

### 7.2 Data Standardisation

In general, multiple variables will be used in cluster analysis. It is not unusual their values range are different. In order to avoid the cluster analysis result being biased to clustering variables with large values, it is useful to standardise the input variables before performing cluster analysis.

### 7.3 Min-Max standardisation

*`normalize()`* of [*heatmaply*](https://cran.r-project.org/web/packages/heatmaply/index.html) package is used to stadardisation the clustering variables by using Min-Max method. The *summary()* is then used to display the summary statistics of the standardised clustering variables.

```{r}
shan_ict.std <- normalize(shan_ict)
summary(shan_ict.std)
```

Note that the values range of the Min-max standardised clustering variables are 0-1 now.

### 7.4 Z-score standardisation

Z-score standardisation can be performed easily by using [*scale()*](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/scale) of Base R. The code chunk below will be used to stadardisation the clustering variables by using Z-score method.

```{r}
shan_ict.z <- scale(shan_ict)
describe(shan_ict.z)
```

::::: markgoals
::: markgoals-header
**Note**
:::

::: markgoals-container
-   Note that the mean and standard deviation of the Z-score standardised clustering variables are 0 and 1 respectively.
-   [*describe()*](https://www.rdocumentation.org/packages/Hmisc/versions/4.4-0/topics/describe) of [**psych**](https://cran.r-project.org/web/packages/psych/index.html) package is used here instead of *summary()* of Base R because the earlier provides standard deviation.
-   [**Z-score standardisation method should only be used if we would assume all variables come from some normal distribution.**]{style="color: #B95E82;"}
:::
:::::

### 7.5 Visualising the standardised clustering variables

Beside reviewing the summary statistics of the standardised clustering variables, it is also a good practice to visualise their distribution graphical.

The code chunk below plot the scaled *Radio_PR* field with histogram and density plots.

::: panel-tabset
## Histogram

```{r}
#| fig-width: 12
#| fig-height: 4
r <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="grey70", 
                 fill="#B8C4A9") +
  ggtitle("Raw values without standardisation")

shan_ict_s_df <- as.data.frame(shan_ict.std)
s <- ggplot(data=shan_ict_s_df, 
       aes(x=`RADIO_PR.x`)) +
  geom_histogram(bins=20, 
                 color="grey70", 
                 fill="#B8C4A9") +
  ggtitle("Min-Max Standardisation")

shan_ict_z_df <- as.data.frame(shan_ict.z)
z <- ggplot(data=shan_ict_z_df, 
       aes(x=`RADIO_PR.x`)) +
  geom_histogram(bins=20, 
                 color="grey70", 
                 fill="#B8C4A9") +
  ggtitle("Z-score Standardisation")

ggarrange(r, s, z,
          ncol = 3,
          nrow = 1)
```

## DensityPlot

```{r}
#| fig-width: 12
#| fig-height: 4
r <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_density(color="black",
               fill="light blue") +
  ggtitle("Raw values without standardisation")

shan_ict_s_df <- as.data.frame(shan_ict.std)
s <- ggplot(data=shan_ict_s_df, 
       aes(x=`RADIO_PR.x`)) +
  geom_density(color="black",
               fill="light blue") +
  ggtitle("Min-Max Standardisation")

shan_ict_z_df <- as.data.frame(shan_ict.z)
z <- ggplot(data=shan_ict_z_df, 
       aes(x=`RADIO_PR.x`)) +
  geom_density(color="black",
               fill="light blue") +
  ggtitle("Z-score Standardisation")

ggarrange(r, s, z,
          ncol = 3,
          nrow = 1)
```
:::

### 7.6 Computing proximity matrix

In R, many packages provide functions to calculate distance matrix. We will compute the proximity matrix by using [*dist()*](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/dist.html) of R.

*dist()* supports six distance proximity calculations, they are: **euclidean, maximum, manhattan, canberra, binary and minkowski**. The default is *euclidean* proximity matrix.

Below is to compute the proximity matrix using *euclidean* and *manhattan* method.

::: panel-tabset
## euclidean

```{r}
proxmat <- dist(shan_ict, method = 'euclidean')
proxmat
```

## manhattan

```{r}
proxmat <- dist(shan_ict, method = 'manhattan')
proxmat
```
:::

### 7.7 Computing hierarchical clustering

In R, several packages provide hierarchical clustering function. In this hands-on exercise, [*hclust()*](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/hclust.html) of R stats will be used.

*hclust()* employed agglomeration method to compute the cluster. Eight clustering algorithms are supported, they are: ward.D, ward.D2, single, complete, average(UPGMA), mcquitty(WPGMA), median(WPGMC) and centroid(UPGMC).

Below performs hierarchical cluster analysis using **ward.D method**. The hierarchical clustering output is stored in an object of class **hclust** which describes the tree produced by the clustering process.

```{r}
hclust_ward <- hclust(proxmat, method = 'ward.D')
```

Next we plot the tree using plot() of R graphics.

```{r}
#| fig-width: 12
#| fig-height: 8
par(bg = "#F7F4EA")
plot(hclust_ward, cex = 0.6)
```

### 7.8 Selecting the optimal clustering algorithm

One of the challenge in performing hierarchical clustering is to identify stronger clustering structures. The issue can be solved by using use [*agnes()*](https://www.rdocumentation.org/packages/cluster/versions/2.1.0/topics/agnes) function of [**cluster**](https://cran.r-project.org/web/packages/cluster/) package. It functions like *hclus()*, however, with the *agnes()* function you can also get the agglomerative coefficient, which measures the amount of clustering structure found (values closer to 1 suggest strong clustering structure).

Below computes the agglomerative coefficients of all hierarchical clustering algorithms.

```{r}
m <- c( "average", "single", "complete", "ward")
names(m) <- c( "average", "single", "complete", "ward")

ac <- function(x) {
  agnes(shan_ict, method = x)$ac
}

map_dbl(m, ac)
```

With reference to the output above, we can see that Ward’s method provides the strongest clustering structure among the four methods assessed. Hence, in the subsequent analysis, only Ward’s method will be used.

### 7.9 Determining Optimal Clusters

Another technical challenge face by data analyst in performing clustering analysis is to determine the optimal clusters to retain.

There are [three](https://statweb.stanford.edu/~gwalther/gap) commonly used methods to determine the optimal clusters, they are:

-   [Elbow Method](https://en.wikipedia.org/wiki/Elbow_method_(clustering))

-   [Average Silhouette Method](https://www.sciencedirect.com/science/article/pii/0377042787901257?via%3Dihub)

-   [Gap Statistic Method](http://www.web.stanford.edu/~hastie/Papers/gap.pdf)

::: panel-tabset
## Gap Statistic Method

The [**gap statistic**](http://www.web.stanford.edu/~hastie/Papers/gap.pdf) compares the total within intra-cluster variation for different values of k with their expected values under null reference distribution of the data. The estimate of the optimal clusters will be value that maximize the gap statistic (i.e., that yields the largest gap statistic). This means that the clustering structure is far away from the random uniform distribution of points.

To compute the gap statistic, [*clusGap()*](https://www.rdocumentation.org/packages/cluster/versions/2.1.0/topics/clusGap) of [**cluster**](https://cran.r-project.org/web/packages/cluster/index.html) package will be used. The [*hcut*](https://rpkgs.datanovia.com/factoextra/reference/hcut.html) function used is from [**factoextra**](https://rpkgs.datanovia.com/factoextra/index.html) package.

```{r}
set.seed(12345)
gap_stat <- clusGap(shan_ict, 
                    FUN = hcut, 
                    nstart = 25, 
                    K.max = 10, 
                    B = 50)
# Print the result
print(gap_stat, method = "firstmax")
```

## Elbow Method

```{r}

```

## Average Silhouette Method
:::

Next, we can visualise the plot by using [*fviz_gap_stat()*](https://rpkgs.datanovia.com/factoextra/reference/fviz_nbclust.html) of [**factoextra**](https://rpkgs.datanovia.com/factoextra/index.html) package.

```{r}
#| fig-width: 12
#| fig-height: 6
fviz_gap_stat(gap_stat) +
  theme(plot.background = element_rect(fill = "#F7F4EA"),
        panel.background = element_rect(fill = "#F7F4EA"))
```

With reference to the gap statistic graph above, the recommended number of cluster to retain is 1. However, it is not logical to retain only one cluster. By examine the gap statistic graph, [the 6-cluster gives the largest gap statistic and should be the next best cluster to pick]{style="color: #B95E82;"}.

**Note:** In addition to these commonly used approaches, the [NbClust](https://cran.r-project.org/web/packages/NbClust/) package, published by Charrad et al., 2014, provides 30 indices for determining the relevant number of clusters and proposes to users the best clustering scheme from the different results obtained by varying all combinations of number of clusters, distance measures, and clustering methods.

### 7.10 Interpreting the dendrograms

In the dendrogram displayed above, each leaf corresponds to one observation. As we move up the tree, observations that are similar to each other are combined into branches, which are themselves fused at a higher height.

The height of the fusion, provided on the vertical axis, indicates the (dis)similarity between two observations. The higher the height of the fusion, the less similar the observations are. Note that, conclusions about the proximity of two observations can be drawn only based on the height where branches containing those two observations first are fused. We cannot use the proximity of two observations along the horizontal axis as a criteria of their similarity.

It’s also possible to draw the dendrogram with a border around the selected clusters by using [*rect.hclust()*](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/rect.hclust.html) of R stats. The argument *border* is used to specify the border colors for the rectangles.

```{r}
#| fig-width: 12
#| fig-height: 8
par(bg = "#F7F4EA")

plot(hclust_ward, cex = 0.6)
rect.hclust(hclust_ward, 
            k = 6, 
            border = 2:5)
```

### 7.11 Visually-driven hierarchical clustering analysis

In this section, we will perform visually-driven hiearchical clustering analysis by using [*heatmaply*](https://cran.r-project.org/web/packages/heatmaply/index.html) package.

With **heatmaply**, we are able to build both highly interactive cluster heatmap or static cluster heatmap.

### 7.11.1 Transforming the data frame into a matrix

The data was loaded into a data frame, but **it has to be a data matrix** to make a heatmap. Below will transform *shan_ict* data frame into a data matrix.

```{r}
shan_ict_mat <- data.matrix(shan_ict)
```

### 7.11.2 Plotting interactive cluster heatmap using *heatmaply()*

[*heatmaply()*](https://talgalili.github.io/heatmaply/reference/heatmaply.html) of [heatmaply](https://talgalili.github.io/heatmaply/index.html) package is used to build an interactive cluster heatmap.

```{r}
#| fig-width: 12
#| fig-height: 10
p <- heatmaply(normalize(shan_ict_mat),
          Colv=NA,
          dist_method = "euclidean",
          hclust_method = "ward.D",
          seriate = "OLO",
          colors = Blues,
          k_row = 6,
          margins = c(NA,200,60,NA),
          fontsize_row = 4,
          fontsize_col = 5,
          main="Geographic Segmentation of Shan State by ICT indicators",
          xlab = "ICT Indicators",
          ylab = "Townships of Shan State"
          ) %>%
  layout(plot_bgcolor = "#D0DDD0",
         paper_bgcolor = "#D0DDD0")
p
```

### 7.12 Mapping the clusters formed

With closed examination of the dendragram above, we have decided to retain six clusters.

[*cutree()*](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/cutree.html) of R Base will be used in the code chunk below to derive a 6-cluster model.

```{r}
groups <- as.factor(cutree(hclust_ward, k=6))
```

The output is called *groups*. It is a *list* object.

In order to visualise the clusters, the *groups* object need to be appended onto *shan_sf* simple feature object.

The code chunk below form the join in three steps:

-   the *groups* list object will be converted into a matrix;
-   *cbind()* is used to append *groups* matrix onto shan_sf to produce an output simple feature object called `shan_sf_cluster`; and
-   *rename* of **dplyr** package is used to rename *as.matrix.groups* field as *CLUSTER*.

```{r}
shan_sf_cluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER`=`as.matrix.groups.`)
```

Next, *qtm()* of **tmap** package is used to plot the choropleth map showing the cluster formed.

```{r}
#| fig-width: 12
#| fig-height: 8
par(bg = "#D0DDD0")
qtm(shan_sf_cluster, "CLUSTER") +
    tm_layout(bg.color = "#D0DDD0",
              outer.margins = c(0.02, 0.02, 0.02, 0.02),
              legend.position = c("right", "top"),
              frame = FALSE)
```

## 8 Spatially constrained clustering: SKATER approach

## 9 Spatially constrained clustering: ClustGeo method

## 10 Visual interpretation of clusters

## 11 Reference
