---
title: "In-class 07 - "
author: "Cathy C."
date: "Oct 18, 2025"
date-modified: "last-modified" #auto render date
format: html
execute: #how YML page work
  echo: true #will show code
  eval: true #will run code every time and show output
  warning: false #dont show warning msg
  freeze: true #controls rendering - no render if no changes
---

## 1 Getting started

We will be using the following R packages:

| R Package | Description |
|------------------------------------|------------------------------------|
| [**olsrr**](https://olsrr.rsquaredacademy.com/) | R package for building OLS and performing diagnostics tests |
| [**GWmodel**](https://cran.r-project.org/web/packages/GWmodel/index.html) | R package for calibrating geographical weighted family of models |
| [**corrplot**](https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html) | R package for multivariate data visualisation and analysis |
| **sf** | Spatial data handling |
| **tidyverse** | Attribute data handling with **readr**, **ggplot2** and **dplyr** |
| **tmap** | Do Choropleth mapping |
| **gtsummary** | provides an elegant and flexible way to create publication-ready analytical and summary tables. he {gtsummary} package summarizes data sets, regression models. |
| **easystats** | R framework for easy statistical modeling |

```{r}
pacman::p_load(olsrr, corrplot, ggpubr, sf, sfdep, GWmodel, tmap, tidyverse, gtsummary,
               performance, RColorBrewer)
```

## 2 The data

### 2.1 Data set

Two data sets will be used in this model building exercise, they are:

-   URA Master Plan subzone boundary in shapefile format (i.e. *MP14_SUBZONE_WEB_PL*)
-   condo_resale_2015 in csv format (i.e. *condo_resale_2015.csv*)

### 2.2 Geospatial data wrangling

#### 2.2.1 Import geospatial data

The geospatial data used in this hands-on exercise is called MP14_SUBZONE_WEB_PL. It is in ESRI shapefile format. The shapefile consists of URA Master Plan 2014’s planning subzone boundaries. Polygon features are used to represent these geographic boundaries. The GIS data is in svy21 projected coordinates systems.

The code chunk below is used to import *MP_SUBZONE_WEB_PL* shapefile by using `st_read()` of **sf** packages.

```{r}
mpsz = st_read(dsn = "data/geospatial", layer = "MP14_SUBZONE_WEB_PL")
```

#### 4.2.2 Update CRS information

The code chunk below updates the newly imported *mpsz* with the **correct ESPG code (i.e. 3414)**

```{r}
mpsz_svy21 <- st_transform(mpsz, 3414)
```

Let's verify the newly transformed *mpsz_svy21*.

```{r}
st_crs(mpsz_svy21)
```

Notice that the EPSG: **is indicated as *3414* now.**

Next, we will reveal the extent of *mpsz_svy21* by using `st_bbox()` of sf package. The print below reports the extent of mpsz_svy21 layer by its l**ower and upper limits**.

```{r}
st_bbox(mpsz_svy21)
```

### 3.3 Aspatial data wrangling

#### 3.3.1 Import Aspatial data

The *condo_resale_2015* is in csv file format. The codes chunk below uses `read_csv()` function of **readr** package to import *condo_resale_2015* into R as a tibble data frame called *condo_resale*.

```{r}
condo_resale = read_csv("data/aspatial/Condo_resale_2015.csv")
```

The print above reveals that the values of *LONGITITUDE* and *LATITUDE* fields are in decimal degree. Most probably wgs84 geographic coordinate system is used.

#### 3.3.2 Convert aspatial data frame into a sf object

Currently, the *condo_resale* tibble data frame is aspatial. We will convert it to a **sf** object. The code chunk below converts condo_resale data frame into a simple feature data frame by using `st_as_sf()` of **sf** packages. Specify Longitude before Latitude.

```{r}
condo_resale.sf <- st_as_sf(condo_resale,
                            coords = c("LONGITUDE", "LATITUDE"),
                            crs=4326) %>%
  st_transform(crs=3414)
```

Let's list the content of *condo_resale.sf* object. Notice that the output is in point feature data frame.

```{r}
head(condo_resale.sf)
```

## 4 Use the package GWmodel to determine the bandwidth.

Generalised GWR model: `bw.ggwr()` Normal bandwidth: `bw.gwr()`

### 4.1 Building Fixed Bandwidth GWR Model

#### 4.1.1 Computing fixed bandwidth

In the code chunk below `bw.gwr()` of GWModel package is used to determine the **optimal fixed bandwidt**h to use in the model. Notice that the argument ***adaptive*** is set to **FALSE** indicates that we are interested to compute the fixed bandwidth.

There are two possible approaches can be uused to determine the stopping rule, they are: **CV cross-validation approach** and **AIC corrected (AICc) approach**. We define the stopping rule using ***approach*** argeement. Below we will use AIC, with comparison with CV on the other tab.

::: panel-tabset
## AIC

```{r}
bw.fixed_aic <- bw.gwr(
  formula = SELLING_PRICE ~ AREA_SQM + AGE  + 
    PROX_CBD + PROX_CHILDCARE + PROX_ELDERLYCARE    +
    PROX_URA_GROWTH_AREA + PROX_MRT + PROX_PARK + 
    PROX_PRIMARY_SCH + PROX_SHOPPING_MALL   + 
    PROX_BUS_STOP   + NO_Of_UNITS + FAMILY_FRIENDLY + 
    FREEHOLD, 
  data=condo_resale.sf, 
  approach="AICc", 
  kernel="gaussian", 
  adaptive=FALSE, 
  longlat=FALSE)
```

#### 4.1.2 GWModel method - fixed bandwidth

Now we can use the code chunk below to calibrate the gwr model using fixed bandwidth and gaussian kernel.

```{r}
gwr.fixed_aic <- gwr.basic(
  formula = SELLING_PRICE ~ AREA_SQM + AGE  + 
    PROX_CBD + PROX_CHILDCARE + PROX_ELDERLYCARE    +
    PROX_URA_GROWTH_AREA + PROX_MRT + PROX_PARK +
    PROX_PRIMARY_SCH + PROX_SHOPPING_MALL   + 
    PROX_BUS_STOP   + NO_Of_UNITS + FAMILY_FRIENDLY + 
    FREEHOLD, 
  data=condo_resale.sf, 
  bw=bw.fixed_aic, #this is from previous
  kernel = 'gaussian', #gaussian is the same as bw. need to be the same.
  longlat = FALSE)
```

The output is saved in a list of class “gwrm”. The code below can be used to display the model output.

```{r}
gwr.fixed_aic
```

## CV

```{r}
bw.fixed <- bw.gwr(
  formula = SELLING_PRICE ~ AREA_SQM + AGE  + 
    PROX_CBD + PROX_CHILDCARE + PROX_ELDERLYCARE    +
    PROX_URA_GROWTH_AREA + PROX_MRT + PROX_PARK + 
    PROX_PRIMARY_SCH + PROX_SHOPPING_MALL   + 
    PROX_BUS_STOP   + NO_Of_UNITS + FAMILY_FRIENDLY + 
    FREEHOLD, 
  data=condo_resale.sf, 
  approach="CV", 
  kernel="gaussian", 
  adaptive=FALSE, 
  longlat=FALSE)
```

Now we can use the code chunk below to calibrate the gwr model using fixed bandwidth and gaussian kernel.

```{r}
gwr.fixed <- gwr.basic(
  formula = SELLING_PRICE ~ AREA_SQM + AGE  + 
    PROX_CBD + PROX_CHILDCARE + PROX_ELDERLYCARE    +
    PROX_URA_GROWTH_AREA + PROX_MRT + PROX_PARK +
    PROX_PRIMARY_SCH + PROX_SHOPPING_MALL   + 
    PROX_BUS_STOP   + NO_Of_UNITS + FAMILY_FRIENDLY + 
    FREEHOLD, 
  data=condo_resale.sf, 
  bw=bw.fixed, #this is from previous
  kernel = 'gaussian', #gaussian is the same as bw. need to be the same.
  longlat = FALSE)
```

The output is saved in a list of class “gwrm”. The code below can be used to display the model output.

```{r}
gwr.fixed
```
:::

Based on the AICc results derived above and section 4.2, the AIC approach for adaptive bandwidth has the smallest value (**41978.63)**. Therefore, we conclude that using the AIC approach for Adaptive bandwidth yields the best result among the four approaches.

| AICc value         | AIC            | CV       |
|--------------------|----------------|----------|
| Fixed Bandwidth    | 42254.7        | 42263.6  |
| Adaptive Bandwidth | **41978.63\*** | 41982.22 |

### 4.2 Building Adaptive Bandwidth GWR Model

In this section, we will calibrate the gwr-based hedonic pricing model by using adaptive bandwidth approach.

#### 4.2.1 Computing the adaptive bandwidth

Similar to the earlier section, we will first use `bw.gwr()` to determine the recommended data point to use.

The code chunk used look very similar to the one used to compute the fixed bandwidth except the `adaptive` argument has changed to **TRUE**.

:::::: panel-tabset
## AIC

```{r}
bw.adaptive_aic <- bw.gwr(
  formula = SELLING_PRICE ~ AREA_SQM + AGE  + 
    PROX_CBD + PROX_CHILDCARE + PROX_ELDERLYCARE    +
    PROX_URA_GROWTH_AREA + PROX_MRT + PROX_PARK +
    PROX_PRIMARY_SCH + PROX_SHOPPING_MALL   + 
    PROX_BUS_STOP   + NO_Of_UNITS + 
    FAMILY_FRIENDLY + FREEHOLD, 
  data=condo_resale.sf, 
  approach="AIC", 
  kernel="gaussian", 
  adaptive=TRUE, 
  longlat=FALSE)
```

::::: markgoals
::: markgoals-header
**Insight**
:::

::: markgoals-container
The result shows that the 27 is the recommended data points to be used.
:::
:::::

#### 4.2.2 Constructing the adaptive bandwidth gwr model

Now, we can go ahead to calibrate the gwr-based hedonic pricing model by using adaptive bandwidth and gaussian kernel as shown in the code chunk below.

```{r}
gwr.adaptive_aic <- gwr.basic(
  formula = SELLING_PRICE ~ AREA_SQM + AGE + 
    PROX_CBD + PROX_CHILDCARE + PROX_ELDERLYCARE +
    PROX_URA_GROWTH_AREA + PROX_MRT + PROX_PARK +
    PROX_PRIMARY_SCH + PROX_SHOPPING_MALL + PROX_BUS_STOP +
    NO_Of_UNITS + FAMILY_FRIENDLY + FREEHOLD,
  data=condo_resale.sf, 
  bw=bw.adaptive_aic, 
  kernel = 'gaussian', 
  adaptive=TRUE, 
  longlat = FALSE)
```

Let's display the model output.

```{r}
gwr.adaptive_aic
```

## CV

```{r}
bw.adaptive <- bw.gwr(
  formula = SELLING_PRICE ~ AREA_SQM + AGE  + 
    PROX_CBD + PROX_CHILDCARE + PROX_ELDERLYCARE    +
    PROX_URA_GROWTH_AREA + PROX_MRT + PROX_PARK +
    PROX_PRIMARY_SCH + PROX_SHOPPING_MALL   + 
    PROX_BUS_STOP   + NO_Of_UNITS + 
    FAMILY_FRIENDLY + FREEHOLD, 
  data=condo_resale.sf, 
  approach="CV", 
  kernel="gaussian", 
  adaptive=TRUE, 
  longlat=FALSE)
```

Now, we can go ahead to calibrate the gwr-based hedonic pricing model by using adaptive bandwidth and gaussian kernel as shown in the code chunk below.

```{r}
gwr.adaptive <- gwr.basic(
  formula = SELLING_PRICE ~ AREA_SQM + AGE + 
    PROX_CBD + PROX_CHILDCARE + PROX_ELDERLYCARE +
    PROX_URA_GROWTH_AREA + PROX_MRT + PROX_PARK +
    PROX_PRIMARY_SCH + PROX_SHOPPING_MALL + PROX_BUS_STOP +
    NO_Of_UNITS + FAMILY_FRIENDLY + FREEHOLD,
  data=condo_resale.sf, 
  bw=bw.adaptive, 
  kernel = 'gaussian', 
  adaptive=TRUE, 
  longlat = FALSE)
```

Let's display the model output.

```{r}
gwr.adaptive
```
::::::

### 4.3 **Converting SDF into *sf* data.frame**

To visualise the fields in **SDF**, we need to first covert it into **sf** data.frame by using the code chunk below.

```{r}
condo_resale.sf.adaptive_aic <-
  st_as_sf(gwr.adaptive_aic$SDF) %>%
  st_transform(crs=3414)
```

Below displays the content of *condo_resale.sf.adaptive_aic* sf data frame.

```{r}
glimpse(condo_resale.sf.adaptive_aic)
```

```{r}
summary(gwr.adaptive_aic$SDF$yhat)
```

### 4.4 **Visualising local R2**

The code chunks below is used to create an interactive point symbol map for local R2.

```{r}
tmap_mode("view")
selected_area <- mpsz_svy21[mpsz_svy21$PLN_AREA_N == "HOUGANG", ]

tm_shape(selected_area)+
  tm_polygons(fill_alpha = 0.4, fill = "pink") +
  tm_text("SUBZONE_N", size = 0.9) +
tm_shape(condo_resale.sf.adaptive_aic) +  
  tm_dots(col = "Local_R2",
          border.col = "gray60",
          border.lwd = 1) +
  tm_view(set.zoom.limits = c(11,20))

tmap_mode("plot")

```
::::: markgoals
::: markgoals-header
**Insight**
:::

::: markgoals-container

As we can see from the thematic maps, Almost all data point's local R\^2 is above 0.85, showing GWR predicts well in the Hougang planning zone. This indicates the selected variables *(AREA_SQM + AGE + PROX_CBD + PROX_CHILDCARE + PROX_ELDERLYCARE + PROX_URA_GROWTH_AREA + PROX_MRT + PROX_PARK + PROX_PRIMARY_SCH + PROX_SHOPPING_MALL + PROX_BUS_STOP + NO_Of_UNITS + FAMILY_FRIENDLY + FREEHOLD)* are highly relevant and capture the main drivers of property values well.
:::
:::::

## References

-   Kam, T. S. (2022, December 8). R for geospatial data science and analytics. <https://r4gdsa.netlify.app/>
