---
title: "08 - Geographically Weighted Predictive Modelling"
author: "Cathy C."
date: "Oct 25, 2025"
date-modified: "last-modified" #auto render date
format: html
execute: #how YML page work
  echo: true #will show code
  eval: true #will run code every time and show output
  warning: false #dont show warning msg
  freeze: true #controls rendering - no render if no changes
---

## 1 Overview

Predictive modelling uses statistical learning or machine learning techniques to **predict outcomes**. By and large, the event wants to predict is in the future. However, a set of known outcome and predictors (also known as variables) will be used to calibrate the predictive models.

Geospatial predictive modelling is conceptually rooted in the principle that the occurrences of events being modeled are limited in distribution. When geographically referenced data are used, occurrences of events are neither uniform nor random in distribution over space. There are geospatial factors (infrastructure, sociocultural, topographic, etc.) that constrain and influence where the locations of events occur. **Geospatial predictive modeling attempts to describe those constraints and influences by spatially correlating occurrences of historical geospatial locations with environmental factors that represent those constraints and influences.**

## 2 Getting started

In this in-class exercise, we will build predictive model by using **geographical random forest method**. By the end of this hands-on exercise, we will achieve:

-   preparing training and test data sets by using appropriate data sampling methods,

-   calibrating predictive models by using both geospatial statistical learning and machine learning methods,

-   comparing and selecting the best model for predicting the future outcome,

-   predicting the future outcomes by using the best model calibrated.

## 3 The data

-   **Aspatial dataset**:

    -   HDB Resale data: a list of HDB resale transacted prices in Singapore from Jan 2017 onwards. It is in csv format which can be downloaded from Data.gov.sg.

-   **Geospatial dataset**:

    -   *MP14_SUBZONE_WEB_PL*: a polygon feature data providing information of URA 2014 Master Plan Planning Subzone boundary data. It is in ESRI shapefile format. This data set was also downloaded from Data.gov.sg

-   **Locational factors with geographic coordinates**:

    -   Downloaded from **Data.gov.sg**.

        -   **Eldercare** data is a list of eldercare in Singapore. It is in shapefile format.
        -   **Hawker Centre** data is a list of hawker centres in Singapore. It is in geojson format.
        -   **Parks** data is a list of parks in Singapore. It is in geojson format.
        -   **Supermarket** data is a list of supermarkets in Singapore. It is in geojson format.
        -   **CHAS clinics** data is a list of CHAS clinics in Singapore. It is in geojson format.
        -   **Childcare service** data is a list of childcare services in Singapore. It is in geojson format.
        -   **Kindergartens** data is a list of kindergartens in Singapore. It is in geojson format.

    -   Downloaded from **Datamall.lta.gov.sg**.

        -   **MRT** data is a list of MRT/LRT stations in Singapore with the station names and codes. It is in shapefile format.
        -   **Bus stops** data is a list of bus stops in Singapore. It is in shapefile format.

-   **Locational factors without geographic coordinates**:

    -   Downloaded from **Data.gov.sg**.
        -   **Primary school** data is extracted from the list on General information of schools from data.gov portal. It is in csv format.
    -   Retrieved/Scraped from **other sources**
        -   **CBD** coordinates obtained from Google.
        -   **Shopping malls** data is a list of Shopping malls in Singapore obtained from [Wikipedia](https://en.wikipedia.org/wiki/List_of_shopping_malls_in_Singapore).
        -   **Good primary schools** is a list of primary schools that are ordered in ranking in terms of popularity and this can be found at [Local Salary Forum](https://www.salary.sg/2021/best-primary-schools-2021-by-popularity).

## 4 Installing and Loading R packages

This code chunk performs 3 tasks: - A list called packages will be created and will consists of all the R packages required to accomplish this exercise. - Check if R packages on package have been installed in R and if not, they will be installed. - After all the R packages have been installed, they will be loaded.

```{r}
pacman::p_load(sf, spdep, GWmodel, SpatialML, 
               tmap, rsample, Metrics, tidyverse)
```

| R Package     | Description |
|---------------|-------------|
| **sf**        |             |
| **spdep**     |             |
| **GWmodel**   |             |
| **SpatialML** |             |
| **tmap**      |             |
| **rsample**   |             |
| **Metrics**   |             |
| **tidyvers**e |             |

## 5 Prepare data

### 5.1 Reading data file to rds

Reading the input data sets. It is in simple feature data frame.

```{r}
mdata <- read_rds("data/rawdata/mdata.rds")
```

### 5.2 Data Sampling

The entire data are split into training and test data sets with 65% and 35% respectively by using *initial_split()* of **rsample** package. rsample is one of the packages of **tigymodels**.

```{r}
set.seed(1234)
resale_split <- initial_split(mdata, 
                              prop = 6.5/10,)
train_data <- training(resale_split)
test_data <- testing(resale_split)
```

```{r}
write_rds(train_data, "data/model/train_data.rds")
write_rds(test_data, "data/model/test_data.rds")
```

## 6 Computing Correlation Matrix

Before loading the predictors into a predictive model, it is always a good practice to **use correlation matrix** to examine if there is sign of **multicolinearity**.

```{r}
#| fig-width: 12
#| fig-height: 10

par(bg = "#D0DDD0") 

mdata_nogeo <- mdata %>%
  st_drop_geometry()
corrplot::corrplot(cor(mdata_nogeo[, 2:17]), 
                   diag = FALSE, 
                   order = "AOE",
                   tl.pos = "td", 
                   tl.cex = 0.8, 
                   method = "number", 
                   type = "upper",
                   number.cex = 0.8,
                   bg = "white")
```

::::: markgoals
::: markgoals-header
**Note**
:::

::: markgoals-container
There is no sign of multicolinearity as all the correlation values are \< 0.8.
:::
:::::

## 7 Retriving the Stored Data

```{r}
train_data <- read_rds("data/model/train_data.rds")
test_data <- read_rds("data/model/test_data.rds")
```

## 8 Building a non-spatial multiple linear regression

```{r}
price_mlr <- lm(resale_price ~ floor_area_sqm +
                  storey_order + remaining_lease_mths +
                  PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER +
                  PROX_MRT + PROX_PARK + PROX_MALL + 
                  PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
                  WITHIN_350M_CHILDCARE + WITHIN_350M_BUS +
                  WITHIN_1KM_PRISCH,
                data=train_data)
summary(price_mlr)
```

## 9 gwr predictive method

In this section, we will calibrate a model to **predict HDB resale price** by using **geographically weighted regression method** of [**GWmodel**](https://cran.r-project.org/web/packages/GWmodel/index.html) package.

### 9.1 Computing adaptive bandwidth

Firstly, `bw.gwr()` of **GWmodel** package will be used to determine the optimal bandwidth to be used. The `bw.gwr()` function in R is part of the GWmodel package and is used for automatic bandwidth selection to calibrate a basic Geographically Weighted Regression (GWR) model.

We use below code to determine adaptive bandwidth and CV method is used to determine the optimal bandwidth.

```{r}
#| eval: false
bw_adaptive <- bw.gwr(
  formula = resale_price ~ floor_area_sqm +
    storey_order + remaining_lease_mths +
    PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER +
    PROX_MRT + PROX_PARK + PROX_MALL + 
    PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
    WITHIN_350M_CHILDCARE + WITHIN_350M_BUS +
    WITHIN_1KM_PRISCH,
  data=train_data,
  approach="CV",
  kernel="gaussian",
  adaptive=TRUE,
  longlat=FALSE)

write_rds(bw_adaptive, "data/model/bw_adaptive.rds")
```

::::: markgoals
::: markgoals-header
**Note**
:::

::: markgoals-container
-   The result shows that 40 neighbour points will be the optimal bandwidth to be used if adaptive bandwidth is used for this data set.

-   The `bw.gwr()` function in R is part of the GWmodel package and is used for automatic bandwidth selection to calibrate a basic Geographically Weighted Regression (GWR) model. The algorithm searches for the bandwidth that minimizes the CV score. Lower CV score means better model fit.

-   **Function Purpose**

    -   It determines the optimal bandwidth (either fixed distance or adaptive, representing the number of nearest neighbors) for a GWR model.
    -   This optimal bandwidth minimizes a chosen criterion, typically the Cross-Validation (CV) score or the corrected Akaike Information Criterion (AICc).
:::
:::::

### 9.2 Constructing the adaptive bandwidth gwr model

First, call the saved bandwidth:

```{r}
bw_adaptive <- read_rds("data/model/bw_adaptive.rds")
```

Now, we can go ahead to calibrate the gwr-based hedonic pricing model by using adaptive bandwidth and Gaussian kernel as shown in the code chunk below.

```{r}
#| eval: false
gwr_adaptive <- gwr.basic(
  formula = resale_price ~ floor_area_sqm +
    storey_order + remaining_lease_mths + 
    PROX_CBD + PROX_ELDERLYCARE + 
    PROX_HAWKER + PROX_MRT + PROX_PARK + 
    PROX_MALL + PROX_SUPERMARKET +
    WITHIN_350M_KINDERGARTEN +
    WITHIN_350M_CHILDCARE + 
    WITHIN_350M_BUS + WITHIN_1KM_PRISCH,
  data=train_data,
  bw=bw_adaptive, 
  kernel = 'gaussian', 
  adaptive=TRUE,
  longlat = FALSE)

write_rds(gwr_adaptive, 
          "data/model/gwr_adaptive.rds")
```

### 9.3 Retrieve gwr output object

First we retrieve the saved gwr model object.

```{r}
gwr_adaptive <- read_rds("data/model/gwr_adaptive.rds")
```

Below code displays the model output:

```{r}
gwr_adaptive
```

### 9.4 Computing adaptive bandwidth for the test data

```{r}
#| eval: false
gwr_bw_test_adaptive <- bw.gwr(
  formula = resale_price ~ floor_area_sqm +
    storey_order + remaining_lease_mths +
    PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER +
    PROX_MRT + PROX_PARK + PROX_MALL + 
    PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
    WITHIN_350M_CHILDCARE + WITHIN_350M_BUS +
    WITHIN_1KM_PRISCH,
  data=test_data,
  approach="CV",
  kernel="gaussian",
  adaptive=TRUE,
  longlat=FALSE)

write_rds(gwr_bw_test_adaptive, 
          "data/model/gwr_bw_test_adaptive.rds")
```

::::: markgoals
::: markgoals-header
**Note**
:::

::: markgoals-container
-   The result shows that 25 neighbour points will be the optimal bandwidth to be used if adaptive bandwidth is used for this data set.
:::
:::::

### 9.5 Computing predicted values of the test data

```{r}
#| eval: false
# Get the local coefficients from fitted model
betas <- gwr_model$SDF@data

# Get coefficient column names (exclude non-coefficient columns)
coef_names <- names(betas)[grep("^(Intercept|floor_area|storey|remaining|PROX|WITHIN)", names(betas))]

predictions <- numeric(nrow(test_sp))

for(i in 1:nrow(test_sp)) {
  # Find distances from this test point to all train points
  dists <- spDists(test_sp[i,], train_sp)
  
  # Find k nearest neighbors (k=40 for adaptive bandwidth)
  nearest <- order(dists)[1:40]
  
  # Get weighted average of coefficients from nearest neighbors
  local_betas <- colMeans(betas[nearest, coef_names])
  
  # Create predictor vector matching coefficient order
  X <- c(
    Intercept = 1,
    floor_area_sqm = test_sp@data[i, "floor_area_sqm"],
    storey_order = test_sp@data[i, "storey_order"],
    remaining_lease_mths = test_sp@data[i, "remaining_lease_mths"],
    PROX_CBD = test_sp@data[i, "PROX_CBD"],
    PROX_ELDERLYCARE = test_sp@data[i, "PROX_ELDERLYCARE"],
    PROX_HAWKER = test_sp@data[i, "PROX_HAWKER"],
    PROX_MRT = test_sp@data[i, "PROX_MRT"],
    PROX_PARK = test_sp@data[i, "PROX_PARK"],
    PROX_MALL = test_sp@data[i, "PROX_MALL"],
    PROX_SUPERMARKET = test_sp@data[i, "PROX_SUPERMARKET"],
    WITHIN_350M_KINDERGARTEN = test_sp@data[i, "WITHIN_350M_KINDERGARTEN"],
    WITHIN_350M_CHILDCARE = test_sp@data[i, "WITHIN_350M_CHILDCARE"],
    WITHIN_350M_BUS = test_sp@data[i, "WITHIN_350M_BUS"],
    WITHIN_1KM_PRISCH = test_sp@data[i, "WITHIN_1KM_PRISCH"]
  )
  
  # Ensure same length and order
  predictions[i] <- sum(local_betas[names(X)] * X)
}

gwr_pred <- list(predictions = predictions)
write_rds(gwr_pred, "data/model/gwr_pred.rds")

```

```{r}
args(gwr.predict)
```

## 10 Calibrating Random Forest Model

In this section, we will calibrate a model to predict HDB resale price by using random forest function of [**ranger**](https://cran.r-project.org/web/packages/ranger/index.html) package.

### 10.1 Preparing coordinates data

Below we will extract the x, y coordinates of the full, training and test data sets.

```{r}
#| eval: false
coords <- st_coordinates(mdata)
coords_train <- st_coordinates(train_data)
coords_test <- st_coordinates(test_data)

coords_train <- write_rds(coords_train, "data/model/coords_train.rds" )
coords_test <- write_rds(coords_test, "data/model/coords_test.rds" )
```

```{r}
#| echo: false
coords_train <- read_rds("data/model/coords_train.rds")
coords_test <- read_rds("data/model/coords_test.rds")

```

### 10.2 Dropping geometry field

In this step we will drop the geometry column with `st_drop_geometry` from **sf** package.

```{r}
train_data_nogeom <- train_data %>% 
  st_drop_geometry()

test_data_nogeom <- test_data %>% 
  st_drop_geometry()
```

### 10.3 Calibrating a non-spatial random forest model

```{r}
#| eval: false
set.seed(1234)
rf <- ranger(
  formula = resale_price ~ floor_area_sqm +
    storey_order + remaining_lease_mths + 
    PROX_CBD + PROX_ELDERLYCARE + 
    PROX_HAWKER + PROX_MRT + PROX_PARK + 
    PROX_MALL + PROX_SUPERMARKET +
    WITHIN_350M_KINDERGARTEN + 
    WITHIN_350M_CHILDCARE + 
    WITHIN_350M_BUS +  
    WITHIN_1KM_PRISCH,
  data=train_data_nogeom)
rf

write_rds(rf, "data/model/rf.rds")
```

```{r}
#| echo: false
rf <- read_rds("data/model/rf.rds")
rf
```

## 11 Calibrating Geographical Weighted Random Forest Model

In this section, you will learn how to calibrate a model to predict HDB resale price by using `grf()` of [**SpatialML**](https://cran.r-project.org/web/packages/ranger/index.html) package.

### 11.1

Below code calibrates a geographic ranform forest model by using `grf()` of **SpatialML** package.

```{r}
#| eval: false
set.seed(1234)
gwRF_adaptive <- grf(
  formula = resale_price ~ floor_area_sqm +
    storey_order + remaining_lease_mths + 
    PROX_CBD + PROX_ELDERLYCARE + 
    PROX_HAWKER + PROX_MRT + PROX_PARK + 
    PROX_MALL + PROX_SUPERMARKET +
    WITHIN_350M_KINDERGARTEN +
    WITHIN_350M_CHILDCARE + 
    WITHIN_350M_BUS + WITHIN_1KM_PRISCH,
  dframe=train_data_nogeom, 
  bw=40,
  kernel="adaptive",
  coords=coords_train)

write_rds(gwRF_adaptive, "data/model/gwRF_adaptive.rds")
```

```{r}
#| echo: false
gwRF_adaptive <- read_rds("data/model/gwRF_adaptive.rds")
```

### 11.2 **Predicting by using test data**

#### 11.2.1 Preparing the test data

TO prepare the test data, we first combine the test data with its corresponding coordinates data.

```{r}
test_data_nogeom <- cbind(
  test_data, coords_test) %>%
  st_drop_geometry()
```

#### 11.2.2 Predicting with test data

Next, `predict.grf()` of spatialML package will be used to predict the resale value by using the test data and gwRF_adaptive model calibrated earlier.

```{r}
#| eval: false
gwRF_pred <- predict.grf(gwRF_adaptive, 
                           test_data_nogeom, 
                           x.var.name="X",
                           y.var.name="Y", 
                           local.w=1,
                           global.w=0)

GRF_pred <- write_rds(gwRF_pred, "data/model/GRF_pred.rds")
```

#### 11.2.3 Converting the predicting output into a data frame

The output of the `predict.grf()` is a **vector** of predicted values. It is wiser to **convert it into a data frame** for further visualisation and analysis.

```{r}
GRF_pred <- read_rds("data/model/GRF_pred.rds")
GRF_pred_df <- as.data.frame(GRF_pred)
```

In the code chunk below, `cbind()` is used to append the predicted values onto test_data.

```{r}
test_data_p <- cbind(
  test_data, GRF_pred_df) %>%
  select(resale_price, GRF_pred)

write_rds(test_data_p, "data/model/test_data_p.rds")
```

### 11.3 **Calculating Root Mean Square Error**

The root mean square error (RMSE) allows us to measure how far predicted values are from observed values in a regression analysis. In the code chunk below, `rmse()` of Metrics package is used to compute the RMSE.

```{r}
rmse(test_data_p$resale_price, 
     test_data_p$GRF_pred)
```

### 11.4 **Visualising the predicted values**

Alternatively, scatterplot can be used to visualise the actual resale price and the predicted resale price by using the code chunk below.

```{r}
ggplot(data = test_data_p,
       aes(x = GRF_pred,
           y = resale_price)) +
  geom_point(size = 0.2) +
  theme_bw() +
  theme(panel.background = element_rect(fill = "grey80"),
        plot.background = element_rect(fill = "#D0DDD0"))


    
```

::::: markgoals
::: markgoals-header
**Note**
:::

::: markgoals-container
A better predictive model should have the scatter point close to the diagonal line. The scatter plot can be also used to detect if any outliers in the model.

:::
:::::

## References

-   Kam, T. S. (2022, December 8). R for geospatial data science and analytics. <https://r4gdsa.netlify.app/>

