---
title: "07 - Calibrating Hedonic Pricing Model for Private Highrise Property with GWR Method"
author: "Cathy C."
date: "Oct 17, 2025"
date-modified: "last-modified" #auto render date
format: html
execute: #how YML page work
  echo: true #will show code
  eval: true #will run code every time and show output
  warning: false #dont show warning msg
  freeze: true #controls rendering - no render if no changes
---

## 1 Overview

**Geographically weighted regression (GWR)** is a spatial statistical technique that takes non-stationary variables into consideration (e.g., climate; demographic factors; physical environment characteristics) and models the local relationships between these independent variables and an outcome of interest (also known as dependent variable). In this hands-on exercise, we will build [hedonic pricing](https://www.investopedia.com/terms/h/hedonicpricing.asp) models by using GWR methods. The dependent variable is the resale prices of condominium in 2015. The independent variables are divided into either structural and locational.

::::: markgoals
::: markgoals-header
**Note**
:::

::: markgoals-container
**Non-stationary variables** = Things that change depending on where you are.

The two types of factors:

-   **Structural factors** - Physical features of the condo itself (size, number of rooms, age of building, facilities)
-   **Locational factors** - Where the condo is located (distance to MRT station, nearby schools, shopping centers)

**Hedonic pricing** = identifies price factors based on the idea that a good's price is influenced by both its internal characteristics and external factors. Hedonic pricing is often used to estimate the [value](https://www.investopedia.com/articles/trading/09/quant-strategies.asp) of environmental or ecosystem services that influence home prices. This valuation method often requires statistical expertise and model specification after collecting data.
:::
:::::

## 2 Getting started

We will be using the following R packages:

| R Package | Description |
|---------------------|---------------------------------------------------|
| [**olsrr**](https://olsrr.rsquaredacademy.com/) | R package for building OLS and performing diagnostics tests |
| [**GWmodel**](https://cran.r-project.org/web/packages/GWmodel/index.html) | R package for calibrating geographical weighted family of models |
| [**corrplot**](https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html) | R package for multivariate data visualisation and analysis |
| **sf** | Spatial data handling |
| **tidyverse** | Attribute data handling with **readr**, **ggplot2** and **dplyr** |
| **tmap** | Do Choropleth mapping |

```{r}
pacman::p_load(olsrr, corrplot, ggpubr, sf, sfdep, GWmodel, tmap, tidyverse, gtsummary,
               performance, RColorBrewer)
```

## 3 About GWmodel

[**GWmodel**](https://www.jstatsoft.org/article/view/v063i17) package provides a collection of localised spatial statistical methods, namely: GW summary statistics, GW principal components analysis, GW discriminant analysis and various forms of GW regression; some of which are provided in basic and robust (outlier resistant) forms. Commonly, outputs or parameters of the GWmodel are mapped to provide a useful exploratory tool, which can often precede (and direct) a more traditional or sophisticated statistical analysis.

## 4 The data

### 4.1 Data set

Two data sets will be used in this model building exercise, they are:

-   URA Master Plan subzone boundary in shapefile format (i.e. *MP14_SUBZONE_WEB_PL*)
-   condo_resale_2015 in csv format (i.e. *condo_resale_2015.csv*)

### 4.2 Geospatial data wrangling

#### 4.2.1 Import geospatial data

The geospatial data used in this hands-on exercise is called MP14_SUBZONE_WEB_PL. It is in ESRI shapefile format. The shapefile consists of URA Master Plan 2014’s planning subzone boundaries. Polygon features are used to represent these geographic boundaries. The GIS data is in svy21 projected coordinates systems.

The code chunk below is used to import *MP_SUBZONE_WEB_PL* shapefile by using `st_read()` of **sf** packages.

```{r}
mpsz = st_read(dsn = "data/geospatial", layer = "MP14_SUBZONE_WEB_PL")
```

#### 4.2.2 Update CRS information

The code chunk below updates the newly imported *mpsz* with the **correct ESPG code (i.e. 3414)**

```{r}
mpsz_svy21 <- st_transform(mpsz, 3414)
```

Let's verify the newly transformed *mpsz_svy21*.

```{r}
st_crs(mpsz_svy21)
```

Notice that the EPSG: **is indicated as *3414* now.**

Next, we will reveal the extent of *mpsz_svy21* by using `st_bbox()` of sf package. The print below reports the extent of mpsz_svy21 layer by its l**ower and upper limits**.

```{r}
st_bbox(mpsz_svy21)
```

### 4.3 Aspatial data wrangling

#### 4.3.1 Import Aspatial data

The *condo_resale_2015* is in csv file format. The codes chunk below uses `read_csv()` function of **readr** package to import *condo_resale_2015* into R as a tibble data frame called *condo_resale*.

```{r}
condo_resale = read_csv("data/aspatial/Condo_resale_2015.csv")
```

```{r}
glimpse(condo_resale)
```

**Look at the data set's longitude and latitude.**

```{r}
head(condo_resale$LONGITUDE)
head(condo_resale$LATITUDE)
```

The print above reveals that the values of *LONGITITUDE* and *LATITUDE* fields are in decimal degree. Most probably wgs84 geographic coordinate system is used.

Next, `summary()` of base R is used to display the summary statistics of *cond_resale* tibble data frame.

```{r}
summary(condo_resale)
```

#### 4.3.2 Convert aspatial data frame into a sf object

Currently, the *condo_resale* tibble data frame is aspatial. We will convert it to a **sf** object. The code chunk below converts condo_resale data frame into a simple feature data frame by using `st_as_sf()` of **sf** packages.

```{r}
condo_resale.sf <- st_as_sf(condo_resale,
                            coords = c("LONGITUDE", "LATITUDE"),
                            crs=4326) %>%
  st_transform(crs=3414)
```

::::: markgoals
::: markgoals-header
**Note**
:::

::: markgoals-container
Notice that `st_transform()` of **sf** package is used to convert the coordinates from wgs84 (i.e. crs:4326) to svy21 (i.e. crs=3414).
:::
:::::

Let's list the content of *condo_resale.sf* object. Notice that the output is in point feature data frame.

```{r}
head(condo_resale.sf)
```

## 5 Exploratory data analysis (EDA)

In the section, we will use statistical graphics functions of **ggplot2** package to perform EDA.

### 5.1 EDA using statistical graphics

We can plot the distribution of *SELLING_PRICE* by using appropriate Exploratory Data Analysis (EDA) as shown in the code chunk below.

```{r}
ggplot(data=condo_resale.sf, 
       aes(x=`SELLING_PRICE`)) +
  geom_histogram(bins=20, 
                 color="#CBCBCB", 
                 fill="#B7B89F")
```

The histogram shows a right skewed distribution. This means that more condominium units were transacted at relatively lower prices.

Statistically, the skewed dsitribution can be normalised by using log transformation. The code chunk below is used to derive a new variable called *LOG_SELLING_PRICE* by using a log transformation on the variable *SELLING_PRICE*. It is performed using `mutate()` of **dplyr** package.

**Below histogram using logarithmic graph looks less skewed after the transformation.**

```{r}
condo_resale.sf <- condo_resale.sf %>%
  mutate(`LOG_SELLING_PRICE` = log(SELLING_PRICE))

ggplot(data=condo_resale.sf, 
       aes(x=`LOG_SELLING_PRICE`)) +
  geom_histogram(bins=20, 
                 color="#CBCBCB", 
                 fill="#B7B89F")
```

### 5.2 Multiple histogram plots distributions of variables

In this section, we will draw a small multiple histograms (also known as trellis plot) by using `ggarrange()` of [**ggpubr**](https://cran.r-project.org/web/packages/ggpubr/index.html) package.

Below we create 12 histograms. Then, `ggarrange()` is used to organised these histogram into a 3 columns by 4 rows small multiple plot.

```{r}
#| fig-width: 12
#| fig-height: 10
#| code-fold: true

custom_theme <- theme(
  axis.text.x = element_text(size = 7),
  axis.title.x = element_text(size = 9),
  plot.background = element_rect(fill = "grey")
)

AREA_SQM <- ggplot(data=condo_resale.sf, 
                   aes(x= `AREA_SQM`)) +
  geom_histogram(bins=20, 
                 color="#777C6D", fill="#B7B89F") +
  custom_theme

AGE <- ggplot(data=condo_resale.sf, 
              aes(x= `AGE`)) +
  geom_histogram(bins=20, 
                 color="#777C6D", fill="#B7B89F") +
  custom_theme

PROX_CBD <- ggplot(data=condo_resale.sf, 
                   aes(x= `PROX_CBD`)) +
  geom_histogram(bins=20, 
                 color="#777C6D", fill="#B7B89F") +
  custom_theme

PROX_CHILDCARE <- ggplot(data=condo_resale.sf, 
                         aes(x= `PROX_CHILDCARE`)) + 
  geom_histogram(bins=20, 
                 color="#777C6D", fill="#B7B89F") +
  custom_theme

PROX_ELDERLYCARE <- ggplot(data=condo_resale.sf, 
                           aes(x= `PROX_ELDERLYCARE`)) +
  geom_histogram(bins=20, 
                 color="#777C6D", fill="#B7B89F") +
  custom_theme

PROX_URA_GROWTH_AREA <- ggplot(data=condo_resale.sf, 
                               aes(x= `PROX_URA_GROWTH_AREA`)) +
  geom_histogram(bins=20, 
                 color="#777C6D", fill="#B7B89F") +
  custom_theme

PROX_HAWKER_MARKET <- ggplot(data=condo_resale.sf, 
                             aes(x= `PROX_HAWKER_MARKET`)) +
  geom_histogram(bins=20, 
                 color="#777C6D", fill="#B7B89F") +
  custom_theme

PROX_KINDERGARTEN <- ggplot(data=condo_resale.sf, 
                            aes(x= `PROX_KINDERGARTEN`)) +
  geom_histogram(bins=20, 
                 color="#777C6D", fill="#B7B89F") +
  custom_theme

PROX_MRT <- ggplot(data=condo_resale.sf, 
                   aes(x= `PROX_MRT`)) +
  geom_histogram(bins=20, 
                 color="#777C6D", fill="#B7B89F") +
  custom_theme

PROX_PARK <- ggplot(data=condo_resale.sf, 
                    aes(x= `PROX_PARK`)) +
  geom_histogram(bins=20, 
                 color="#777C6D", fill="#B7B89F") +
  custom_theme

PROX_PRIMARY_SCH <- ggplot(data=condo_resale.sf, 
                           aes(x= `PROX_PRIMARY_SCH`)) +
  geom_histogram(bins=20, 
                 color="#777C6D", fill="#B7B89F") +
  custom_theme

PROX_TOP_PRIMARY_SCH <- ggplot(data=condo_resale.sf, 
                               aes(x= `PROX_TOP_PRIMARY_SCH`)) +
  geom_histogram(bins=20, 
                 color="#777C6D", fill="#B7B89F") +
  custom_theme

ggarrange(AREA_SQM, AGE, PROX_CBD, PROX_CHILDCARE, 
          PROX_ELDERLYCARE, PROX_URA_GROWTH_AREA,
          PROX_HAWKER_MARKET, PROX_KINDERGARTEN, PROX_MRT,
          PROX_PARK, PROX_PRIMARY_SCH, PROX_TOP_PRIMARY_SCH,  
          ncol = 3, nrow = 4) 
```

### 5.3 Draw statistical point map

Finally, we want to reveal the geospatial distribution condominium resale prices in Singapore. The map will be prepared by using **tmap** package.

First, we will turn on the interactive mode of tmap.

```{r}
tmap_mode("view")
```

Then, we will create a interactive point symbol map.

```{r}
tm_shape(mpsz_svy21)+
  tm_polygons() +
  tm_text("PLN_AREA_N", size = 0.5) +
tm_shape(condo_resale.sf) +  
  tm_dots(fill = "SELLING_PRICE",
          fill_alpha = 0.6,
          size = 0.5,
          fill.scale = tm_scale_intervals(
            style="quantile")) +
  tm_view(set_zoom_limits = c(11,14))
```

::::: markgoals
::: markgoals-header
**Note**
:::

::: markgoals-container
-   [`tm_dots()`](https://www.rdocumentation.org/packages/tmap/versions/2.2/topics/tm_symbols) is used instead of `tm_bubbles(), so it's not oversized.`

-   `set_zoom_limits` argument of `tm_view()` sets the minimum and maximum zoom level to 11 and 14 respectively.
:::
:::::

## 6 Hedonic pricing modelling in R

### 6.1 Simple linear regression method

### 6.2 Multiple linear regression method

### 6.3 **Building a hedonic pricing model using multiple linear regression method**

### 6.4 **Revising the model**

### 6.5 **Preparing Publication Quality Table: gtsummary method**

### 6.6 **Regression Diagnostics**

#### 6.6.1 Multicollinearity test

#### 6.6.2 Test for Non-Linearity

#### 6.6.3 Test for Normality Assumption

#### 6.6.4 Test for Spatial Autocorrelation

## **7 Building Hedonic Pricing Models using GWmodel**

### 7.1 **Building Fixed Bandwidth GWR Model**

#### 7.1.1 Computing fixed bandwidth

#### 7.1.2 GWModel method - fixed bandwidth

### 7.2 **Building Adaptive Bandwidth GWR Model**

#### 7.2.1 Computing the adaptive bandwidth

#### 7.2.2 Constructing the adaptive bandwidth gwr model

### 7.3 **Visualising GWR Output**

### 7.4 **Converting SDF into *sf* data.frame**

### 7.5 **Visualising local R2**

### 7.6 **Visualising coefficient estimates**

## References
