---
title: "07 - Calibrating Hedonic Pricing Model for Private Highrise Property with GWR Method"
author: "Cathy C."
date: "Oct 17, 2025"
date-modified: "last-modified" #auto render date
format: html
execute: #how YML page work
  echo: true #will show code
  eval: true #will run code every time and show output
  warning: false #dont show warning msg
  freeze: true #controls rendering - no render if no changes
---

## 1 Overview

**Geographically weighted regression (GWR)** is a spatial statistical technique that takes non-stationary variables into consideration (e.g., climate; demographic factors; physical environment characteristics) and models the local relationships between these independent variables and an outcome of interest (also known as dependent variable). In this hands-on exercise, we will build [hedonic pricing](https://www.investopedia.com/terms/h/hedonicpricing.asp) models by using GWR methods. The dependent variable is the resale prices of condominium in 2015. The independent variables are divided into either structural and locational.

::::: markgoals
::: markgoals-header
**Note**
:::

::: markgoals-container
**Non-stationary variables** = Things that change depending on where you are.

The two types of factors:

-   **Structural factors** - Physical features of the condo itself (size, number of rooms, age of building, facilities)
-   **Locational factors** - Where the condo is located (distance to MRT station, nearby schools, shopping centers)

**Hedonic pricing** = identifies price factors based on the idea that a good's price is influenced by both its internal characteristics and external factors. Hedonic pricing is often used to estimate the [value](https://www.investopedia.com/articles/trading/09/quant-strategies.asp) of environmental or ecosystem services that influence home prices. This valuation method often requires statistical expertise and model specification after collecting data.
:::
:::::

## 2 Getting started

We will be using the following R packages:

| R Package | Description |
|------------------------------------|------------------------------------|
| [**olsrr**](https://olsrr.rsquaredacademy.com/) | R package for building OLS and performing diagnostics tests |
| [**GWmodel**](https://cran.r-project.org/web/packages/GWmodel/index.html) | R package for calibrating geographical weighted family of models |
| [**corrplot**](https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html) | R package for multivariate data visualisation and analysis |
| **sf** | Spatial data handling |
| **tidyverse** | Attribute data handling with **readr**, **ggplot2** and **dplyr** |
| **tmap** | Do Choropleth mapping |
| **gtsummary** | provides an elegant and flexible way to create publication-ready analytical and summary tables. he {gtsummary} package summarizes data sets, regression models. |

```{r}
pacman::p_load(olsrr, corrplot, ggpubr, sf, sfdep, GWmodel, tmap, tidyverse, gtsummary,
               performance, RColorBrewer)
```

## 3 About GWmodel

[**GWmodel**](https://www.jstatsoft.org/article/view/v063i17) package provides a collection of localised spatial statistical methods, namely: GW summary statistics, GW principal components analysis, GW discriminant analysis and various forms of GW regression; some of which are provided in basic and robust (outlier resistant) forms. Commonly, outputs or parameters of the GWmodel are mapped to provide a useful exploratory tool, which can often precede (and direct) a more traditional or sophisticated statistical analysis.

## 4 The data

### 4.1 Data set

Two data sets will be used in this model building exercise, they are:

-   URA Master Plan subzone boundary in shapefile format (i.e. *MP14_SUBZONE_WEB_PL*)
-   condo_resale_2015 in csv format (i.e. *condo_resale_2015.csv*)

### 4.2 Geospatial data wrangling

#### 4.2.1 Import geospatial data

The geospatial data used in this hands-on exercise is called MP14_SUBZONE_WEB_PL. It is in ESRI shapefile format. The shapefile consists of URA Master Plan 2014’s planning subzone boundaries. Polygon features are used to represent these geographic boundaries. The GIS data is in svy21 projected coordinates systems.

The code chunk below is used to import *MP_SUBZONE_WEB_PL* shapefile by using `st_read()` of **sf** packages.

```{r}
mpsz = st_read(dsn = "data/geospatial", layer = "MP14_SUBZONE_WEB_PL")
```

#### 4.2.2 Update CRS information

The code chunk below updates the newly imported *mpsz* with the **correct ESPG code (i.e. 3414)**

```{r}
mpsz_svy21 <- st_transform(mpsz, 3414)
```

Let's verify the newly transformed *mpsz_svy21*.

```{r}
st_crs(mpsz_svy21)
```

Notice that the EPSG: **is indicated as *3414* now.**

Next, we will reveal the extent of *mpsz_svy21* by using `st_bbox()` of sf package. The print below reports the extent of mpsz_svy21 layer by its l**ower and upper limits**.

```{r}
st_bbox(mpsz_svy21)
```

### 4.3 Aspatial data wrangling

#### 4.3.1 Import Aspatial data

The *condo_resale_2015* is in csv file format. The codes chunk below uses `read_csv()` function of **readr** package to import *condo_resale_2015* into R as a tibble data frame called *condo_resale*.

```{r}
condo_resale = read_csv("data/aspatial/Condo_resale_2015.csv")
```

```{r}
glimpse(condo_resale)
```

**Look at the data set's longitude and latitude.**

```{r}
head(condo_resale$LONGITUDE)
head(condo_resale$LATITUDE)
```

The print above reveals that the values of *LONGITITUDE* and *LATITUDE* fields are in decimal degree. Most probably wgs84 geographic coordinate system is used.

Next, `summary()` of base R is used to display the summary statistics of *cond_resale* tibble data frame.

```{r}
summary(condo_resale)
```

#### 4.3.2 Convert aspatial data frame into a sf object

Currently, the *condo_resale* tibble data frame is aspatial. We will convert it to a **sf** object. The code chunk below converts condo_resale data frame into a simple feature data frame by using `st_as_sf()` of **sf** packages.

```{r}
condo_resale.sf <- st_as_sf(condo_resale,
                            coords = c("LONGITUDE", "LATITUDE"),
                            crs=4326) %>%
  st_transform(crs=3414)
```

::::: markgoals
::: markgoals-header
**Note**
:::

::: markgoals-container
Notice that `st_transform()` of **sf** package is used to convert the coordinates from wgs84 (i.e. crs:4326) to svy21 (i.e. crs=3414).
:::
:::::

Let's list the content of *condo_resale.sf* object. Notice that the output is in point feature data frame.

```{r}
head(condo_resale.sf)
```

## 5 Exploratory data analysis (EDA)

In the section, we will use statistical graphics functions of **ggplot2** package to perform EDA.

### 5.1 EDA using statistical graphics

We can plot the distribution of *SELLING_PRICE* by using appropriate Exploratory Data Analysis (EDA) as shown in the code chunk below.

```{r}
ggplot(data=condo_resale.sf, 
       aes(x=`SELLING_PRICE`)) +
  geom_histogram(bins=20, 
                 color="#CBCBCB", 
                 fill="#B7B89F") +
  theme(plot.background = element_rect(fill = "grey"))
```

The histogram shows a right skewed distribution. This means that more condominium units were transacted at relatively lower prices.

Statistically, the skewed dsitribution can be normalised by using log transformation. The code chunk below is used to derive a new variable called *LOG_SELLING_PRICE* by using a log transformation on the variable *SELLING_PRICE*. It is performed using `mutate()` of **dplyr** package.

**Below histogram using logarithmic graph looks less skewed after the transformation.**

```{r}
condo_resale.sf <- condo_resale.sf %>%
  mutate(`LOG_SELLING_PRICE` = log(SELLING_PRICE))

ggplot(data=condo_resale.sf, 
       aes(x=`LOG_SELLING_PRICE`)) +
  geom_histogram(bins=20, 
                 color="#CBCBCB", 
                 fill="#B7B89F") +
  theme(plot.background = element_rect(fill = "grey"))
```

### 5.2 Multiple histogram plots distributions of variables

In this section, we will draw a small multiple histograms (also known as trellis plot) by using `ggarrange()` of [**ggpubr**](https://cran.r-project.org/web/packages/ggpubr/index.html) package.

Below we create 12 histograms. Then, `ggarrange()` is used to organised these histogram into a 3 columns by 4 rows small multiple plot.

```{r}
#| fig-width: 12
#| fig-height: 10
#| code-fold: true

custom_theme <- theme(
  axis.text.x = element_text(size = 7),
  axis.title.x = element_text(size = 9),
  plot.background = element_rect(fill = "grey")
)

AREA_SQM <- ggplot(data=condo_resale.sf, 
                   aes(x= `AREA_SQM`)) +
  geom_histogram(bins=20, 
                 color="#777C6D", fill="#B7B89F") +
  custom_theme

AGE <- ggplot(data=condo_resale.sf, 
              aes(x= `AGE`)) +
  geom_histogram(bins=20, 
                 color="#777C6D", fill="#B7B89F") +
  custom_theme

PROX_CBD <- ggplot(data=condo_resale.sf, 
                   aes(x= `PROX_CBD`)) +
  geom_histogram(bins=20, 
                 color="#777C6D", fill="#B7B89F") +
  custom_theme

PROX_CHILDCARE <- ggplot(data=condo_resale.sf, 
                         aes(x= `PROX_CHILDCARE`)) + 
  geom_histogram(bins=20, 
                 color="#777C6D", fill="#B7B89F") +
  custom_theme

PROX_ELDERLYCARE <- ggplot(data=condo_resale.sf, 
                           aes(x= `PROX_ELDERLYCARE`)) +
  geom_histogram(bins=20, 
                 color="#777C6D", fill="#B7B89F") +
  custom_theme

PROX_URA_GROWTH_AREA <- ggplot(data=condo_resale.sf, 
                               aes(x= `PROX_URA_GROWTH_AREA`)) +
  geom_histogram(bins=20, 
                 color="#777C6D", fill="#B7B89F") +
  custom_theme

PROX_HAWKER_MARKET <- ggplot(data=condo_resale.sf, 
                             aes(x= `PROX_HAWKER_MARKET`)) +
  geom_histogram(bins=20, 
                 color="#777C6D", fill="#B7B89F") +
  custom_theme

PROX_KINDERGARTEN <- ggplot(data=condo_resale.sf, 
                            aes(x= `PROX_KINDERGARTEN`)) +
  geom_histogram(bins=20, 
                 color="#777C6D", fill="#B7B89F") +
  custom_theme

PROX_MRT <- ggplot(data=condo_resale.sf, 
                   aes(x= `PROX_MRT`)) +
  geom_histogram(bins=20, 
                 color="#777C6D", fill="#B7B89F") +
  custom_theme

PROX_PARK <- ggplot(data=condo_resale.sf, 
                    aes(x= `PROX_PARK`)) +
  geom_histogram(bins=20, 
                 color="#777C6D", fill="#B7B89F") +
  custom_theme

PROX_PRIMARY_SCH <- ggplot(data=condo_resale.sf, 
                           aes(x= `PROX_PRIMARY_SCH`)) +
  geom_histogram(bins=20, 
                 color="#777C6D", fill="#B7B89F") +
  custom_theme

PROX_TOP_PRIMARY_SCH <- ggplot(data=condo_resale.sf, 
                               aes(x= `PROX_TOP_PRIMARY_SCH`)) +
  geom_histogram(bins=20, 
                 color="#777C6D", fill="#B7B89F") +
  custom_theme

ggarrange(AREA_SQM, AGE, PROX_CBD, PROX_CHILDCARE, 
          PROX_ELDERLYCARE, PROX_URA_GROWTH_AREA,
          PROX_HAWKER_MARKET, PROX_KINDERGARTEN, PROX_MRT,
          PROX_PARK, PROX_PRIMARY_SCH, PROX_TOP_PRIMARY_SCH,  
          ncol = 3, nrow = 4) 
```

### 5.3 Draw statistical point map

Finally, we want to reveal the geospatial distribution condominium resale prices in Singapore. The map will be prepared by using **tmap** package.

First, we will turn on the interactive mode of tmap.

```{r}
#tmap_mode("view")
```

Then, we will create a interactive point symbol map.

```{r}
tm_shape(mpsz_svy21)+
  tm_polygons() +
  tm_text("SUBZONE_N", size = 0.5) +
tm_shape(condo_resale.sf) +  
  tm_dots(fill = "SELLING_PRICE",
          fill_alpha = 0.6,
          size = 0.5,
          fill.scale = tm_scale_intervals(
            style="quantile")) +
  tm_view(set_zoom_limits = c(11,16))
```

::::: markgoals
::: markgoals-header
**Note**
:::

::: markgoals-container
-   [`tm_dots()`](https://www.rdocumentation.org/packages/tmap/versions/2.2/topics/tm_symbols) is used instead of `tm_bubbles(), so it's not oversized.`

-   `set_zoom_limits` argument of `tm_view()` sets the minimum and maximum zoom level to 11 and 14 respectively.
:::
:::::

**Don't forget to return to the *"plot"* mode.**

```{r}
tmap_mode("plot")
```

## 6 Hedonic pricing modelling in R

In this section, we will build hedonic pricing models for condominium resale units using [`lm()`](https://www.rdocumentation.org/packages/stats/versions/3.5.2/topics/lm) of R base.

### 6.1 Simple linear regression method

First, we will build a simple linear regression model by using *SELLING_PRICE* as the dependent variable and *AREA_SQM* as the independent variable. **\[*AREA_SQM*➡️ *SELLING_PRICE* \]**

`lm()` returns an object of class “lm” or for multiple responses of class c(“mlm”, “lm”).

```{r}
condo.slr <- lm(formula=SELLING_PRICE ~ AREA_SQM, 
                data = condo_resale.sf)
```

The functions `summary()` and `anova()` can be used to obtain and print a summary and analysis of variance table of the results. The generic accessor functions coefficients, effects, fitted.values and residuals extract various useful features of the value returned by `lm`.

```{r}
summary(condo.slr)
```

::::: markgoals
::: markgoals-header
\*\*Learning\*
:::

::: markgoals-container
The output report reveals that the SELLING_PRICE can be explained by using the formula:

```         
      *y = -258121.1 + 14719x1*
```

-   The R-squared of 0.4518 reveals that the simple regression model built is able to explain about 45% of the resale prices.

-   Since p-value is much smaller than 0.0001, we will **reject the null hypothesis** that mean is a good estimator of *SELLING_PRICE*. This will allow us to infer that **simple linear regression model above is a good estimator of *SELLING_PRICE*.**

-   The **Coefficients:** section of the report reveals that the p-values of both the estimates of the Intercept and ARA_SQM are smaller than 0.001. In view of this, the null hypothesis of the B0 and B1 are equal to 0 will be rejected. As a results, we will be able to infer that the B0 and B1 are good parameter estimates.
:::
:::::

To visualise the best fit curve on a scatterplot, we can incorporate `lm()` as a method function in ggplot’s geometry as shown in the code chunk below.

```{r}
#| fig-width: 12
#| fig-height: 8
ggplot(data=condo_resale.sf,  
       aes(x=`AREA_SQM`, y=`SELLING_PRICE`)) +
  geom_point(size = 0.3, color = "#777C6D") +
  geom_smooth(method = lm) +
  theme(plot.background = element_rect(fill = "grey"))
```

::::: markgoals
::: markgoals-header
**Insight**
:::

::: markgoals-container
Figure above reveals that there are a few statistical outliers with relatively high selling prices.
:::
:::::

### 6.2 Multiple linear regression method

#### 6.2.1 Visualising the relationships of the independent variables

Before building a multiple regression model, it is important to ensure that the indepdent variables used are not highly correlated to each other. If these highly correlated independent variables are used in building a regression model by mistake, the quality of the model will be compromised. This phenomenon is known as **multicollinearity** in statistics.

Correlation matrix is commonly used to visualise the relationships between the independent variables. Beside the `pairs()` of R, there are many packages support the display of a correlation matrix. In this section, the [**corrplot**](https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html) package will be used.

Below we will plot a scatterplot matrix of the relationship between the independent variables in *condo_resale* data.frame.

```{r}
#| fig-width: 12
#| fig-height: 8
par(bg = "grey") 
corrplot(cor(condo_resale[, 5:23]), 
         bg = "#E7F0DC",
         diag = FALSE, 
         order = "AOE",
         tl.pos = "td", 
         tl.cex = 0.7, 
         tl.col = "black",
         tl.srt = 45,
         method = "number", 
         number.cex = 0.9,
         type = "upper")
```

::::: markgoals
::: markgoals-header
**Learning**
:::

::: markgoals-container
-   Matrix reorder is very important for mining the hiden structure and patter in the matrix. There are four methods in corrplot (parameter order), named “AOE”, “FPC”, “hclust”, “alphabet”. In the code chunk above, **AOE order** is used. It orders the variables by using the *angular order of the eigenvectors* method suggested by [Michael Friendly](https://www.datavis.ca/papers/corrgram.pdf).

-   From the scatterplot matrix, it is clear that ***Freehold*** is highly correlated to ***LEASE_99YEAR***. In view of this, it is wiser to only include either one of them in the subsequent model building. As a result, ***LEASE_99YEAR*** is excluded in the subsequent model building.
:::
:::::

### 6.3 Building a hedonic pricing model using multiple linear regression method

The code chunk below using `lm()` to calibrate the **multiple linear regression** model.

`lm()` of Base R is used to calibrate a multiple linear regression model. The model output is stored in an lm object called *condo.mlr*. - `summary()` is used to print the model output.

```{r}
condo.mlr <- lm(
  formula = SELLING_PRICE ~ AREA_SQM + AGE  + 
    PROX_CBD + PROX_CHILDCARE + PROX_ELDERLYCARE +
    PROX_URA_GROWTH_AREA + PROX_HAWKER_MARKET   +
    PROX_KINDERGARTEN   + PROX_MRT  + PROX_PARK +
    PROX_PRIMARY_SCH + PROX_TOP_PRIMARY_SCH +
    PROX_SHOPPING_MALL  + PROX_SUPERMARKET + 
    PROX_BUS_STOP   + NO_Of_UNITS + 
    FAMILY_FRIENDLY + FREEHOLD, 
  data=condo_resale.sf)
summary(condo.mlr)
```

### 6.4 Revising the model

With reference to the report above, it is clear that not all the independent variables are statistically significant. We will revised the model by **removing those variables which are not statistically significant, including *PROX_HAWKER_MARKET*, *PROX_KINDERGARTEN*, *PROX_TOP_PRIMARY_SCH*, and *PROX_SUPERMARKET*.**

Now, we are ready to calibrate the revised model.

```{r}
condo.mlr1 <- lm(
  formula = SELLING_PRICE ~ AREA_SQM + AGE + 
    PROX_CBD + PROX_CHILDCARE + PROX_MRT +
    PROX_ELDERLYCARE  + PROX_URA_GROWTH_AREA +
    PROX_PARK + PROX_PRIMARY_SCH + 
    PROX_SHOPPING_MALL+ PROX_BUS_STOP + 
    NO_Of_UNITS + FAMILY_FRIENDLY + FREEHOLD,
  data=condo_resale.sf)
summary(condo.mlr1)
```

::::: markgoals
::: markgoals-header
**Insight**
:::

::: markgoals-container
The output above shows that all explanatory variables are now statistically significant at 95% confident level.
:::
:::::

### 6.5 Preparing Publication Quality Table: gtsummary method

The [**gtsummary**](https://www.danieldsjoberg.com/gtsummary/index.html) package provides an elegant and flexible way to create publication-ready summary tables in R. In the code chunk below, [`tbl_regression()`](https://www.danieldsjoberg.com/gtsummary/reference/tbl_regression.html) is used to create a well formatted regression report.

``` r
# Install broom.helpers and its dependencies
install.packages("broom.helpers", dependencies = TRUE)

# Or install both packages together
install.packages(c("gtsummary", "broom.helpers"), dependencies = TRUE)
```

```{r}
#remove.packages("gtsummary")
#install.packages("gtsummary")
library(gtsummary)
```

```{r}
tbl_regression(condo.mlr1, intercept = TRUE)
```

With gtsummary package, model statistics can be included in the report by either appending them to the report table by using [`add_glance_table()`](https://www.danieldsjoberg.com/gtsummary/reference/add_glance.html) or adding as a table source note by using [`add_glance_source_note()`](https://www.danieldsjoberg.com/gtsummary/reference/add_glance.html) as shown in the code chunk below.

::: panel-tabset
## add_glance_source_note()

Use [`add_glance_source_note()`](https://www.danieldsjoberg.com/gtsummary/reference/add_glance.html)

```{r}
tbl_regression(condo.mlr1, 
               intercept = TRUE) %>% 
  add_glance_source_note(
    label = list(sigma ~ "\U03C3"),
    include = c(r.squared, adj.r.squared, 
                AIC, statistic,
                p.value, sigma))
```

## add_glance_table()

Use [`add_glance_table()`](https://www.danieldsjoberg.com/gtsummary/reference/add_glance.html)

```{r}
tbl_regression(condo.mlr1, 
               intercept = TRUE) %>% 
  add_glance_table(
    label = list(sigma ~ "\U03C3"),
    include = c(r.squared, adj.r.squared, 
                AIC, statistic,
                p.value, sigma))
```
:::

### 6.6 Regression Diagnostics

Regression diagnostics are a set of procedures used to check if a regression model’s assumptions are met and how well the model fits the data. These diagnostics involve checking for issues like non-linear relationships, non-normal errors, non-constant variance, and influential observations to ensure the model’s conclusions are valid and reliable. Common methods include graphical analysis, like residual plots and QQ-plots, and quantitative tests

In this section, we would like to introduce you a fantastic R package specially programmed for performing OLS regression diagnostics. It is called [**olsrr**](https://olsrr.rsquaredacademy.com/). It provides a collection of very useful methods for building better multiple linear regression models:

-   comprehensive regression output
-   residual diagnostics
-   measures of influence
-   heteroskedasticity tests
-   collinearity diagnostics
-   model fit assessment
-   variable contribution assessment
-   variable selection procedures

#### 6.6.1 Multicollinearity test

**Multicollinearity** occurs when independent variables are **not** truly independent, meaning a change in one is associated with a change in another. This makes it hard for the model to isolate each variable’s influence on the outcome.

Performing a multicollinearity test is crucial in multiple linear regression because it ensures the reliability and interpretability of the model’s results. High multicollinearity, where independent variables are highly correlated, inflates the variance of the estimated coefficients, making them unstable, unreliable, and difficult to interpret. This instability can lead to misleading statistical conclusions, such as a variable appearing statistically insignificant when it is not.

[`check_collinearity()`](https://easystats.github.io/performance/reference/check_collinearity.html) of **performance** package is used to test if there are sign of multicollinearity.

```{r}
check_collinearity(condo.mlr1)
```

```{r}
mlr.vif <- check_collinearity(condo.mlr1)
plot(mlr.vif)
```

::::: markgoals
::: markgoals-header
**Insight**
:::

::: markgoals-container
Since the VIF of the independent variables are less than 10. We can safely conclude that **there are no sign of multicollinearity** among the independent variables.
:::
:::::

#### 6.6.2 Test for Non-Linearity

In multiple linear regression, it is important for us to test the assumption that linearity and additivity of the relationship between dependent and independent variables.

::: panel-tabset
## ols_plot_resid_fit

[`ols_plot_resid_fit()`](https://olsrr.rsquaredacademy.com/reference/ols_plot_resid_fit.html) of **olsrr** package is used to perform linearity assumption test.

The figures below reveals that most of the data poitns are scattered around the 0 line, hence we can safely conclude that the relationships between the dependent variable and independent variables are linear.

```{r}
ols_plot_resid_fit(condo.mlr1)
```

## check_model

.

```{r}
check_model(condo.mlr1, check = "linearity")

```
:::

#### 6.6.3 Test for Normality Assumption

The normality assumption test for multiple linear regression checks **if the model’s residuals (the differences between observed and predicted values) are normally distributed**. This is crucial for accurate hypothesis testing and confidence intervals. To test this, you can use visual methods like histograms and Q-Q plots of the residuals, or conduct statistical tests like Shapiro-Wilk test and Kolmogorov-Smirnov test.

In the code chunk below, `check_normality()` of performance package is used to perform normality assumption test on condo.mlr1 model.

```{r}
check_normality(condo.mlr1)
```

::::: markgoals
::: markgoals-header
**Insight**
:::

::: markgoals-container
The print above reveals that the p-value of the normality assumption test is less than alpha value of 0.05. Hence we reject the normality assumption at 95% confident level.
:::
:::::

-   `check_normality()` calls *stats::shapiro.test* and checks the standardized residuals (or studentized residuals for mixed models) for normal distribution.

-   Note that this formal test almost always yields significant results for the distribution of residuals and visual inspection (e.g. Q-Q plots) are preferable.

Instead of showing the test statistic, `plot()` of **see** package can be used to plot a the output of `check_normality()` for visual inspection as shown below.

::: panel-tabset
## check_normality

Q-Q plot below shows that majority of the data points are felt along the zero line.

```{r}
plot(check_normality(condo.mlr1), 
     type = "qq")
```

## check_model

Another way to check for normality assumption visual is by using [`check_model()`](https://easystats.github.io/performance/reference/check_model.html) of performance package. The figure reveals that the residual of the multiple linear regression model (i.e. condo.mlr1) is resemble normal distribution.

```{r}
check_model(condo.mlr1,
            check= "normality")
```
:::

#### 6.6.4 Test for Spatial Autocorrelation

The hedonic model we try to build are using geographically referenced data. Hence, it is crucial to check for spatial autocorrelation because its presence can produce unreliable and misleading results. Traditional regression models, such as ordinary least squares (OLS), assume that observations are independent of one another. However, spatial data often violates this assumption.

Spatial autocorrelation is the correlation of a variable with itself across different spatial locations. Positive spatial autocorrelation means nearby features tend to be more similar, while negative autocorrelation means they tend to be more dissimilar. This phenomenon is based on the first law of geography: “Everything is related to everything else, but nearby things are more related than distant things”.

Ignoring spatial autocorrelation in a regression model can lead to serious statistical issues:

-   Biased and inefficient coefficient estimates: If autocorrelation is present, standard errors of the model coefficients can be wrong, leading to unreliable hypothesis tests (p-values). The model might appear more significant than it is.

-   Misleading significance tests: Standard regression models cannot distinguish between true explanatory power and the influence of spatial patterns, resulting in inaccurate p-values.

-   Model misspecification: Significant spatial autocorrelation in the regression residuals often signals that important explanatory variables are missing from the model. The spatial patterning of the residuals (over- and under-predictions) can provide clues about what these missing variables might be.

-   Inflated Type I error rates: Researchers might incorrectly reject a true null.

To test for spatial autocorrelation, We can run a Moran’s I test on the model’s residuals. Significant spatial autocorrelation in the residuals means the model is not capturing the full spatial story.

In order to perform spatial autocorrelation test, we need to export the residual of the hedonic pricing model and save it as a data frame first.

```{r}
mlr.output <- as.data.frame(condo.mlr1$residuals)
```

Next, we will join the newly created data frame with *condo_resale.sf* object.

```{r}
condo_resale.sf <- cbind(condo_resale.sf, 
                        condo.mlr1$residuals) %>%
rename(`MLR_RES` = `condo.mlr1.residuals`)
```

Next, we will use **tmap** package to display the distribution of the residuals on an interactive map.

```{r}
#tmap_mode("view")

tm_shape(mpsz_svy21) +
  tm_polygons(fill_alpha = 0.4) +
  tm_text("SUBZONE_N", size = 0.5) +
tm_shape(condo_resale.sf) +  
  tm_dots(
    fill = "MLR_RES",
    size = 0.7,
    col = "black",
    fill.scale = tm_scale(
      n = 10,
      values = rev(
        brewer.pal(11, "RdBu")),
      style = "quantile",
      midpoint = NA),
    fill.legend = tm_legend(title = "Residuals")
  ) + 
  tm_title("LM Residuals (Quantile Classification)") +
  tm_layout(legend.outside = TRUE) +
  tm_view(set_zoom_limits = c(11,16))

tmap_mode("plot")
```

The figure above reveal that there is sign of spatial autocorrelation.

To proof that our observation is indeed true, Global Moran’s I test will be performed

First, we will compute the distance-based weight matrix by using [`st_dist_band()`](https://sfdep.josiahparry.com/reference/st_dist_band) function of **sfdep**.

```{r}
condo_resale.sf <- condo_resale.sf %>% 
  mutate(
    nb = st_dist_band(
      st_geometry(geometry), 
                 upper = 1500),
    wt = st_weights(
      nb, style = "W"),
    .before = 1) 
```

Next, [`global_moran_perm()`](https://sfdep.josiahparry.com/reference/global_moran_perm) of **sfdep** package will be used to perform Moran’s I test for residual spatial autocorrelation

```{r}
set.seed(1234)
global_moran_perm(
  condo_resale.sf$MLR_RES, 
  nb = condo_resale.sf$nb, 
  wt = condo_resale.sf$wt,
  alternative = "two.sided", 
  nsim = 499)
```

::::: markgoals
::: markgoals-header
**Insight**
:::

::: markgoals-container
-   The Global Moran’s I test for residual spatial autocorrelation shows that it’s p-value is less than 0.00000000000000022 which is less than the alpha value of 0.05. Hence, we will reject the null hypothesis that the residuals are randomly distributed.

-   Since the Observed Global Moran I = 0.14389 which is greater than 0, we can infer that the residuals resemble **cluster distribution.**
:::
:::::

## 7 Building Hedonic Pricing Models using GWmodel

In this section, we will model hedonic pricing using both the fixed and adaptive bandwidth schemes.

### 7.1 Building Fixed Bandwidth GWR Model

#### 7.1.1 Computing fixed bandwidth

In the code chunk below `bw.gwr()` of GWModel package is used to determine the **optimal fixed bandwidt**h to use in the model. Notice that the argument ***adaptive*** is set to **FALSE** indicates that we are interested to compute the fixed bandwidth.

There are two possible approaches can be uused to determine the stopping rule, they are: **CV cross-validation approach** and **AIC corrected (AICc) approach**. We define the stopping rule using ***approach*** argeement.

```{r}
bw.fixed <- bw.gwr(
  formula = SELLING_PRICE ~ AREA_SQM + AGE  + 
    PROX_CBD + PROX_CHILDCARE + PROX_ELDERLYCARE    +
    PROX_URA_GROWTH_AREA + PROX_MRT + PROX_PARK + 
    PROX_PRIMARY_SCH + PROX_SHOPPING_MALL   + 
    PROX_BUS_STOP   + NO_Of_UNITS + FAMILY_FRIENDLY + 
    FREEHOLD, 
  data=condo_resale.sf, 
  approach="CV", 
  kernel="gaussian", 
  adaptive=FALSE, 
  longlat=FALSE)
```

::::: markgoals
::: markgoals-header
**Insight**
:::

::: markgoals-container
-   The result shows that the recommended bandwidth is 971.3405 metres.

-   Because of longlat=FALSE parameter, it indicates our data is in a **projected coordinate system** rather than geographic coordinates. Our sf is with SVY21 projection, and SVY21 uses meters as the base unit, so the unit is in meters.
:::
:::::

#### 7.1.2 GWModel method - fixed bandwidth

Now we can use the code chunk below to calibrate the gwr model using fixed bandwidth and gaussian kernel.

```{r}
gwr.fixed <- gwr.basic(
  formula = SELLING_PRICE ~ AREA_SQM + AGE  + 
    PROX_CBD + PROX_CHILDCARE + PROX_ELDERLYCARE    +
    PROX_URA_GROWTH_AREA + PROX_MRT + PROX_PARK +
    PROX_PRIMARY_SCH + PROX_SHOPPING_MALL   + 
    PROX_BUS_STOP   + NO_Of_UNITS + FAMILY_FRIENDLY + 
    FREEHOLD, 
  data=condo_resale.sf, 
  bw=bw.fixed, 
  kernel = 'gaussian', 
  longlat = FALSE)
```

The output is saved in a list of class “gwrm”. The code below can be used to display the model output.

```{r}
gwr.fixed
```

::::: markgoals
::: markgoals-header
**Insight**
:::

::: markgoals-container
The report shows that the AICc of the gwr is 42263.61 which is significantly smaller than the globel multiple linear regression model of 42967.1.
:::
:::::

### 7.2 **Building Adaptive Bandwidth GWR Model**

In this section, we will calibrate the gwr-based hedonic pricing model by using adaptive bandwidth approach.

#### 7.2.1 Computing the adaptive bandwidth

Similar to the earlier section, we will first use `bw.gwr()` to determine the recommended data point to use.

The code chunk used look very similar to the one used to compute the fixed bandwidth except the `adaptive` argument has changed to **TRUE**.

```{r}
bw.adaptive <- bw.gwr(
  formula = SELLING_PRICE ~ AREA_SQM + AGE  + 
    PROX_CBD + PROX_CHILDCARE + PROX_ELDERLYCARE    +
    PROX_URA_GROWTH_AREA + PROX_MRT + PROX_PARK +
    PROX_PRIMARY_SCH + PROX_SHOPPING_MALL   + 
    PROX_BUS_STOP   + NO_Of_UNITS + 
    FAMILY_FRIENDLY + FREEHOLD, 
  data=condo_resale.sf, 
  approach="CV", 
  kernel="gaussian", 
  adaptive=TRUE, 
  longlat=FALSE)
```

::::: markgoals
::: markgoals-header
**Insight**
:::

::: markgoals-container
The result shows that the 30 is the recommended data points to be used.
:::
:::::

#### 7.2.2 Constructing the adaptive bandwidth gwr model

Now, we can go ahead to calibrate the gwr-based hedonic pricing model by using adaptive bandwidth and gaussian kernel as shown in the code chunk below.

```{r}
gwr.adaptive <- gwr.basic(
  formula = SELLING_PRICE ~ AREA_SQM + AGE + 
    PROX_CBD + PROX_CHILDCARE + PROX_ELDERLYCARE +
    PROX_URA_GROWTH_AREA + PROX_MRT + PROX_PARK +
    PROX_PRIMARY_SCH + PROX_SHOPPING_MALL + PROX_BUS_STOP +
    NO_Of_UNITS + FAMILY_FRIENDLY + FREEHOLD,
  data=condo_resale.sf, 
  bw=bw.adaptive, 
  kernel = 'gaussian', 
  adaptive=TRUE, 
  longlat = FALSE)
```

Let's display the model output.

```{r}
gwr.adaptive
```

::::: markgoals
::: markgoals-header
**Insight**
:::

::: markgoals-container
The report shows that the AICc the adaptive distance gwr is 41982.22 which is even **smaller** than the AICc of the fixed distance gwr of 42263.61.
:::
:::::

### 7.3 **Visualising GWR Output**

In addition to regression residuals, the output feature class table includes fields for observed and predicted y values, condition number (cond), Local R2, residuals, and explanatory variable coefficients and standard errors:

-   **Condition Number:** this diagnostic evaluates local collinearity. In the presence of strong local collinearity, results become unstable. Results associated with condition numbers larger than 30, may be unreliable.

-   **Local R2:** these values range between 0.0 and 1.0 and indicate how well the local regression model fits observed y values. Very low values indicate the local model is performing poorly. Mapping the Local R2 values to see where GWR predicts well and where it predicts poorly may provide clues about important variables that may be missing from the regression model.

-   **Predicted:** these are the estimated (or fitted) y values 3. computed by GWR.

-   **Residuals:** to obtain the residual values, the fitted y values are subtracted from the observed y values. Standardized residuals have a mean of zero and a standard deviation of 1. A cold-to-hot rendered map of standardized residuals can be produce by using these values.

-   **Coefficient Standard Error**: these values measure the reliability of each coefficient estimate. Confidence in those estimates are higher when standard errors are small in relation to the actual coefficient values. Large standard errors may indicate problems with local collinearity.

They are all stored in a SpatialPointsDataFrame or SpatialPolygonsDataFrame object integrated with fit.points, GWR coefficient estimates, y value, predicted values, coefficient standard errors and t-values in its “data” slot in an object called **SDF** of the output list.

### 7.4 **Converting SDF into *sf* data.frame**

To visualise the fields in **SDF**, we need to first covert it into **sf** data.frame by using the code chunk below.

```{r}
condo_resale.sf.adaptive <-
  st_as_sf(gwr.adaptive$SDF) %>%
  st_transform(crs=3414)
```

Below displays the content of *condo_resale.sf.adaptive* sf data frame.

```{r}
glimpse(condo_resale.sf.adaptive)
```

```{r}
summary(gwr.adaptive$SDF$yhat)
```

### 7.5 **Visualising local R2**

The code chunks below is used to create an interactive point symbol map for local R2.

```{r}
#tmap_mode("view")
tm_shape(mpsz_svy21)+
  tm_polygons(fill_alpha = 0.1) +
  tm_text("SUBZONE_N", size = 0.5) +
tm_shape(condo_resale.sf.adaptive) +  
  tm_dots(col = "Local_R2",
          border.col = "gray60",
          border.lwd = 1) +
  tm_view(set.zoom.limits = c(11,14))

tmap_mode("plot")
```

### 7.6 **Visualising coefficient estimates**

```{r}
#| fig-width: 12

#tmap_mode("view")
AREA_SQM_SE <- tm_shape(mpsz_svy21)+
  tm_polygons(alpha = 0.1) +
  tm_text("SUBZONE_N", size = 0.5) +
tm_shape(condo_resale.sf.adaptive) +  
  tm_dots(col = "AREA_SQM_SE",
          border.col = "gray60",
          border.lwd = 1) +
  tm_view(set.zoom.limits = c(11,14))

AREA_SQM_TV <- tm_shape(mpsz_svy21)+
  tm_polygons(alpha = 0.1) +
tm_shape(condo_resale.sf.adaptive) +  
  tm_dots(col = "AREA_SQM_TV",
          border.col = "gray60",
          border.lwd = 1) +
  tm_view(set.zoom.limits = c(11,14))

tmap_arrange(AREA_SQM_SE, AREA_SQM_TV, 
             asp=1, ncol=2,
             sync = TRUE)

tmap_mode("plot")
```

## References

-   Kam, T. S. (2022, December 8). R for geospatial data science and analytics. <https://r4gdsa.netlify.app/>
