---
title: "08a - Kick starter - Geographically Weighted Predictive Modelling"
author: "Cathy C."
date: "Oct 25, 2025"
date-modified: "last-modified" #auto render date
format: html
execute: #how YML page work
  echo: true #will show code
  eval: true #will run code every time and show output
  warning: false #dont show warning msg
  freeze: true #controls rendering - no render if no changes
---

## 1 Overview

This exercise is to:

-   prepare and geocode HDB resale price data for geospatial modelling; and
-   perform proximity analysis to count the number of geographic entities located within a defined distance from each HDB property.

## 2 Installing and Loading R packages

One new package called [jsonlite]{style="color: #F08787;"} will be installed and loaded onto R environment. It will be used to parse json file into R as data frame.

```{r}
pacman::p_load(httr, tidyverse, sf, tmap, jsonlite, progress)
```

### Additional packages for predive modelling

```{r}
pacman::p_load(spdep, GWmodel, SpatialML, rsample, Metrics, knitr,
               kableExtra, spatialRF, randomForestExplainer)
```

progress - to monitor and track the computation.

## 3 Geocoding HDB resale data

We use OneMap Geocode API:

-   https://www.onemap.gov.sg/apidocs/.

-   ttps://www.onemap.gov.sg/apidocs/reverseGeocode

One must download *Resale flat prices based on registration date from Jan-2017 onwards* from [Singapore’s open data portal](https://isss626-ay2025-26aug.netlify.app/in-class_ex/in-class_ex08/data.gov.sg).

### 3.1 Selecting data of interest

This data set comprises of HDB resale transactions from Jan 2017 onwards. For the purpose of this in-class exercise, the code chunk below will be used to extract a sub-set of transaction records on 4-Room flat transacted on the month September 2025.

```{r}
#| eval: false
HDBresale <- read_csv(
  "data/rawdata/HDBresale.csv") %>%
  filter(flat_type == "4 ROOM") %>%
  filter(month >= "2025-09" & month < "2025-10") # filter the sample data for the exercise.

write_rds(HDBresale, "data/rds/HDBresale.rds")
```

```{r}
#| echo: false
HDBresale <- read_rds("data/rds/HDBresale.rds")
glimpse(HDBresale)
```

We do further processing for the data as the street names with Saint are usually written as St.

```{r}
# street names is ST, but it's actually Sait.
HDBresale$street_name <- gsub(
  "ST\\.", "SAINT",
  HDBresale$street_name
)
```

Now we will prepare a geocoding function. It aims to perform the following tasks:

-   Combining the block and street name into an address. The addresses will be used to geocode instead of postal code that you learned in In-class Exercise 3b,
-   Passing the address as the **searchVal** in our query,
-   Sending the query to **OneMapSG search**. Note: Since we don’t need all the address details, we can set getAddrDetails as ‘N’,
-   Converting response (JSON object) to text,
-   Saving response in text form as a dataframe, and
-   Retain the latitude and longitude for our output.

```{r}
# for multiple queries
geocode <- function(block, streetname) {
  base_url <- "https://onemap.gov.sg/api/common/elastic/search"
  address <- paste(block, streetname, sep = " ")
  query <- list("searchVal" = address,
                "returnGeom" = "Y",
                "getAddrDetails" = "N", #No. can change to Y for comparison
                "pageNum" = "1") 
  res <- GET(base_url, query = query) #get URL query
  restext <- content(res, as = "text")
  
  output <- fromJSON(restext) %>%
    as.data.frame %>%
    select(results.LATITUDE, results.LONGITUDE)
  
  return(output)
}
```

Need below to run the functions

```{r}
HDBresale$LATITUDE <- 0
HDBresale$LONGITUDE <- 0

for(i in 1:nrow(HDBresale)){
  temp_output <- geocode(HDBresale[i, 4], HDBresale[i, 5])
  
  HDBresale$LATITUDE[i] <- temp_output$results.LATITUDE
  HDBresale$LONGITUDE[i] <- temp_output$results.LONGITUDE
}
```

## 4 Proximity analysis

We will learn how to count the number of geographic entities located within a defined distance from each HDB property.

For the purpose of this exercise, we are interested to count the number of preschools located with 350m of each resale HDB unit.

### 4.1 Transform to sf data frame

Before performing proximity analysis, it is important to ensure that:

-   both input data sets must be in sf objects, and
-   they must be in similar projected coordinates systems.

```{r}
st_crs(HDBresale)
HDBresale_sf <- HDBresale %>%
  st_as_sf(
    coords = c("LONGITUDE", "LATITUDE"),
    crs = 4326) %>% #if decimal degree
  st_transform(crs = 3414)

```

::::: markgoals
::: markgoals-header
**Note**
:::

::: markgoals-container
-   `st_as_sf()` is used to convert HDBresale tibble data.frame to sf object by using values from LONGITUDE and LATITUDE columns to form the geometry features.
-   `st_transform()` is then used to transform the sf object into svy21 projected coordinates system.
-   the output sf object is called\* HDBresale_sf\*. It is in sf [data.frame format]{style="color: #F08787;"}.
:::
:::::

### 4.2 Load preschool data

```{r}
preschool = st_read("data/rawdata/PreSchoolsLocation.kml") %>%
  st_transform(crs = 3414)
```

From this we will have the geometry data for the preschools.

### 4.3 Counting number of preschools

Next we will derive a binary data to find out which schools are within 350 meters.

```{r}
within_350m <- st_is_within_distance(
  HDBresale_sf, preschool, dist = 350)

HDBresale_sf <- HDBresale_sf %>%
  mutate(prischool_350m = map_int(within_350m, length)
)
```

::::: markgoals
::: markgoals-header
**Note**
:::

::: markgoals-container
-   `st_is_within_distance()` of [**sf**]{style="color: #F08787;"} package is used to determines whether two spatial objects (HDB property and preschool) are within 350m distance of each other. It returns a logical matrix (or a sparse list of indices if sparse = TRUE) indicating for each geometry in HDB property which geometries in preschool are within 350m distance.
-   `map_int()` of [**purr**]{style="color: #F08787;"} package is used to the number of preschools found in the newly created WITHIN_350M_PRISCHOOL field.
:::
:::::

### 4.4 Computing shortest distance

Instead of counting number of preschools located within 350m of each HDB resale property, code chunk below is used to compute the distance between the nearest preschool from each HDB resale property.

```{r}
HDBresale_sf <- HDBresale_sf %>%
  mutate(
    PROX_PRESCHOOL = as.numeric(
      st_distance(geometry,
                  preschool[st_nearest_feature(
                    geometry, preschool), ], 
                  by_element = TRUE)
      )
    )
```

# 5 Predictive modeling

## 5.1 Data set

### 5.1.1 Data import

```{r}
mdata <- read_rds("data/rawdata/mdata.rds")
```

### 5.1.2 Data sampling

Calibrating predictive models are computational intensive, especially random forest method is used. For quick prototyping, a 10% sample will be selected at random from the data by using the code chunk below:

```{r}
set.seed(1234)
HDB_sample <- mdata %>%
  sample_n(1500)
```

## 5.2 Checking of overlapping point

\*\*\* When using GWmodel to calibrate explanaroty or predictive models, it is very important to ensure that there are no overlapping point features \*\*\*

Below is to check if there are overlapping point feature - to do spatial jitter.

```{r}
overlapping_points <- HDB_sample %>%
  mutate(overlap = lengths(st_equals(.,.)) > 1)

```

Below we can have a look at overlapping_points, and see the new column overlap showing whether the location has overlapping points. We then calculate how many has overlapping points. In this sample data, we observed 453 overlapping points.

```{r}
head(overlapping_points)
sum(overlapping_points$overlap == "TRUE")
```

## 5.3 Spatial jitter

st_jitter is used to move the point feature by 5m to avoid overlapping point features.

```{r}
HDB_sample <- HDB_sample %>%
  st_jitter(amount = 1)
```

```{r}
set.seed(1234)
resale_split <- initial_split(HDB_sample,
                              prop = 6.67/10,)
train_data <- training(resale_split)
test_data <- testing(resale_split)
```

## 5.4 Constructing the adaptive bandwidth gwr model

```{r}
#| eval: false
bw_adaptive <- bw.gwr(
  formula = resale_price ~ floor_area_sqm +
    storey_order + remaining_lease_mths +
    PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER +
    PROX_MRT + PROX_PARK + PROX_MALL + 
    PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
    WITHIN_350M_CHILDCARE + WITHIN_350M_BUS +
    WITHIN_1KM_PRISCH,
  data=train_data,
  approach="CV",
  kernel="gaussian",
  adaptive=TRUE,
  longlat=FALSE)

write_rds(bw_adaptive, "data/model/bw_adaptive.rds")
```

## 5.5 Constructing the adaptive bandwidth gwr model

```{r}
#| echo: false
bw_adaptive <- read_rds("data/model/bw_adaptive.rds")
```

```{r}
#| eval: false
gwr_adaptive <- gwr.basic(
  formula = resale_price ~ floor_area_sqm +
    storey_order + remaining_lease_mths + 
    PROX_CBD + PROX_ELDERLYCARE + 
    PROX_HAWKER + PROX_MRT + PROX_PARK + 
    PROX_MALL + PROX_SUPERMARKET +
    WITHIN_350M_KINDERGARTEN +
    WITHIN_350M_CHILDCARE + 
    WITHIN_350M_BUS + WITHIN_1KM_PRISCH,
  data=train_data,
  bw=bw_adaptive, 
  kernel = 'gaussian', 
  adaptive=TRUE,
  longlat = FALSE)

write_rds(gwr_adaptive, 
          "data/model/gwr_adaptive.rds")
```

## 5.6 GWR Output

```{r}
#| echo: false
gwr_adaptive <- read_rds("data/model/gwr_adaptive.rds")
gwr_adaptive
```

-   Based on the result of the modelling above, p-value is much smaller than 0.05. We see the adjusted R\^2 = 0.7387, meaning that this model can explain 73% of the data using the variables.

-   The AICc = 24901.56 for Global Regression model is much larger than Geographically Weighted Regression's AICc = 24103.97. The difference of approximately 797.6 points is quite significant. This suggests the **GWR model provides meaningfully better predictions and fit**.

Explanatory variables:

Plot out the significant variables to show distribution and explain which ones have higher/lower influence over the model.

## 5.7 Compute Collinearity Matrix

```{r}
#| fig-width: 12
#| fig-height: 10
par(bg = "#D0DDD0") 

mdata_nogeo <- mdata %>%
  st_drop_geometry()
corrplot::corrplot(cor(mdata_nogeo[, 2:17]), 
                   diag = FALSE, 
                   order = "AOE",
                   tl.pos = "td", 
                   tl.cex = 0.8, 
                   method = "number", 
                   type = "upper",
                   number.cex = 0.8,
                   bg = "white")
```

# 6 Predictive Modeling using MLR

## 6.1 **Building a non-spatial multiple linear regression**

```{r}
#| eval: false
price_mlr <- lm(resale_price ~ floor_area_sqm +
                  storey_order + remaining_lease_mths +
                  PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER +
                  PROX_MRT + PROX_PARK + PROX_MALL + 
                  PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
                  WITHIN_350M_CHILDCARE + WITHIN_350M_BUS +
                  WITHIN_1KM_PRISCH,
                data=train_data)
write_rds(price_mlr, "data/model/price_mlr.rds")
```

```{r}
#| echo: false
price_mlr <- read_rds("data/model/price_mlr.rds")
summary(price_mlr)
```

# 7 **Random Forest Model**

## 7.1 Prepare data

```{r}
coords <- st_coordinates(mdata)
coords_train <- st_coordinates(train_data)
coords_test <- st_coordinates(test_data)
```

```{r}
coords_train <- write_rds(coords_train, 
  "data/model/coords_train.rds" )
coords_test <- write_rds(
  coords_test, 
  "data/model/coords_test.rds" )
```

## 7.2 Dropping geometry field

We will drop geometry column of the sf data.frame by using `st_drop_geometry()` of sf package.

```{r}
train_data_nogeom <- train_data %>% 
  st_drop_geometry()

test_data_nogeom <- test_data %>% 
  st_drop_geometry()
```

## 7.3 Calibrating a non-spatial random forest model

```{r}
#| eval: false
set.seed(1234)
rf <- ranger(
  formula = resale_price ~ floor_area_sqm +
    storey_order + remaining_lease_mths + 
    PROX_CBD + PROX_ELDERLYCARE + 
    PROX_HAWKER + PROX_MRT + PROX_PARK + 
    PROX_MALL + PROX_SUPERMARKET +
    WITHIN_350M_KINDERGARTEN + 
    WITHIN_350M_CHILDCARE + 
    WITHIN_350M_BUS +  
    WITHIN_1KM_PRISCH,
  data=train_data_nogeom)

write_rds(rf, "data/model/rf.rds")
```

```{r}
#| echo: false

rf <- read_rds("data/model/rf.rds")
rf
```

## 7.4 **Calibrating Geographical Weighted Random Forest Model**

### 7.4.1 Calibrating with training data

Below we calibrate a **geographic random forest model** by using `grf()` of **SpatialML** package.

```{r}
#| eval: false
set.seed(1234)
gwRF_adaptive <- grf(
  formula = resale_price ~ floor_area_sqm +
    storey_order + remaining_lease_mths + 
    PROX_CBD + PROX_ELDERLYCARE + 
    PROX_HAWKER + PROX_MRT + PROX_PARK + 
    PROX_MALL + PROX_SUPERMARKET +
    WITHIN_350M_KINDERGARTEN +
    WITHIN_350M_CHILDCARE + 
    WITHIN_350M_BUS + WITHIN_1KM_PRISCH,
  dframe=train_data_nogeom, 
  bw=40,
  kernel="adaptive",
  coords=coords_train)

write_rds(gwRF_adaptive, "data/model/gwRF_adaptive.rds")
```

```{r}
#| echo: false

gwRF_adaptive <- read_rds("data/model/gwRF_adaptive.rds")

```

::::: markgoals
::: markgoals-header
Training results
:::

::: markgoals-container
-   **MLR vs GWR:** Use **AICc** → GWR wins (24,104 \< 24,902)
-   **RF vs GRF:** Use **OOB R²** → GRF slightly better (84.87% \> 84.11%)
-   Need to test predictions to determine which is the best.
:::
:::::

# 8 Predicting with test data

## 8.1 GW Radnom Forest

### 8.1.1 Prepare data

```{r}
test_data_nogeom <- cbind(
  test_data, coords_test) %>%
  st_drop_geometry()
```

### 8.1.2 Predicting with test data

```{r}
#| eval: false
gwRF_pred <- predict.grf(gwRF_adaptive, 
                           test_data_nogeom, 
                           x.var.name="X",
                           y.var.name="Y", 
                           local.w=1,
                           global.w=0)

write_rds(gwRF_pred, "data/model/GRF_pred.rds")
```

```{r}
#| echo: false
GRF_pred <- read_rds("data/model/GRF_pred.rds")
```

### 81.3 Converting the predicting output into a data frame

```{r}
GRF_pred <- read_rds("data/model/GRF_pred.rds")
GRF_pred_df <- as.data.frame(GRF_pred)
```

```{r}
test_data_p <- cbind(
  test_data, GRF_pred_df) %>%
  select(resale_price, GRF_pred)
```

### 8.1.4 Calculating Root Mean Square Error

```{r}
rmse(test_data_p$resale_price, 
     test_data_p$GRF_pred)
```

### 8.1.5 Visualising the predicted values

```{r}
ggplot(data = test_data_p,
       aes(x = GRF_pred,
           y = resale_price)) +
  geom_point()
```

## 8.1 GWR Predictive model

### 8.1.1 Computing predicted values of the test data

```{r}
gwr_pred <- gwr.predict(
  formula = resale_price ~ floor_area_sqm + 
    storey_order + remaining_lease_mths + 
    PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER + 
    PROX_MRT + PROX_PARK + PROX_MALL + 
    PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
    WITHIN_350M_CHILDCARE + WITHIN_350M_BUS + 
    WITHIN_1KM_PRISCH, 
  data=train_data, 
  predictdata = test_data, 
  bw=40, 
  kernel = 'gaussian', 
  adaptive=TRUE, 
  longlat = FALSE)

write_rds(gwr_pred, "data/model/gwr_pred.rds")
```
