---
title: "08a - Kick starter - Geographically Weighted Predictive Modelling"
author: "Cathy C."
date: "Oct 25, 2025"
date-modified: "last-modified" #auto render date
format: html
execute: #how YML page work
  echo: true #will show code
  eval: true #will run code every time and show output
  warning: false #dont show warning msg
  freeze: true #controls rendering - no render if no changes
---

## Installing and Loading R packages


```{r}
pacman::p_load(httr, tidyverse, sf, tmap, jsonlite, progress)
```

### Additional packages for predive modelling
```{r}
pacman::p_load(spdep, GWmodel, SpatialML, rsample, Metrics, knitr,
               kableExtra, spatialRF, randomForestExplainer)
```

progress - to monitor and track the computation.

## Geocoding HDB resale data
We use OneMap https://www.onemap.gov.sg/apidocs/
https://www.onemap.gov.sg/apidocs/reverseGeocode


### Basic version
```{r}
#| eval: false
HDBresale <- read_csv(
  "data/rawdata/HDBresale.csv") %>%
  filter(flat_type == "4 ROOM") %>%
  filter(month >= "2025-09" & month < "2025-10") # filter the sample data for the exercise.

write_rds(HDBresale, "data/rds/HDBresale.rds")
```
```{r}
HDBresale <- read_rds("data/rds/HDBresale.rds")
glimpse(HDBresale)
```

```{r}
# street names is ST, but it's actually Sait.
HDBresale$street_name <- gsub(
  "ST\\.", "SAINT",
  HDBresale$street_name
)
```

```{r}
# for multiple queries
geocode <- function(block, streetname) {
  base_url <- "https://onemap.gov.sg/api/common/elastic/search"
  address <- paste(block, streetname, sep = " ")
  query <- list("searchVal" = address,
                "returnGeom" = "Y",
                "getAddrDetails" = "N", #No. can change to Y for comparison
                "pageNum" = "1") 
  res <- GET(base_url, query = query) #get URL query
  restext <- content(res, as = "text")
  
  output <- fromJSON(restext) %>%
    as.data.frame %>%
    select(results.LATITUDE, results.LONGITUDE)
  
  return(output)
}
```

Need below to run the functions

```{r}
HDBresale$LATITUDE <- 0
HDBresale$LONGITUDE <- 0

for(i in 1:nrow(HDBresale)){
  temp_output <- geocode(HDBresale[i, 4], HDBresale[i, 5])
  
  HDBresale$LATITUDE[i] <- temp_output$results.LATITUDE
  HDBresale$LONGITUDE[i] <- temp_output$results.LONGITUDE
}
```

## Transform to sf data frame
```{r}
st_crs(HDBresale)
HDBresale_sf <- HDBresale %>%
  st_as_sf(
    coords = c("LONGITUDE", "LATITUDE"),
    crs = 4326) %>% #if decimal degree
  st_transform(crs = 3414)

```


## Load preschool data
```{r}
preschool = st_read("data/rawdata/PreSchoolsLocation.kml") %>%
  st_transform(crs = 3414)
```
From this we will have the geometry data for the preschools.

Next we will derive a binary data to find out which schools are within 350 meters.
```{r}
within_350m <- st_is_within_distance(
  HDBresale_sf, preschool, dist = 350)

HDBresale_sf <- HDBresale_sf %>%
  mutate(prischool_350m = map_int(within_350m, length)
)
```


# Predictive modeling
## Data import
```{r}
mdata <- read_rds("data/rawdata/mdata.rds")
```

## Data sampling
Calibrating predictive models are computational intensive, especially random forest method is used.
For quick prototyping, a 10% sample will be selected at random from the data by using the code chunk below:

```{r}
set.seed(1234)
HDB_sample <- mdata %>%
  sample_n(1500)
```

### Checking of overlapping point

*** When using GWmodel to calibrate explanaroty or predictive models, it is very important to ensure that there are no overlapping point features *** 

Below is to check if there are overlapping point feature - to do spatial jitter. 

```{r}
overlapping_points <- HDB_sample %>%
  mutate(overlap = lengths(st_equals(.,.)) > 1)

```

Below we can have a look at overlapping_points, and see the new column overlap showing whether the location has overlapping points. We then calculate how many has overlapping points. In this sample data, we observed 453 overlapping points. 
```{r}
head(overlapping_points)
sum(overlapping_points$overlap == "TRUE")
```

## Spatial jitter

st_jitter is used to move the point feature by 5m to avoid overlapping point features.

```{r}
HDB_sample <- HDB_sample %>%
  st_jitter(amount = 1)
```

```{r}
set.seed(1234)
resale_split <- initial_split(HDB_sample,
                              prop = 6.67/10,)
train_data <- training(resale_split)
test_data <- testing(resale_split)
```


### Constructing the adaptive bandwidth gwr model

```{r}
#| eval: false
bw_adaptive <- bw.gwr(
  formula = resale_price ~ floor_area_sqm +
    storey_order + remaining_lease_mths +
    PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER +
    PROX_MRT + PROX_PARK + PROX_MALL + 
    PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
    WITHIN_350M_CHILDCARE + WITHIN_350M_BUS +
    WITHIN_1KM_PRISCH,
  data=train_data,
  approach="CV",
  kernel="gaussian",
  adaptive=TRUE,
  longlat=FALSE)

write_rds(bw_adaptive, "data/model/bw_adaptive.rds")
```

### Constructing the adaptive bandwidth gwr model
```{r}
#| echo: false
bw_adaptive <- read_rds(
  "data/model/bw_adaptive.rds")
```

```{r}
#| eval: false
gwr_adaptive <- gwr.basic(
  formula = resale_price ~ floor_area_sqm +
    storey_order + remaining_lease_mths + 
    PROX_CBD + PROX_ELDERLYCARE + 
    PROX_HAWKER + PROX_MRT + PROX_PARK + 
    PROX_MALL + PROX_SUPERMARKET +
    WITHIN_350M_KINDERGARTEN +
    WITHIN_350M_CHILDCARE + 
    WITHIN_350M_BUS + WITHIN_1KM_PRISCH,
  data=train_data,
  bw=bw_adaptive, 
  kernel = 'gaussian', 
  adaptive=TRUE,
  longlat = FALSE)

write_rds(gwr_adaptive, 
          "data/model/gwr_adaptive.rds")
```

### Output

```{r}
#| echo: false
gwr_adaptive <- read_rds("data/model/gwr_adaptive.rds")
gwr_adaptive
```

- Based on the result of the modelling above, p-value is much smaller than 0.05. 
We see the adjusted R^2 = 0.7387, meaning that this model can explain 73% of the data using the variables. 

- The AICc = 24901.56 for Global Regression model is much larger than Geographically Weighted Regression's AICc = 24103.97. The difference of approximately 797.6 points is quite significant. This suggests the **GWR model provides meaningfully better predictions and fit**. 


Explanatory variables:

Plot out the significant variables to show distribution and explain which ones have higher/lower influence over the model. 

## Collinearity

```{r}

```

## Plot VIF
```{.r}
plot(vif) +
  theme(axis.test.x - element_text(
    angle = 45, hjust = 1
  ))
```


## Predive Modeling using MLR
