---
title: "03 - Spatio-Temporal Point Patterns Analysis"
author: "Cathy C."
date: "Sep 9, 2025"
date-modified: "last-modified" #auto render date
format: html
execute: #how YML page work
  echo: true #will show code
  eval: true #will run code every time and show output
  warning: false #dont show warning msg
  freeze: true #controls rendering - no render if no changes
---

## Overview

A **spatio-temporal point process** (also called space-time or spatial-temporal point process) is a **random collection of points**, where each point represents the **time** and **location** of an event. Examples of events include incidence of disease, sightings or births of a species, or the occurrences of fires, earthquakes, lightning strikes, tsunamis, or volcanic eruptions.

The analysis of spatio-temporal point patterns is becoming increasingly necessary, given the rapid emergence of geographically and temporally indexed data in a wide range of fields. Several spatio-temporal point patterns analysis methods have been introduced and implemented in R in the last ten years.

In this exercise, various R packages will be combined to run a set of **spatio-temporal point pattern analyses**.

A real world forest fire events in Kepulauan Bangka Belitung, Indonesia from 1st January 2023 to 31st December 2023 is used to illustrate the methods, procedures and interpretations.

## 1 Objectives

Specific questions we would like to answer are:

-   are the locations of forest fire in Kepulauan Bangka Belitung spatial and spatio-temporally **independen**t?

-   if the answer is NO, **where and when** the observed forest fire locations tend to cluster?

## 2 The data

For the purpose of this exercise, two data sets will be used, they are:

-   *forestfires*, a csv file provides locations of forest fire detected from the Moderate Resolution Imaging Spectroradiometer (MODIS) sensor data. The data are downloaded from [Fire Information for Resource Management System](https://firms.modaps.eosdis.nasa.gov/download/). Only forest fires within Kepulauan Bangka Belitung will be used for this exercise.

-   *Kepulauan_Bangka_Belitung*, an **ESRI** **shapefile** showing the sub-district (i.e. [kelurahan](https://en.wikipedia.org/wiki/Villages_of_Indonesia)) boundary of Kepulauan Bangka Belitung. The data set was downloaded from [Indonesia Geospatial](https://www.indonesia-geospasial.com/2023/05/download-shapefile-batas-administrasi.html) portal. The original data covers the whole Indonesia. Only sub-districts within Kepulauan Bangka Belitung are extracted for this exercise.

## 3 Installing and Loading the R packages

| R Package | Description |
|--------------------|:---------------------------------------------------|
| [**sf**](https://r-spatial.github.io/sf/) | a relatively new R package specially designed to import, manage and process vector-based geospatial data in R. |
| [**raster**](https://rspatial.org/raster/) | handling raster data in R. |
| [**spatstat**](https://cran.r-project.org/web/packages/spatstat/index.html) | performing Spatial Point Patterns Analysis such as kcross, Lcross, etc. |
| [**sparr**](https://tilmandavies.github.io/sparr/index.html) | provides functions to estimate fixed and adaptive kernel-smoothed spatial relative risk surfaces via the density-ratio method and perform subsequent inference. Fixed-bandwidth spatiotemporal density and relative risk estimation is also supported. |
| [**tmap**](https://r-tmap.github.io/tmap/) | provides functions to produce cartographic quality thematic maps. |
| [**tidyverse**](https://www.tidyverse.org/) | a family of R packages that provide functions to perform common data science tasks including and not limited to data import, data transformation, data wrangling and data visualisation. |

: In this hands-on exercise, six R packages will be used, they are: {tbl-colwidths="\[20,80\]"}

```{r}
pacman::p_load(sf, raster, sparr, tmap, tidyverse, spatstat)
```

## 4 Importing and Preparing Study Area

### 4.1 Importing study area

First import study area (i.e. Kepulauan Bangka Belitung) into R environment and transform the data.

```{r}
kbb_sf <- st_read(dsn="data/rawdata",
               layer = "Kepulauan_Bangka_Belitung") %>%
  st_union() %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 32748)
```

### 4.2 Converting OWIN

Next, `as.owin()` is used to convert *kbb* into an **owin** object.

```{r}
kbb_owin <- as.owin(kbb_sf)
kbb_owin
```

Let's confirm kbb_own is indeed an *OWIN* object.

```{r}
class(kbb_owin)
```

## 5 Importing and Preparing Forest Fire data

Next, we will import the forest fire data set (i.e. *forestfires.csv*) into R environment.

```{r}
fire_sf <- read_csv("data/rawdata/forestfires.csv") %>%
  st_as_sf(coords = c("longitude", "latitude"), #convert to spatial format
           crs = 4326) %>% #WGS84
  st_transform(crs = 32748) #transform to UTM meter system
```

The result is a spatial object where each forest fire location is now a point geometry with coordinates in the UTM 48S projection system. Distance and area calculations are more accurate in projected systems.

Because **ppp** object only accepts numerical or character as mark. The code chunk below is used to convert data type of ***acq_date*** to **numeric.**

```{r}
fire_sf <- fire_sf %>%
  mutate(DayofYear = yday(acq_date)) %>%
  mutate(Month_num = month(acq_date)) %>%
  mutate(Month_fac = month(acq_date,
                            label = TRUE, abbr = FALSE))
```

## 6 Visualising the Fire Points

### 6.1 Overall plot

::: panel-tabset
## DIY

```{r}
#| fig-align: center

par(bg = "#EEEEEE", mar = c(1, 1, 1, 1))
plot(st_geometry(kbb_sf), col = "white")
plot(st_geometry(fire_sf), add = TRUE, col = "#BA487F", pch = 16, cex = 0.45)

```

## Key

```{r}
par(bg = "#EEEEEE")
tm_shape(kbb_sf)+
  tm_polygons() +
tm_shape(fire_sf) +
  tm_dots() +
tm_layout(
    outer.bg.color = "#EEEEEE"
  )
```
:::

### 6.2 Visuaising geographic distribution of forest fires by month

::: panel-tabset
## DIY

```{r}
#| fig-align: center

tmap_options(bg.color = "#EEEEEE")
tm_shape(kbb_sf)+
  tm_polygons(fill = "#A8BBA3") +
tm_shape(fire_sf) +
  tm_dots(fill = "#D25D5D", size = 0.2) +
  tm_facets(by = "Month_fac",
            nrow = 3, ncols = 4,
            free.coords = FALSE, drop.units = TRUE) 
```

## Key

```{r}
tm_shape(kbb_sf)+
  tm_polygons() +
tm_shape(fire_sf) +
  tm_dots(size = 0.1) +
tm_facets(by="Month_fac", 
            free.coords=FALSE, 
            drop.units = TRUE)
```
:::

## 7 Computing STKDE by Month

In this section, we will compute STKDE by using [`spattemp.density()`](https://tilmandavies.github.io/sparr/reference/spattemp.density.html) of **sparr** package. Before using the function, it is highly recommended you read the function’s reference guide in detail in order to understand the input data requirements and the output object generated.

### 7.1 **Extracting forest fires by month**

`as.ppp()` only need the **mark field** and **geometry field** from the input sf data.frame, so below we want to remove the unwanted fields from *fire_sf* sf data.frame.

```{r}
fire_month <- fire_sf %>%
  select(Month_num)
```

### 7.2 **Creating ppp**

Next we will derive a **ppp** object called *fire_month* from *fire_month* sf data.frame.

```{r}
fire_month_ppp <- as.ppp(fire_month)
fire_month_ppp
```

Use below code to check the output is in the correct object class.

```{r}
summary(fire_month_ppp)
```

We also want to check if there are duplicated point events.

```{r}
any(duplicated(fire_month_ppp))
```

### 7.3 **Including Owin object**

We use the code below to combine *origin_am_ppp* and *am_owin* objects into one.

```{r}
fire_month_owin <- fire_month_ppp[kbb_owin]
summary(fire_month_owin)
```

Next we will examine the correctness of the output.

```{r}
par(bg = "#D0DDD0")
plot(fire_month_owin)
```

### 7.4 **Computing Spatio-temporal KDE**

Then, `spattemp.density()` of **sparr** package is used to compute the STKDE.

```{r}
st_kde <- spattemp.density(fire_month_owin)
summary(st_kde)
```

### 7.5 **Plotting the spatio-temporal KDE object**

We use `plot()` to plot the KDE for between July 2023 - December 2023.

```{r}
par(bg = "#D0DDD0")
tims <- c(7, 8, 9, 10, 11, 12)
par(mfcol = c(2,3))
for (i in tims){
  plot(st_kde, i,
       override.par = FALSE,
       fix.range = TRUE)
  title(main = paste("Month", i))
}
```

## 8 Computing STKDE by Day of Year

In this section, we will compute the STKDE of forest fires by day of year.

### 8.1 **Creating ppp object**

DayofYear field is included in the output ppp object with the code below.

```{r}
fire_yday_ppp <- fire_sf %>%
  select(DayofYear) %>%
  as.ppp()

```

### 8.2 **Including Owin object**

Next we will combine the ppp object and the owin object.

```{r}
fire_yday_owin <- fire_yday_ppp[kbb_owin]
summary(fire_yday_owin)
```

### 8.3 Plot KDE object

```{r}
kde_yday <- spattemp.density(fire_yday_owin)
summary(kde_yday)
```

**Let's plot it.**

```{r}
#| fig-width: 5
#| fig-height: 3
par(bg = "#D0DDD0")
plot(kde_yday)
```

## 9 Computing STKDE by Day of Year: Improved method
