---
title: "03 - Spatio-Temporal Point Patterns Analysis"
author: "Cathy C."
date: "Sep 9, 2025"
date-modified: "last-modified" #auto render date
format: html
execute: #how YML page work
  echo: true #will show code
  eval: true #will run code every time and show output
  warning: false #dont show warning msg
  freeze: true #controls rendering - no render if no changes
---

## Overview

A **spatio-temporal point process** (also called space-time or spatial-temporal point process) is a **random collection of points**, where each point represents the **time** and **location** of an event. Examples of events include incidence of disease, sightings or births of a species, or the occurrences of fires, earthquakes, lightning strikes, tsunamis, or volcanic eruptions.

The analysis of spatio-temporal point patterns is becoming increasingly necessary, given the rapid emergence of geographically and temporally indexed data in a wide range of fields. Several spatio-temporal point patterns analysis methods have been introduced and implemented in R in the last ten years.

In this exercise, various R packages will be combined to run a set of **spatio-temporal point pattern analyses**.

A real world forest fire events in Kepulauan Bangka Belitung, Indonesia from 1st January 2023 to 31st December 2023 is used to illustrate the methods, procedures and interpretations.

## 1 Objectives

Specific questions we would like to answer are:

-   are the locations of forest fire in Kepulauan Bangka Belitung spatial and spatio-temporally **independen**t?

-   if the answer is NO, **where and when** the observed forest fire locations tend to cluster?

## 2 The data

For the purpose of this exercise, two data sets will be used, they are:

-   *forestfires*, a csv file provides locations of forest fire detected from the Moderate Resolution Imaging Spectroradiometer (MODIS) sensor data. The data are downloaded from [Fire Information for Resource Management System](https://firms.modaps.eosdis.nasa.gov/download/). Only forest fires within Kepulauan Bangka Belitung will be used for this exercise.

-   *Kepulauan_Bangka_Belitung*, an **ESRI** **shapefile** showing the sub-district (i.e. [kelurahan](https://en.wikipedia.org/wiki/Villages_of_Indonesia)) boundary of Kepulauan Bangka Belitung. The data set was downloaded from [Indonesia Geospatial](https://www.indonesia-geospasial.com/2023/05/download-shapefile-batas-administrasi.html) portal. The original data covers the whole Indonesia. Only sub-districts within Kepulauan Bangka Belitung are extracted for this exercise.

## 3 Installing and Loading the R packages

| R Package | Description |
|-----------------|:------------------------------------------------------|
| [**sf**](https://r-spatial.github.io/sf/) | a relatively new R package specially designed to import, manage and process vector-based geospatial data in R. |
| [**raster**](https://rspatial.org/raster/) | handling raster data in R. |
| [**spatstat**](https://cran.r-project.org/web/packages/spatstat/index.html) | performing Spatial Point Patterns Analysis such as kcross, Lcross, etc. |
| [**sparr**](https://tilmandavies.github.io/sparr/index.html) | provides functions to estimate fixed and adaptive kernel-smoothed spatial relative risk surfaces via the density-ratio method and perform subsequent inference. Fixed-bandwidth spatiotemporal density and relative risk estimation is also supported. |
| [**tmap**](https://r-tmap.github.io/tmap/) | provides functions to produce cartographic quality thematic maps. |
| [**tidyverse**](https://www.tidyverse.org/) | a family of R packages that provide functions to perform common data science tasks including and not limited to data import, data transformation, data wrangling and data visualisation. |

: In this hands-on exercise, six R packages will be used, they are: {tbl-colwidths="\[20,80\]"}

```{r}
pacman::p_load(sf, raster, sparr, tmap, tidyverse, spatstat)
```

## 4 Importing and Preparing Study Area

### 4.1 Importing study area

First import study area (i.e. Kepulauan Bangka Belitung) into R environment and transform the data.

```{r}
kbb_sf <- st_read(dsn="data/rawdata",
               layer = "Kepulauan_Bangka_Belitung") %>%
  st_union() %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 32748)
```

### 4.2 Converting OWIN

Next, `as.owin()` is used to convert *kbb* into an **owin** object.

```{r}
kbb_owin <- as.owin(kbb_sf)
kbb_owin
```

Let's confirm kbb_own is indeed an *OWIN* object.

```{r}
class(kbb_owin)
```

## 5 Importing and Preparing Forest Fire data

Next, we will import the forest fire data set (i.e. *forestfires.csv*) into R environment.

```{r}
fire_sf <- read_csv("data/rawdata/forestfires.csv") %>%
  st_as_sf(coords = c("longitude", "latitude"), #convert to spatial format
           crs = 4326) %>% #WGS84
  st_transform(crs = 32748) #transform to UTM meter system
```

The result is a spatial object where each forest fire location is now a point geometry with coordinates in the UTM 48S projection system. Distance and area calculations are more accurate in projected systems.

## 6 Visualising the Fire Points

## 7 Computing STKDE by Month

## 8 Computing STKDE by Day of Year

## 9 Computing STKDE by Day of Year: Improved method
