---
title: "02b - Second Order Spatial Point Patterns Analysis Methods"
author: "Cathy C."
date: "Sep 3, 2025"
date-modified: "last-modified" #auto render date
format: html
execute: #how YML page work
  echo: true #will show code
  eval: true #will run code every time and show output
  warning: false #dont show warning msg
  freeze: true #controls rendering - no render if no changes
---

## Overview

**Second-order spatial point pattern analysis** Second-order spatial point pattern analysis examines the spatial relationships between points in a pattern, specifically focusing on how the presence of one point influences the location of others. It goes beyond simply describing the overall density of points (first-order effects) by investigating clustering, dispersion, or randomness at various spatial scales.

## 1 Getting started

Using appropriate functions of [spatstat](https://cran.r-project.org/web/packages/spatstat/), in this hands-on exercise we aim to discover the **spatial point processes** of childecare centres in Singapore.

The specific questions we would like to answer are as follows:

-   are the childcare centres in Singapore randomly distributed throughout the country?
-   if the answer is not, then the next logical question is where are the locations with higher concentration of childcare centres?

## 2 The data

We will use two data sets for the analysis.

-   *Child Care Services* data from data.gov.sg, a **point feature data** providing both location and attribute information of childcare centres.

-   *Master Plan 2019 Subzone Boundary (No Sea)*, a **polygon feature data** providing information of URA 2019 Master Plan Planning Subzone boundary data.

Both data sets are available at [Singapore’s open data portal](data.gov.sg). They are provided in kml and geojson format.

## 3 Installing and Loading the R packages

+---------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| R Package                                                           | Description                                                                                                                                                                                                                                                                               |
+=====================================================================+:==========================================================================================================================================================================================================================================================================================+
| [**sf**](https://r-spatial.github.io/sf/)                           | a relatively new R package specially designed to import, manage and process vector-based geospatial data in R.                                                                                                                                                                            |
+---------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [**spatstat**](https://spatstat.org/)                               | has a wide range of useful functions for point pattern analysis. In this hands-on exercise, it will be used to perform 1st- and 2nd-order spatial point patterns analysis and derive kernel density estimation (KDE) layer.                                                               |
+---------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [**tmap**](https://cran.r-project.org/web/packages/tmap/index.html) | provides functions for plotting cartographic quality static point patterns maps or interactive maps by using [**leaflet**](https://leafletjs.com/) API.                                                                                                                                   |
+---------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [**tidyverse**](https://www.tidyverse.org/)                         | a family of R packages designed for modern data science. These packages are developed to work together seamlessly, sharing a common design philosophy, grammar, and data structures, which aims to make data manipulation, analysis, and visualization in R more intuitive and efficient. |
+---------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [**rvest**](https://rvest.tidyverse.org/)                           | for scraping (or harvesting) data from web pages                                                                                                                                                                                                                                          |
|                                                                     |                                                                                                                                                                                                                                                                                           |
|                                                                     | \*need this to read html.                                                                                                                                                                                                                                                                 |
+---------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+

: In this hands-on exercise, four R packages will be used, they are: {tbl-colwidths="\[20,80\]"}

```{r}
pacman::p_load(sf, spatstat, tmap, tidyverse, rvest)
```

## 4 Data Import and Preparation

The processes of importing and preparing the geospatial data to meet the analysis are similar to Exercise 02a.

-   [Importing and Wrangling Geospatial Data Sets](https://catgeo.netlify.app/hands-on_ex/hands-on_ex02/hands-on_ex02a#importing-and-wrangling-geospatial-data-sets)
-   [Geospatial Data wrangling](https://catgeo.netlify.app/hands-on_ex/hands-on_ex02/hands-on_ex02a#geospatial-data-wrangling) - to prepare geospatial data in ppp object class required by spatstat package.

```{r}
#| code-fold: true
#| code-summary: "Show code for data wranggling"
mpsz_sf <- st_read("data/MasterPlan2019SubzoneBoundaryNoSeaKML.kml") %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 3414)

extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)
  rows <- page %>% html_elements("tr")
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>%
    html_element("td") %>%
    html_text2()
  
  if (length(value) == 0) NA_character_ else value
}

mpsz_sf <- mpsz_sf %>%
  mutate(
    REGION_N = map_chr(Description, extract_kml_field, "REGION_N"),
    PLN_AREA_N = map_chr(Description, extract_kml_field, "PLN_AREA_N"),
    SUBZONE_N = map_chr(Description, extract_kml_field, "SUBZONE_N"),
    SUBZONE_C = map_chr(Description, extract_kml_field, "SUBZONE_C")
  ) %>%
  select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())

mpsz_cl <- mpsz_sf %>%
  filter(SUBZONE_N != "SOUTHERN GROUP",
         PLN_AREA_N != "WESTERN ISLANDS",
         PLN_AREA_N != "NORTH-EASTERN ISLANDS")

childcare_sf <- st_read("data/ChildCareServices.kml") %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 3414)
```

```{r}
#| code-fold: true
#| code-summary: "Show code for ppp conversion"
#convert to ppp
childcare_ppp <- as.ppp(childcare_sf)

#create owin object
sg_owin <- as.owin(mpsz_cl)

#combien point events object and owin object
childcareSG_ppp = childcare_ppp[sg_owin]

#extract study areas
pg <- mpsz_cl %>%
  filter(PLN_AREA_N == "PUNGGOL")
tm <- mpsz_cl %>%
  filter(PLN_AREA_N == "TAMPINES")
ck <- mpsz_cl %>%
  filter(PLN_AREA_N == "CHOA CHU KANG")
jw <- mpsz_cl %>%
  filter(PLN_AREA_N == "JURONG WEST")

```

```{r}
#| code-fold: true
#| code-summary: "Show code for plotting"
#| fig-width: 12
#| fig-height: 8
#create owin object
pg_owin = as.owin(pg)
tm_owin = as.owin(tm)
ck_owin = as.owin(ck)
jw_owin = as.owin(jw)

#combine point events object and owin object
childcare_pg_ppp = childcare_ppp[pg_owin]
childcare_tm_ppp = childcare_ppp[tm_owin]
childcare_ck_ppp = childcare_ppp[ck_owin]
childcare_jw_ppp = childcare_ppp[jw_owin]

#transform measurement to km
childcare_pg_ppp.km = rescale.ppp(childcare_pg_ppp, 1000, "km")
childcare_tm_ppp.km = rescale.ppp(childcare_tm_ppp, 1000, "km")
childcare_ck_ppp.km = rescale.ppp(childcare_ck_ppp, 1000, "km")
childcare_jw_ppp.km = rescale.ppp(childcare_jw_ppp, 1000, "km")

#plot
par(bg = "#e6f0e6", mfrow = c(2,2), mar = c(1, 1, 2, 1))
plot(unmark(childcare_pg_ppp.km), main = "Punggol")
plot(unmark(childcare_tm_ppp.km), main = "Tampines")
plot(unmark(childcare_ck_ppp.km), main = "Choa Chu Kang")
plot(unmark(childcare_jw_ppp.km), main = "Jurong West")
```

## 6 Analysing Spatial Point Process Using **G-Function**

The G function measures the **distribution of the distances** from an arbitrary event to its nearest event. In this section, we will compute G-function estimation by using [*Gest()*](https://rdrr.io/cran/spatstat/man/Gest.html) of **spatstat** package. We will also perform Monte Carlo simulation test using [*envelope()*](https://rdrr.io/cran/spatstat/man/envelope.html) of **spatstat** package.

::: panel-tabset
## 6.1 Choa Chu Kang planning area

#### 6.1.1 Computing G-function estimation

```{r}
set.seed(1234)
```

Compute G-function using *Gest()* of **spatstat** package.
```{r}
par(bg = "#f5e6e6")
G_CK = Gest(childcare_ck_ppp, correction = "border")
plot(G_CK, xlim = c(0,500))
```

#### 6.1.2 Performing Complete Spatial Randomness Test
To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:
H₀ = The distribution of childcare services at Choa Chu Kang are randomly distributed.
H₁ = The distribution of childcare services at Choa Chu Kang are not randomly distributed.

The null hypothesis will be rejected if *p*-value is smaller than alpha value of 0.001.
Monte Carlo test with G-fucntion

```{r}
par(bg = "#f5e6e6")
G_CK.csr <- envelope(childcare_ck_ppp, Gest, nsim = 999)
```
**Let's plot the figure.**
```{r}
par(bg = "#f5e6e6")
plot(G_CK.csr)
```


## 6.2 Tampines planning area

#### 6.2.1 Computing G-function estimation

#### 6.2.2 Performing Complete Spatial Randomness Test
:::

## 7 Analysing Spatial Point Process Using **F-Function**

::: panel-tabset
## 7.1 Choa Chu Kang planning area

#### 7.1.1 Computing F-function estimation

#### 7.1.2 Performing Complete Spatial Randomness Test

## 7.2 Tampines planning area

#### 7.2.1 Computing F-function estimation

#### 7.2.2 Performing Complete Spatial Randomness Test
:::

## 8 Analysing Spatial Point Process Using **K-Function**

::: panel-tabset
## 8.1 Choa Chu Kang planning area

#### 8.1.1 Computing K-fucntion estimate

#### 8.1.2 Performing Complete Spatial Randomness Test

## 8.2 Tampines planning area

#### 8.2.1 Computing K-fucntion estimation

#### 8.2.2 Performing Complete Spatial Randomness Test
:::

## 9 Analysing Spatial Point Process Using **L-Function**

::: panel-tabset
## 9.1 Choa Chu Kang planning area

#### 9.1.1 Computing **L Fucntion** estimation

#### 9.1.2 Performing Complete Spatial Randomness Test

## 9.2 Tampines planning area

#### 9.2.1 Computing L-fucntion estimate

#### 9.2.2 Performing Complete Spatial Randomness Test
:::
