---
title: "Hands-on Exercise 02a"
author: "Cathy C."
date: "Sep 1, 2025"
date-modified: "last-modified" #auto render date
format: html
execute: #how YML page work
  echo: true #will show code
  eval: true #will run code every time and show output
  warning: false #dont show warning msg
  freeze: true #controls rendering - no render if no changes
---

## Overview

**Spatial Point Pattern Analysis (SPPA)** is the evaluation of the pattern or distribution of a set of points on a surface. The points may represent:

-   events such as crimes, traffic accidents, or disease onsets, or
-   business services (e.g., coffee shops and fast-food outlets) or facilities such as childcare centres and eldercare centres.

**First-order Spatial Point Pattern Analysis (1st-SPPA)** focuses on understanding the intensity or **density of points** across a study area. It examines how the distribution of points varies over space, essentially identifying trends or patterns in point density. This type of analysis deals with the individual locations of points and their distribution, without considering interactions between them.

In essence, 1st-SPPA helps answer questions such as:

-   Where are points most densely located within the study area?
-   Is point density uniform, or does it vary across space?
-   How spread out is the point pattern?

## 1 Getting started

In this chapter, we will gain hands-on experience with [**spatstat**](https://cran.r-project.org/web/packages/spatstat/) to perform two commonly used 1st-SPPA methods:

The specific questions we would like to answer are as follows:

-   Are the childcare centres in Singapore randomly distributed throughout the country?
-   If the answer is not, then the next logical question is where are the locations with higher concentration of childcare centres?

## 2 The data

We will use two data sets to answer the questions above:

-   *Child Care Services* data from data.gov.sg, a point feature data providing both location and attribute information of childcare centres.

-   *Master Plan 2019 Subzone Boundary (No Sea)*, a polygon feature data providing information of URA 2019 Master Plan Planning Subzone boundary data.

Both data sets are provided in **kml** and **geojson** format.

## 1.3 Installing and Loading the R packages

| R Package | Description |
|----|:---|
| [**sf**](https://r-spatial.github.io/sf/) | a relatively new R package specially designed to import, manage and process vector-based geospatial data in R. |
| [**spatstat**](https://spatstat.org/) | has a wide range of useful functions for point pattern analysis. In this hands-on exercise, it will be used to perform 1st- and 2nd-order spatial point patterns analysis and derive kernel density estimation (KDE) layer. |
| [**terra**](https://rspatial.github.io/terra/index.html) | a modern spatial data analysis package designed to replace the raster package. It offers improved speed and efficiency when working with both raster and vector spatial data, particularly with large datasets. terra provides functionalities for creating, reading, manipulating, and writing raster and vector data, and it’s built on top of GDAL and PROJ libraries for enhanced performance. In this hands-on exercise, it will be used to convert image output generate by spatstat into terra format. |
| [**tmap**](https://cran.r-project.org/web/packages/tmap/index.html) | provides functions for plotting cartographic quality static point patterns maps or interactive maps by using [**leaflet**](https://leafletjs.com/) API. |
| [**rvest**](https://rvest.tidyverse.org/) | for scraping (or harvesting) data from web pages |

: In this hands-on exercise, five R packages will be used, they are: {tbl-colwidths="\[20,80\]"}

```{r}
pacman::p_load(sf, terra, spatstat, tmap, rvest, tidyverse)
```

## 4 Importing and Wrangling Geospatial Data Sets

::::: taskgoals
::: taskgoals-header
**TASK**
:::

::: taskgoals-container
-   Import the two geospatial datasets into R using the methods from Hands-on Exercises 1 and 2.
-   Tidy the datasets and, if necessary, transform their projections.
-   Validate that both datasets use the same projection.
-   Ensure that the final outputs are stored in sf format.
:::
:::::

First we import Master Plan 2019 Subzone (No Sea) data set into R environment.

### 4.1 Mapping the geospatial data sets

```{r}
mpsz_sf <- st_read("data/MasterPlan2019SubzoneBoundaryNoSeaKML.kml") %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 3414)
```

Learning:

`st_zm()` is used to used to remove Z (elevation) and M (measure) dimensions from geospatial geometries.

Next, we build a function called *extract_kml_field* for extracting REGION_N, PLN_AREA_N, SUBZONE_N, SUBZONE_C from Description field with the code below.

```{r}
extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)
  rows <- page %>% html_elements("tr")
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>%
    html_element("td") %>%
    html_text2()
  
  if (length(value) == 0) NA_character_ else value
}
```

```{r}
mpsz_sf <- mpsz_sf %>%
  mutate(
    REGION_N = map_chr(Description, extract_kml_field, "REGION_N"),
    PLN_AREA_N = map_chr(Description, extract_kml_field, "PLN_AREA_N"),
    SUBZONE_N = map_chr(Description, extract_kml_field, "SUBZONE_N"),
    SUBZONE_C = map_chr(Description, extract_kml_field, "SUBZONE_C")
  ) %>%
  select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())
```

```{r}
mpsz_cl <- mpsz_sf %>%
  filter(SUBZONE_N != "SOUTHERN GROUP",
         PLN_AREA_N != "WESTERN ISLANDS",
         PLN_AREA_N != "NORTH-EASTERN ISLANDS")
```

```{r}
write_rds(mpsz_cl,"data/mpsz_cl.rds")
```

## 5 Geospatial Data wrangling

### 5.1 Converting sf data frames to ppp class

### 5.2 Creating owin object

### 5.3 Combining point events object and owin object

## 6 Clark-Evan Test for Nearest Neighbour Analysis

### 6.1 Perform the Clark-Evans test without CSR

### 6.2 Perform the Clark-Evans test with CSR

## 7 Kernel Density Estimation Method

### 7.1 Working with automatic bandwidth selection method

### 7.2 Rescalling KDE values

### 7.3 Working with different automatic badwidth methods

### 7.4 Working with different kernel methods

## 8 Fixed and Adaptive KDE

### 8.1 Computing KDE by using fixed bandwidth

### 8.2 Computing KDE by using adaptive bandwidth

## 9 Plotting cartographic quality KDE map

### 9.1 Converting gridded output into raster

### 9.2 Assigning projection systems

### 9.3 Plotting KDE map with tmap

## 10 First Order SPPA at the Planning Subzone Level

### 10.1 Geospatial data wrangling

#### 10.1.1 Extracting study area

#### 10.1.2 Creating owin object

#### 10.1.3 Combining point events object and owin object

### 10.2 Clark and Evans Test

#### 10.2.1 Choa Chu Kang planning area

#### 10.2.2 Tampines planning area

### 10.3 Computing KDE surfaces by planning area
