---
title: "05b - Local Measures of Spatial Autocorrelation"
author: "Cathy C."
date: "Sep 29, 2025"
date-modified: "last-modified" #auto render date
format: html
execute: #how YML page work
  echo: true #will show code
  eval: true #will run code every time and show output
  warning: false #dont show warning msg
  freeze: true #controls rendering - no render if no changes
---

## 1 Overview

Local Measures of Spatial Autocorrelation (LMSA) focus on the **relationships between each observation and its surroundings**, rather than providing a single summary of these relationships across the map. In this sense, they are not summary statistics but scores that allow us to learn more about the spatial structure in our data.

The general intuition behind the metrics however is similar to that of global ones. Some of them are even mathematically connected, where the global version can be decomposed into a collection of local ones. One such example are **Local Indicators of Spatial Association (LISA)**. Beside LISA, **Getis-Ord’s Gi-statistics** will be introduce as an alternative LMSA statistics that present complementary information or allow us to obtain similar insights for geographically referenced data.

### Objectives

In this hands-on exercise, we will compute Local Measures of Spatial Autocorrelation (LMSA) by using [**sfdep**](https://sfdep.josiahparry.com/) package. We wil do the following:

-   import geospatial data using appropriate function(s) of **sf** package,
-   import csv file using appropriate function of **readr** package,
-   perform relational join using appropriate join function of **dplyr** package,
-   compute Local Indicator of Spatial Association (LISA) statistics for detecting clusters and outliers by using appropriate functions **sfdep** package;
-   compute Getis-Ord’s Gi-statistics for detecting hot spot or/and cold spot area by using appropriate functions of **sfdep** package; and to visualise the analysis output by using **tmap** package.

## 2 Getting started

### 2.1 The analytical question

In spatial policy, one of the main development objective of the local govenment and planners is to ensure equal distribution of development in the province. Our task in this study, hence, is to apply appropriate spatial statistical methods to **discover if development are evenly distributed geographically**. If the answer is **No**. Then, our next question will be [“is there sign of spatial clustering?”]{style="color: #B95E82;"}. And, if the answer for this question is **yes**, then our next question will be [“where are these clusters?”]{style="color: #B95E82;"}

In this case study, we are interested to examine the spatial pattern of a selected development indicator (i.e. GDP per capita) of Hunan Provice, People Republic of China.(https://en.wikipedia.org/wiki/Hunan)

### 2.2 Setting the Analytical Tools

We will use **sf**, **sfdep**, **tmap** and **tidyverse** packages of R.

| R package | Description |
|------------------|------------------------------------------------------|
| sf | importing and handling geospatial data in R |
| tidyverse | wrangling attribute data in R |
| sfdep | compute spatial weights, global and local spatial autocorrelation statistics |
| tmap | prepare cartographic quality choropleth map |

The code chunk below is used to perform the following tasks:

-   creating a package list containing the necessary R packages,
-   checking if the R packages in the package list have been installed in R, if they have yet to be installed, RStudio will installed the missing packages,
-   launching the packages into R environment.

```{r}
pacman::p_load(sf, tidyverse, sfdep, tmap)
```

## 3 The data

### 3.1 The Study Area and Data

Two data sets will be used in this hands-on exercise, they are:

-   Hunan province administrative boundary layer at county level. This is a geospatial data set in ESRI shapefile format.

-   Hunan_2012.csv: This csv file contains selected Hunan’s local development indicators in 2012.

### 3.2 Import shapefile into R environment

The code chunk below uses [`st_read()`](https://r-spatial.github.io/sf/reference/st_read.html) of **sf** package to import Hunan shapefile into R. The imported shapefile will be **simple features** Object of **sf**.

```{r}
hunan <- st_read(dsn = "data/geospatial", layer = "Hunan") %>%
  st_transform(crs = 32650)
```

::::: markgoals
::: markgoals-header
**Tips**
:::

::: markgoals-container
The raw data is in WGS 84 geographic coordinates system. For geospatial analysis, it is appropriate to use projected coordinates system. In the code chunk above, `st_transform()` is used to transform Hunan geospatial data from WGS 84 to UTM zone 50N (i.e. EPSG: 32650).
:::
:::::

### 3.3 Import csv file into R environment

Next, we will import *Hunan_2012.csv* into R by using [`read_csv()`](https://readr.tidyverse.org/reference/read_delim.html) of **readr** package. The output is R data frame class.

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

### 3.4 Performing relational join

Below we will update the attribute table of *hunan*’s SpatialPolygonsDataFrame with the attribute fields of *hunan2012* dataframe. This is performed using [`left_join()`](https://dplyr.tidyverse.org/reference/mutate-joins.html) of **dplyr** package.

```{r}
hunan <- left_join(hunan, hunan2012) %>%
  dplyr::select(1:4, 7, 15)
```

### 3.5 Visualising Regional Development Indicator

Now, we are going to prepare a basemap and a choropleth map showing the distribution of GDPPC 2012 using `qtm()` of **tmap** package.

```{r}
#| fig-width: 12
#| fig-height: 8

equal <- tm_shape(hunan) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 5,
                values = "brewer.blues")) +
  tm_text("NAME_3", size = 0.5) +
  tm_borders(fiil_alpha = 0.5) + 
  tm_layout(legend.position = c("left", "bottom"), 
            main.title = "Equal interval classification")

quantile <- tm_shape(hunan) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5)) +
  tm_text("NAME_3", size = 0.5) +
  tm_borders(fiil_alpha = 0.5) + 
  tm_layout(legend.position = c("left", "bottom"), 
            main.title = "Quantile classification")

tmap_arrange(equal, 
             quantile, 
             asp=1, 
             ncol=2)
```

::::: markgoals
::: markgoals-header
**Insights**
:::

::: markgoals-container
-   **Does the plot above reveal any outliers or clusters?**

    -   **outliers** - from both maps, we can see that county **Lengshuijiang** appears to be dark blue and is surrounded by lighter colors counties. **Zixing**, could also be considered as an outlier as most of the surrounding counties are lighter colors. Another outlier is **Anren** that appears to be the bottom of GDPPC but surrounded by higher GDPPC counties.

    -   **clusters** - We see clusters appearing in the Hunan province. In the northeast and southeast there is a high GDPPC cluster each. In central south from Xinhua to Tongdao counties, these form a low GDPPC cluster.

-   **Does the plot above indicate the presence of hot spots or cold spots?**
Based on our observation for clusters: 

    -   Hot spots include the clusters surrounding **Changsha** and **Guiyang**.
    -   Cold spot is the cluster from **Xinhua** to **Tongdao** that we identified for low GDPPC cluster.
:::
:::::

## 4 Local Indicators of Spatial Association(LISA)

### 4.1 Computing Contiguity Spatial Weights

### 4.2 Row-standardised weights matrix

### 4.3 Computing local Moran’s I

### 4.4 Preparing and Visualising LISA Map

## 5 Hot Spots and Cold Spots Analysis (HCSA)

## 5.1 Deriving distance weight matrix

## 5.2 Computing local Gi\*

### 5.3 Preparing and Visualising HCSA Map

## 6 References
