---
title: "03 - Predictive Modeling for Singapore HDB Resale Prices"
author: "Cathy C."
date: "Nov 09, 2025"
date-modified: "last-modified" #auto render date
format: html
execute: #how YML page work
  echo: true #will show code
  eval: true #will run code every time and show output
  warning: false #dont show warning msg
  freeze: true #controls rendering - no render if no changes
---

## 1 Overview

### 1.1 Background

### 1.2 Objective

The objective of this take-home exercise is to develop a spatially informed model to predict HDB resale prices in Singapore, making use of appropriate geospatial analytics methods. We will calibrate a predictive model to predict HDB resale prices for the period July to September 2025, using HDB resale transaction data from July 2024 to June 2025.

### 1.3 The Task

In this study, we will first determine potential factors that influence the HDB resale price, import the data of these factors, and process them for predictive modeling. The modeling we will investigate in this task include Multi-linear Regression (MLR), Geographically Weighted Regression (GWR), non-spatial Random Forest, as well as Geographical Weighted Random Forest Model (gwRF). Among these models, we will examine and compare them to derive the best performing model.

## 2 Installing and Loading R packages

One new package called [**jsonlite**]{style="color: #F08787;"} will be installed and loaded onto R environment. It will be used to parse json file into R as data frame.

```{r}
pacman::p_load(httr, tidyverse, sf, tmap, jsonlite, progress,
               spdep, GWmodel, SpatialML, rsample, Metrics, knitr,
               kableExtra, spatialRF, randomForestExplainer, rvest,
               stringr, lubridate)
```

<package intro> progress - to monitor and track the computation.

## 3 The Data

We will make use of the following factors ...

-   Structural factors

    -   Area of the unit
    -   Floor level
    -   Remaining lease
    -   Age of the unit
    -   Main Upgrading Program (MUP) completed (optional)
    -   Average population age in the area??

-   Location factors

    -   Proxomity to CBD
    -   Proxomity to business park
    -   Proximity to eldercare
    -   Proximity to foodcourt/hawker centres
    -   Proximity to MRT
    -   Proximity to park
    -   Proximity to top primary school
    -   Proximity to shopping mall
    -   Proximity to supermarket
    -   Proximity to healthcare facilities
    -   Proximity to sports halls
    -   Proximity to community clubs   
    -   Numbers of kindergartens within 350m
    -   Numbers of childcare centres within 350m
    -   Numbers of bus stop within 350m
    -   Numbers of primary school within 1km

+--------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------+
| Data set                       | Link                                                                                                                                                                            | Source        |
+================================+=================================================================================================================================================================================+===============+
| CBD                            |                                                                                                                                                                                 |               |
+--------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------+
| Business Parks                 | [JTC Business Park Land](https://data.gov.sg/datasets?query=park&page=1&resultId=d_7555d52fb87e75da7397bd65d4f107dd)                                                            |               |
+--------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------+
| eldercare                      | [Eldercare Services](https://data.gov.sg/datasets?query=eldercare&page=1&resultId=d_f0fd1b3643ed8bd34bd403dedd7c1533)                                                           |               |
+--------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------+
| foodcourt/hawker centres       | [**Hawker Centres (GEOJSON)**](https://data.gov.sg/datasets?query=hawker+centre&page=1&resultId=d_4a086da0a5553be1d89383cd90d07ecd)                                             | data.gov.sg   |
+--------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------+
| MRT                            | [TrainStation_Aug2025](https://datamall.lta.gov.sg/content/datamall/en/static-data.html)                                                                                        | DataMall      |
+--------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------+
| Major interchange stations     | Derive from MRT based on [LTAguru](https://landtransportguru.net/bus-interchanges-and-terminals/)                                                                               |               |
+--------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------+
| Upcoming bus interchange       | Derive from MRT based on [LTAguru](https://landtransportguru.net/bus-interchanges-and-terminals/)                                                                               |               |
+--------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------+
| park                           | [Park Facilities](https://data.gov.sg/datasets?query=park&page=1&resultId=d_14d807e20158338fd578c2913953516e)                                                                   |               |
+--------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------+
| shopping mall                  | wiki + Kaggle                                                                                                                                                                   | wiki + Kaggle |
+--------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------+
| supermarket                    | Supermarkets (GEOJSON)                                                                                                                                                          | data.gov.sg   |
+--------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------+
| kindergartens within 350m      | [Kindergartens](https://data.gov.sg/datasets?query=kindergarten&resultId=d_7fe9a72b1afff18e48111772c8d0fd39&page=1)                                                             |               |
+--------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------+
| childcare centres within 350m  | [Pre-Schools Location](https://data.gov.sg/datasets?query=preschool&resultId=d_61eefab99958fd70e6aab17320a71f1c&page=1)                                                         |               |
|                                |                                                                                                                                                                                 |               |
|                                | [Listing of Centres](https://data.gov.sg/datasets?query=kindergarten&resultId=d_696c994c50745b079b3684f0e90ffc53&page=1)                                                        |               |
+--------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------+
| bus stop within 350m           |                                                                                                                                                                                 | DataMall      |
+--------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------+
| primary school within 1km      | [School Directory and Information](https://data.gov.sg/datasets?query=school+information&resultId=457&page=1)                                                                   |               |
+--------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------+
| good primary school within 1km | Primary school with data from [sophia education](https://sophiaeducation.sg/2025-singapore-psle-school-rankings/#%F0%9F%8F%AB_Top_20_Primary_Schools_in_Singapore_2025_Edition) |               |
+--------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------+
| healthcare facilities          | [CHAS Clinics](https://data.gov.sg/datasets?query=clinic&resultId=d_548c33ea2d99e29ec63a7cc9edcccedc&page=1)                                                                    |               |
+--------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------+
| Sports hall                    | [SportsFields\@SG](https://data.gov.sg/datasets?query=sports+hub&page=1&resultId=d_f71b449b4b43a69b5ecfe411b440d249)                                                            | data.gov.sg   |
+--------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------+

### 2.1 Import Data

#### 2.1.0 Singapore Maps

Below we will import *Master Plan 2019 Subzone Boundary (No Sea).*

```{r}
#| eval: false
#| code-fold: true
mpsz <- st_read("data/rawdata/MasterPlan2019SubzoneBoundaryNoSeaKML.kml") %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 3414)

# extract values from the HTML description
extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)
  rows <- page %>% html_elements("tr")
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>%
    html_element("td") %>%
    html_text2()
  
  if (length(value) == 0) NA_character_ else value
}

# extract info
mpsz <- mpsz %>%
  mutate(
    REGION_N = map_chr(Description, extract_kml_field, "REGION_N"),
    PLN_AREA_N = map_chr(Description, extract_kml_field, "PLN_AREA_N"),
    SUBZONE_N = map_chr(Description, extract_kml_field, "SUBZONE_N"),
    SUBZONE_C = map_chr(Description, extract_kml_field, "SUBZONE_C")
  ) %>%
  select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())

write_rds(mpsz, "data/rds/mpsz.rds")
```


```{r}
#| echo: false
mpsz <- read_rds("data/rds/mpsz.rds")
st_crs(mpsz)
```


#### 2.1.1 HDB resale data Preparation
-   **Load HDB Resale Data**
This data set comprises of HDB resale transactions from July 2017 onwards. Below we will extract a sub-set of transaction records on 4-Room flat transacted during July 2024 to June 2025.

```{r}
#| eval: false
#| code-fold: true

HDBresale <- read_csv(
  "data/rawdata/ResaleflatpricesJan2017.csv") %>%
  filter(flat_type == "4 ROOM") %>%
  filter(month >= "2024-07" & month < "2025-06") # filter the sample data for the exercise.

write_rds(HDBresale, "data/rds/HDBresale.rds")
```

-   **Further processing for street names**
We do further processing for the data as the street names with Saint are usually written as St.

``` {r}
#| eval: false
#| code-fold: true
# street names is ST, but it's actually Sait.
HDBresale$street_name <- gsub(
  "ST\\.", "SAINT",
  HDBresale$street_name
)
```

```{r}
#| echo: false
HDBresale <- read_rds("data/rds/HDBresale.rds")
glimpse(HDBresale)
```

-   **Geocoding HDB Resale Data**
We use OneMap Geocode API: - <https://www.onemap.gov.sg/apidocs/.> - [ttps://www.onemap.gov.sg/apidocs/reverseGeocode](ttps://www.onemap.gov.sg/apidocs/reverseGeocode)

Below we will prepare a geocoding function. It aims to perform the following tasks:

-   Combining the block and street name into an address. The addresses will be used to geocode instead of postal code
-   Passing the address as the **searchVal** in our query
-   Sending the query to **OneMapSG search**. Note: Since we don’t need all the address details, we can set getAddrDetails as ‘N’
-   Converting response (JSON object) to text
-   Saving response in text form as a dataframe
-   Retain the latitude and longitude for our output.

``` {r}
#| eval: false
# for multiple queries
geocode <- function(block, streetname) {
  base_url <- "https://onemap.gov.sg/api/common/elastic/search"
  address <- paste(block, streetname, sep = " ")
  query <- list("searchVal" = address,
                "returnGeom" = "Y",
                "getAddrDetails" = "N", #No. can change to Y for comparison
                "pageNum" = "1") 
  res <- GET(base_url, query = query) #get URL query
  restext <- content(res, as = "text")
  
  output <- fromJSON(restext) %>%
    as.data.frame %>%
    select(results.LATITUDE, results.LONGITUDE)
  
  return(output)
}
```

-   **Run functions**

``` {r}
#| eval: false
HDBresale$LATITUDE <- 0
HDBresale$LONGITUDE <- 0

for(i in 1:nrow(HDBresale)){
  temp_output <- geocode(HDBresale[i, 4], HDBresale[i, 5])
  
  HDBresale$LATITUDE[i] <- temp_output$results.LATITUDE
  HDBresale$LONGITUDE[i] <- temp_output$results.LONGITUDE
}
```

-   **Transform to sf data frame**
We want to ensure that:
-   both input data sets must be in sf objects
-   they must be in similar projected coordinates systems

```{r}
#| eval: false
#| code-fold: true

st_crs(HDBresale)
HDBresale_sf <- HDBresale %>%
  st_as_sf(
    coords = c("LONGITUDE", "LATITUDE"),
    crs = 4326) %>% #if decimal degree
  st_transform(crs = 3414)

write_rds(HDBresale_sf, "data/rds/HDBresale_sf.rds")
```
```{r}
#| echo: false
HDBresale_sf <- read_rds("data/rds/HDBresale_sf.rds")
```


#### 2.1.2 Load location factors data

::: panel-tabset

## Education

-   **Childcare centers**

```{r}
#| eval: false
#| code-fold: true
preschool_data <- st_read("data/rawdata/PreschoolsLocation.kml") %>%
  st_transform(crs = 3414)
write_rds(preschool_data, "data/rds/preschool_data.rds")
```

```{r}
#| echo: false
preschool_data <- read_rds("data/rds/preschool_data.rds")
```


Make sure the childcare coordinate system is SVY21.
```{r}
#| code-fold: true
st_crs(preschool_data)
```

-   **Kindergartens**
```{r}
#| eval: false
#| code-fold: true
kindergarten_data <- st_read("data/rawdata/Kindergartens.geojson") %>%
  st_transform(crs = 3414)

write_rds(kindergarten_data, "data/rds/kindergarten_data.rds")
```
```{r}
#| echo: false
#| code-fold: true
kindergarten_data <- read_rds("data/rds/kindergarten_data.rds")
```


-   **Primary schools data**

```{r}
#| code-fold: true
#| eval: false
primschool_data <- read_csv("data/rawdata/General Information of schools.csv") %>%
  select(school_name, postal_code, address)
```

Below we need to geocode the primary school location because currently we have their postal codes.
```{r}
#| code-fold: true
#| eval: false
geocode_postal <- function(postal_code) {
  url <- paste0("https://www.onemap.gov.sg/api/common/elastic/search?",
                "searchVal=", postal_code,
                "&returnGeom=Y&getAddrDetails=Y")
  
  response <- GET(url)
  
  if (status_code(response) == 200) {
    result <- content(response, "text") %>% fromJSON()
    
    if (result$found > 0) {
      return(data.frame(
        latitude = as.numeric(result$results$LATITUDE[1]),
        longitude = as.numeric(result$results$LONGITUDE[1])
      ))
    }
  }
  
  return(data.frame(latitude = NA, longitude = NA))
}

# Geocode all primary schools
primschool_coords <- lapply(primschool_data$postal_code, function(pc) {
  Sys.sleep(0.2)  
  geocode_postal(pc)
})

primschool_coords_df <- bind_rows(primschool_coords)

primschool_data <- primschool_data %>%
  bind_cols(primschool_coords_df)

# Convert to sf object
primschool_data <- primschool_data %>%
  filter(!is.na(latitude)) %>%  # Remove any that failed
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(3414)  # Transform to SVY21

nrow(primschool_data)  # How many successfully geocoded
```

Check what schools we will filter out using keywords:
```{r}
#| code-fold: true
#| eval: false
primschool_data %>%
  filter(
    grepl("SECONDARY|HIGH|INSTITUTE|INSTITUTION|SCHOOL OF|JUNIOR|COLLEGE", 
               school_name, 
               ignore.case = TRUE)) %>%
  select(school_name) %>%
  print(n = 50)
```



```{r}
#| code-fold: true
#| eval: false
# Remove non-primary schools 
primschool_data <- primschool_data %>%
  filter(school_name %in% c("MARIS STELLA HIGH SCHOOL", "CATHOLIC HIGH SCHOOL")  |
    !grepl("SECONDARY|HIGH|INSTITUTE|INSTITUTION|SCHOOL OF|JUNIOR|COLLEGE", 
                school_name, 
                ignore.case = TRUE)) %>%
  filter(!school_name %in% c("CHIJ KATONG CONVENT", "CHIJ ST. JOSEPH'S CONVENT"))

nrow(primschool_data)
write_rds(primschool_data, "data/rds/primschool_data.rds") 
```

```{r}
#| echo: false
primschool_data <- read_rds("data/rds/primschool_data.rds")
```


-   **Top Primary schools**

```{r}
#| code-fold: true
#| eval: false
top_primschool_data <- primschool_data %>%
  filter(school_name %in% c("RAFFLES GIRLS' PRIMARY SCHOOL", "NANYANG PRIMARY SCHOOL",
                          "HENRY PARK PRIMARY SCHOOL", "ROSYTH SCHOOL",
                          "AI TONG SCHOOL", "NAN HUA PRIMARY SCHOOL", "ST. HILDA'S PRIMARY SCHOOL",
                          "TAO NAN SCHOOL", "CHIJ ST. NICHOLAS GIRLS' SCHOOL",
                          "CATHOLIC HIGH SCHOOL", "PEI HWA PRESBYTERIAN PRIMARY SCHOOL",
                          "METHODIST GIRLS' SCHOOL (PRIMARY)", "KONG HWA SCHOOL",
                          "RED SWASTIKA SCHOOL", "SINGAPORE CHINESE GIRLS' SCHOOL", "MARIS STELLA HIGH SCHOOL",
                          "ANGLO-CHINESE SCHOOL (PRIMARY)", "TEMASEK PRIMARY SCHOOL","RIVER VALLEY PRIMARY SCHOOL",
                          "CEDAR PRIMARY SCHOOL"))

write_rds(top_primschool_data, "data/rds/top_primschool_data.rds")
```

```{r}
#| echo: false

top_primschool_data <- read_rds("data/rds/top_primschool_data.rds")
```

## Business

-   **CBD**

For CBD, we will extract the area for CBD from `mpsz`.
```{r}
#| code-fold: true
#| eval: false
cbd_data <- mpsz %>%
  filter(SUBZONE_N %in% c("RAFFLES PLACE", "MARINA CENTRE", "TANJONG PAGAR", "CITY HALL", "CENTRAL SUBZONE")) %>%
  st_union()

write_rds(cbd_data, "data/rds/cbd_data.rds")
```

```{r}
#| echo: false
cbd_data <- read_rds("data/rds/cbd_data.rds")
```

```{r}
glimpse(cbd_data)
```
cbd_data is now a single merged polygon representing the entire CBD area. We will use this to calculate distance in later section.

-   **Business park**

Business parks are typically smaller than planning areas, and we have multiple business parks to consider. Therefore, we will calculate centroid for measuring proximity, and transform polygon data to centroid point data.


```{r}
#| code-fold: true
#| eval: false

bizpark_data <- st_read("data/rawdata/JTCBusinessParkLandGEOJSON.geojson") %>% 
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 3414) %>%
  st_centroid()

write_rds(bizpark_data, "data/rds/bizpark_data.rds")
```

```{r}
#| echo: false
bizpark_data <- read_rds("data/rds/bizpark_data.rds")
```


## Recreation

-   **Shopping malls**

```{r}
#| code-fold: true
#| eval: false

sgmalls_rawdata <- read_csv("data/rawdata/singapore_malls.csv") %>%
  filter(!Mall_Name %in% c('JCube', 'JCube (closed)', 'Liang Court', 'Lot 1', 
                           'Nex', 'Novena Square Shopping Mall'))

sgmalls_coordinates <- read_csv("data/rawdata/shopping_mall_coordinates.csv") %>%
  rename(Mall_Name = 'Mall Name')

merged_malls <- full_join(sgmalls_rawdata, sgmalls_coordinates, by = "Mall_Name")

```
After the merge, we examine if any shopping malls turn out without coordinates data. 
```{r}
#| code-fold: true
#| eval: false
malls_without_coords <- merged_malls %>%
  filter(is.na(LONGITUDE) | is.na(LATITUDE)) %>%
  pull(Mall_Name)

malls_without_coords
```

Amber Mansions - demolished in 1984
Bedok Point
Capitol Centre
Ellenborough Market
Funan DigitaLife Mall - duplicate
Great World - duplicate
GV Yishun - cinema
Jurong Entertainment Centre - demolished 2010
Shaw Tower - demolished
Specialists' Shopping Centre - redeveloped as Orchard Gateway
Tanglin Shopping Centre - CLOSED
The Golden Mile - CLOSED
The Verge - closed
Woodland Mall - change to Woods square mall

For this case, we manually check with Google maps
```{r}
#| code-fold: true
#| eval: false

coords_to_fill <- tibble(
  Mall_Name = c("Alexandra Retail Centre", "Capitol Singapore", "Change Alley",
                "Chinatown Point", "Clementi Mall", "Dawson Place", 
                "Esplanade Mall", "Forum The Shopping Mall", "Fu Lu Shou Complex", 
                "HarbourFront Centre", "Holland Road Shopping Centre", 
                "Holland Village Shopping Mall", 
                "Mandarin Gallery", "Mustafa Centre", "Novena Square", 
                "Orchard Towers", "Paya Lebar Quarter", "Peninsula Plaza",
                "Punggol Coast Mall", "Scotts Shopping Centre", "Seletar Mall",
                "Shaw House and Centre", "Singpost Center",
                "South Beach", "Suntec City Mall",
                "Tang Plaza", "The Cathay", 
                "The Majestic", "The Rail Mall", 
                "VivoCity", "Woodland Mall"),
  
  LONGITUDE = c(103.8015, 103.8511, 103.8518, 
                103.8446, 103.7642, 103.8107,
                103.8561, 103.8284, 103.8543,
                103.8201, 103.7952, 
                103.7937, 
                103.8363, 103.8554, 103.8439,
                103.8292, 103.8930, 103.8508,
                103.9104, 103.8329, 103.8758,
                103.8318, 103.8940, 
                103.8557, 103.8581, 
                103.8329, 103.8477,
                103.8432, 103.7678,
                103.8231, 103.7859
                ),
  LATITUDE = c(1.2741176, 1.293231, 1.284643, 
               1.285434, 1.315111, 1.293505, 
               1.290017, 1.306391, 1.301906,
               1.265355, 1.310419, 1.319192,
               
               1.311708, 
               1.317859, 1.292544, 1.302237, 
               1.310192, 1.320230, 1.307743, 
               1.415818, 1.306093, 1.392464,
               1.306507, 
               1.294485, 1.295778, 
               1.305221, 1.300409, 
               1.284871, 1.358212, 
               1.265357, 1.438082
               )
) 

merged_malls <- merged_malls %>%
  rows_update(coords_to_fill, by = "Mall_Name", unmatched = "ignore")

merged_malls <-  merged_malls %>%
  mutate(
    Mall_Name = case_when(
      Mall_Name == "Woods Square Mall" ~ "Woodland Mall",
      TRUE ~ Mall_Name
    )
  )

# add new mall data
new_malls <- tibble(
  Mall_Name = "Woodleigh Mall",
  LONGITUDE = 103.8718,
  LATITUDE = 1.339560
)  

merged_malls <- merged_malls %>%
  bind_rows(new_malls)
```

Check the malls with missing coordinates, meaning that we did not find the coordinates for these mall as they're closed.
```{r}
#| code-fold: true
#| eval: false
malls_with_na <- merged_malls %>%
  filter(is.na(LONGITUDE) | is.na(LATITUDE)) 

malls_with_na
```
These malls are the ones duplicate or closed. We will go ahead to remove them. We will also remove 2 duplicate malls observed from *merged_malls*. 
```{r}
#| code-fold: true
#| eval: false
merged_malls <- merged_malls %>%
  filter(!is.na(LONGITUDE) & !is.na(LATITUDE)) %>%
  distinct(Mall_Name, .keep_all = TRUE) %>%
  filter(!Mall_Name %in% c("Suntec City Mall", "Capitol Piazza"))

# Check how many remain
nrow(merged_malls)

# merged_malls has long, lat coordinates. We want to convert it to sf object, then SVY21.
merged_malls <- merged_malls %>%
  st_as_sf(coords = c("LONGITUDE", "LATITUDE"), crs = 4326) %>%  # it's WGS84
  st_transform(3414)  

write_rds(merged_malls, "data/rds/malls_data.rds")
```

```{r}
#| echo: false

malls_data <- read_rds("data/rds/malls_data.rds")
```


-   **Sports halls**

```{r}
#| code-fold: true
#| eval: false
sports_data <- st_read("data/rawdata/SportsFieldsSG.geojson") %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 3414)

extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)
  rows <- page %>% html_elements("tr")
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>%
    html_element("td") %>%
    html_text2()
  
  if (length(value) == 0) NA_character_ else value
}

# extract info
sports_data <- sports_data %>%
  mutate(
    name = map_chr(Description, extract_kml_field, "NAME"),
    description = map_chr(Description, extract_kml_field, "DESCRIPTION")
  ) %>%
  select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())

write_rds(sports_data, "data/rds/sports_data.rds")
```

```{r}
#| echo: false
sports_data <- read_rds("data/rds/sports_data.rds")

```

-   **Parks**

```{r}
#| code-fold: true
#| eval: false
park_data <- st_read("data/rawdata/MasterPlan2019SDCPParkandOpenSpacelayer.kml") %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_make_valid() %>%
  st_transform(crs = 3414)

# Calculate centroids 
park_data <- park_data %>%
  mutate(centroid = st_centroid(geometry))

write_rds(park_data, "data/rds/park_data.rds")
```
```{r}
#| echo: false
park_data <- read_rds("data/rds/park_data.rds")
```

-   **Supermarkets**


```{r}
#| code-fold: true
#| eval: false

supermarkets_data <- st_read("data/rawdata/SupermarketsGEOJSON.geojson") %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 3414)
# # calculate centroids
#   mutate(centroid = st_centroid(geometry)) %>%
#   select(OBJECTID_1, centroid)

extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)
  rows <- page %>% html_elements("tr")
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>%
    html_element("td") %>%
    html_text2()
  
  if (length(value) == 0) NA_character_ else value
}

# extract info
supermarkets_data <- supermarkets_data %>%
  mutate(
    name = map_chr(Description, extract_kml_field, "LIC_NAME"),
    postcode = map_chr(Description, extract_kml_field, "POSTCODE")
  ) %>%
  select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())  

write_rds(supermarkets_data, "data/rds/supermarkets_data.rds")  
```
```{r}
#| echo: false
supermarkets_data <- read_rds("data/rds/supermarkets_data.rds")
```

-   **Community clubs**
```{r}
#| code-fold: true
#| eval: false
cc_data <- st_read("data/rawdata/CommunityClubs.geojson") %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 3414)

extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)
  rows <- page %>% html_elements("tr")
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>%
    html_element("td") %>%
    html_text2()
  
  if (length(value) == 0) NA_character_ else value
}

cc_data <- cc_data %>%
  mutate(
    name = map_chr(Description, extract_kml_field, "NAME")
  ) %>%
  select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())

write_rds(cc_data, "data/rds/cc_data.rds")
```

```{r}
#| echo: false
cc_data <- read_rds("data/rds/cc_data.rds")
```

-   **Foodcourt/Hawker centers**
```{r}
#| code-fold: true
#| eval: false
hkcenter_data <- st_read("data/rawdata/HawkerCentresGEOJSON.geojson") %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 3414)

write_rds(hkcenter_data, "data/rds/hkcenter_data.rds")

```

```{r}
#| echo: false
hkcenter_data <- read_rds("data/rds/hkcenter_data.rds")

```
Data is also transformed to SVY21. 

## Transportation

-   **MRT**

Below we will import MRT station exit shapefile. The MRT stations may have multiple exits, and these exits matter when people go through them to enter or exit the stations. We will use the data to compute for proximity to HDB units.

```{r}
#| code-fold: true
mrt_data <- st_read("data/rawdata/TrainStationExit_Feb2025/Train_Station_Exit_Layer.shp") %>%
  st_transform(crs = 3414)

```

-   **Bus stops**

Below we will import *Bus Stop Location* shapfile from LTA DataMall.

```{r}
#| code-fold: true
#| eval: false
bus_data = st_read(dsn = "data/rawdata/BusStopLocation_Aug2025",
                  layer = "BusStop")
bus_data <- bus_data %>%
  st_transform(crs = 3414)

write_rds(bus_data, "data/rds/bus_data.rds")
```

```{r}
#| echo: false
bus_data <- read_rds("data/rds/bus_data.rds")

```

## Healthcare

-   **Clinics**


```{r}
#| code-fold: true
#| eval: false

clinics_data <- st_read("data/rawdata/CHASClinics.geojson") %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 3414)

extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)
  rows <- page %>% html_elements("tr")
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>%
    html_element("td") %>%
    html_text2()
  
  if (length(value) == 0) NA_character_ else value
}

# extract info
clinics_data <- clinics_data %>%
  mutate(
    name = map_chr(Description, extract_kml_field, "HCI_NAME")
  ) %>%
  select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())

write_rds(clinics_data, "data/rds/clinics_data.rds")

```

```{r}
#| echo: false
clinics_data <- read_rds("data/rds/clinics_data.rds")

```

Above data is verified to conform **SVY21**. 

-   **Eldercare**

```{r}
#| code-fold: true
#| eval: false

eldercare_data <- st_read("data/rawdata/EldercareServices.geojson") %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 3414)

extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)
  rows <- page %>% html_elements("tr")
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>%
    html_element("td") %>%
    html_text2()
  
  if (length(value) == 0) NA_character_ else value
}

# extract info
eldercare_data <- eldercare_data %>%
  mutate(
    name = map_chr(Description, extract_kml_field, "NAME")
  ) %>%
  select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())

write_rds(eldercare_data, "data/rds/eldercare_data.rds")

```


```{r}
#| echo: false
eldercare_data <- read_rds("data/rds/eldercare_data.rds")
```
Above data is verified to conform **SVY21**. 

:::

#### 2.1.3 Initial data inspection
We validate the coordinates of the location factor data below. It shows us that all now conform to SVY21, and ready for proximity computation.

```{r}
cat('bizpark_data: ')
st_crs(bizpark_data)$epsg
cat('bd_data: ')
st_crs(cbd_data)$epsg
cat('clinics_data: ')
st_crs(clinics_data)$epsg
cat('eldercare_data: ')
st_crs(eldercare_data)$epsg
cat('hkcenter_data: ')
st_crs(hkcenter_data)$epsg
cat('malls_data: ')
st_crs(malls_data)$epsg
cat('mrt_data: ')
st_crs(mrt_data)$epsg
cat('park_data: ')
st_crs(park_data)$epsg
cat('preschool_data: ')
st_crs(preschool_data)$epsg
cat('primschool_data: ')
st_crs(primschool_data)$epsg
cat('supermarkets_data: ')
st_crs(supermarkets_data)$epsg
cat('sports_data: ')
st_crs(sports_data)$epsg
cat('bus_data: ')
st_crs(bus_data)$epsg
cat('kindergarten_data: ')
st_crs(kindergarten_data)$epsg
cat('cc_data: ')
st_crs(cc_data)$epsg


```


### 2.2 Structural factors
The tructural factors for this study include flat type, floor level, town, planning area, area of unit, age of unit, main upgrading program, and flat model. 
Currently in our data, we have the factors shown below:
```{r}
names(HDBresale_sf)
```

We will rename some of the factors and keep the factors being used for the study:
```{r}
#| code-fold: true
#| eval: false
HDBresale_data <- HDBresale_sf %>%
  select(flat_type, month, storey_range, town, floor_area_sqm, flat_model, 
         remaining_lease, resale_price, geometry) %>%
  rename(transact_date = month,
         remaining_lease_mths = remaining_lease,
         floor_level = storey_range,
         plan_area = town) 
write_rds(HDBresale_data, "data/rds/HDBresale_data.rds")  
```

```{r}
#| echo: false
HDBresale_data <- read_rds("data/rds/HDBresale_data.rds")
glimpse(HDBresale_data)
```
For predictive modeling, we should change data type for the following variables: `transact_date` and `remaining_lease_mths` to date and numeric data type respectively.

```{r}
#| code-fold: true
#| eval: false

# convert transction date to date data type
HDB_data <- HDBresale_data %>%
  mutate(
    transact_date = as.Date(paste0(transact_date, "-01")),
    
    # Extract years
    years = as.numeric(str_extract(remaining_lease_mths, "\\d+(?= year)")),
    # Extract months
    months = as.numeric(str_extract(remaining_lease_mths, "\\d+(?= month)")),

    years = ifelse(is.na(years), 0, years),
    months = ifelse(is.na(months), 0, months),

    remaining_lease_mths_num = years * 12 + months
  ) %>%
  select(-years, -months)  # Remove temporary columns
write_rds(HDB_data, "data/rds/HDB_data.rds")
```

Below we confirm the transction date is now `Date` data type and the remaining lease is in `numeric` data type as months.
```{r}
#| echo: false
HDB_data <- read_rds("data/rds/HDB_data.rds")

head(HDB_data$transact_date, 10)
class(HDB_data$transact_date)
class(HDB_data$remaining_lease_mths_num)
```


### 2.3 Proximity Analysis


In this section, we will prepare location factors of their proximity to the HDB units and the availability of these facilities within a fixed distance of the HDB housing. 

#### 2.3.1 Computing shortest distance

Instead of counting number of preschools located within 350m of each HDB resale property, code chunk below is used to compute the distance between the nearest preschool from each HDB resale property. 

`st_is_within_distance()` of [**sf**]{style="color: #F08787;"} package is used to determine whether two spatial objects (HDB property and preschool) are within a set distance of each other. It returns a logical matrix (or a sparse list of indices if sparse = TRUE) indicating for each geometry in HDB property which geometries in the facility are within the set distance.

check for outliers / errors Verify coordinates are in Singapore (plot tmaps)

::: panel-tabset

## Education
```{r}
#| code-fold: true
#| eval: false

# Top Primary school
HDB_combine_edu <- HDB_data %>%
  mutate(
    PROX_TOPSCHOOL = as.numeric(
      st_distance(geometry,
                  top_primschool_data[st_nearest_feature(
                    geometry, top_primschool_data), ], 
                  by_element = TRUE)
      )
    )
```

## F&B
```{r}
#| code-fold: true
#| eval: false

# Hawker centers
HDB_combine_2fb <- HDB_combine_edu %>%
  mutate(
    PROX_HKCENTER = as.numeric(
      st_distance(geometry,
                  hkcenter_data[st_nearest_feature(
                    geometry, hkcenter_data), ], 
                  by_element = TRUE)),
# Supermarkets
    PROX_SUPERMKT =  as.numeric(
      st_distance(geometry,
                  supermarkets_data[st_nearest_feature(
                    geometry, supermarkets_data), ], 
                  by_element = TRUE))
    )

```

## Healthcare
```{r}
#| code-fold: true
#| eval: false
# Clinics
HDB_combine_3hc <- HDB_combine_2fb %>%
  mutate(
    PROX_CLINIC = as.numeric(
      st_distance(geometry,
                  clinics_data[st_nearest_feature(
                    geometry, clinics_data), ], 
                  by_element = TRUE)),
# Eldercare
    PROX_ELDERCARE =  as.numeric(
      st_distance(geometry,
                  eldercare_data[st_nearest_feature(
                    geometry, eldercare_data), ], 
                  by_element = TRUE))
    )

```

## Recreation
```{r}
#| code-fold: true
#| eval: false
# Shopping mall
HDB_combine_4rc <- HDB_combine_3hc %>%
  mutate(
    PROX_MALLS = as.numeric(
      st_distance(geometry,
                  malls_data[st_nearest_feature(
                    geometry, malls_data), ], 
                  by_element = TRUE)),
# Sports hall
    PROX_SPORTS =  as.numeric(
      st_distance(geometry,
                  sports_data[st_nearest_feature(
                    geometry, sports_data), ], 
                  by_element = TRUE)),
# Community Club
    PROX_CC =  as.numeric(
      st_distance(geometry,
                  cc_data[st_nearest_feature(
                    geometry, cc_data), ], 
                  by_element = TRUE)),
# Parks
    PROX_PARK =  as.numeric(
      st_distance(geometry,
                  park_data[st_nearest_feature(
                    geometry, park_data), ], 
                  by_element = TRUE))
    )

```

## Transportation

``` {r}
#| code-fold: true
#| eval: false
# Shopping mall
HDB_combine_5transp <- HDB_combine_4rc %>%
  mutate(
    PROX_MRT = as.numeric(
      st_distance(geometry,
                  mrt_data[st_nearest_feature(
                    geometry, mrt_data), ], 
                  by_element = TRUE))
    )

```

## Business
```{r}
#| code-fold: true
#| eval: false
# CBD
HDB_combine_6biz <- HDB_combine_5transp %>%
  mutate(
    PROX_CBD = as.numeric(
      st_distance(geometry,
                  cbd_data)),
    PROX_BIZPK = as.numeric(
      st_distance(geometry,
                  cbd_data)),
# Business parks
    PROX_BIZPK = as.numeric(
      st_distance(geometry,
                  bizpark_data[st_nearest_feature(
                    geometry, bizpark_data), ], 
                  by_element = TRUE))
    )
```


:::



-   `map_int()` of [**purr**]{style="color: #F08787;"} package is used to the number of preschools found in the newly created WITHIN_350M_PRISCHOOL field.



#### **2.3.2 Create binary proximity indicators**

Next we will derive a binary data to find out which schools are within 350 meters.

::: panel-tabset

## Top primary school

Within 1 km
``` {r}
#| code-fold: true
#| eval: false
within_1km <- st_is_within_distance(
  HDB_data, top_primschool_data, dist = 1000)

HDB_combine_7bi_primschool <- HDB_combine_6biz %>%
  mutate(
    WITHIN_1km_TOPPRISCH = map_int(within_1km, length),
    WITHIN_1km_TOPPRISCH_BINARY = ifelse(WITHIN_1km_TOPPRISCH > 0, 1, 0)
)
```

## MRT 
Within 500 meters
```{r}
#| code-fold: true
#| eval: false
within_500m_mrt <- st_is_within_distance(
  HDB_data, mrt_data, dist = 500)

HDB_combine_7bi_mrt <- HDB_combine_7bi_primschool %>%
  mutate(
    WITHIN_500m_MRT_BINARY = as.numeric(lengths(within_500m_mrt) > 0)
)
```


## Hawker centre 
Within 500 meters

```{r}
#| code-fold: true
#| eval: false
within_500m_hkcenter <- st_is_within_distance(
  HDB_data, hkcenter_data, dist = 500)

HDB_combine_7bi_hkcenter <- HDB_combine_7bi_mrt %>%
  mutate(
    WITHIN_500m_HKC_BINARY = as.numeric(lengths(within_500m_hkcenter) > 0)
)
```
:::

### 2.4 Amenity Density Analysis

Computing number of location factors

::: panel-tabset
## kindergarten

within 350m

## childcare

within 350m

## bus stop

within 350m

## primary school

within 1km
:::

### 2.5 Final Dataset Assembly

#### 2.5.1 Merge all features into master dataset

#### 2.5.2 Handle remaining missing values

#### 2.5.3 Save processed dataset

## 3 Data Quality & Spatial Overview

### 3.1 Data Quality Check

-   Summary statistics
-   Check for outliers for price
-   Coordinate validation

### 3.2 Spatial Visualization of Amenities

#### 3.2.1 Target Variable Analysis

HDB resale price distribution

#### 3.2.2 Interactive map of all amenities

::: panel-tabset
## HDB Units

-   All resale units (color by price)

## Education

-   Primary schools (top 20 highlighted)
-   Kindergartens
-   Childcare centres

## Transport

-   MRT stations (current + future)
-   Major bus interchanges

## Commercial

-   Shopping malls
-   Hawker centres
-   Supermarkets

## Business

-   CBD
-   BUsiness parks

## Recreation & Others

-   Parks
-   Sports facilities
-   Healthcare
:::

## 4 Pre-modeling Preparation

### 4.1 Multicollinearity Check

-   Correlation matrix of predictors
-   VIF scores
-   Identify multicollinearity issues Remove highly correlated features

### 4.2 Feature Selection

-   Final feature list for modeling

### 4.3 Train/Test Split

#### 4.3.1 Stratified Spatial Sampling

``` r
library(dplyr)
library(sf)

set.seed(123)

# Stratify by town AND year (if multi-year)
hdb_sample <- hdb_data %>%
  # Ensure temporal coverage
  group_by(year, town) %>%
  # Sample proportionally
  slice_sample(prop = 5000/nrow(hdb_data)) %>%
  ungroup()

# If you get slightly more/less than 5000, adjust:
if(nrow(hdb_sample) > 5000) {
  hdb_sample <- hdb_sample %>% sample_n(5000)
} else if(nrow(hdb_sample) < 5000) {
  # Add more randomly to reach 5000
  additional <- hdb_data %>%
    anti_join(hdb_sample) %>%
    sample_n(5000 - nrow(hdb_sample))
  hdb_sample <- bind_rows(hdb_sample, additional)
}

# Verify distribution
hdb_sample %>% count(town)
hdb_sample %>% count(flat_type)

# Save
write_csv(hdb_sample, "data/processed/hdb_sample_5000.csv")
```

``` r
cat("Original dataset:", nrow(hdb_data), "transactions
")
cat("Sample size:", nrow(hdb_sample), "transactions
")
cat("Sampling rate:", round(5000/nrow(hdb_data)*100, 2), "%
")
```

-   Total original dataset size
-   Sampling method used
-   Why 5000 was chosen
-   How representativeness was ensured

``` r
# Compare distributions
par(mfrow = c(2,2))

# Price distribution
plot(density(hdb_data$resale_price), main = "Full Dataset", col = "blue")
plot(density(hdb_sample$resale_price), main = "Sample", col = "red")

# Town distribution
barplot(table(hdb_data$town), main = "Full Dataset - Towns")
barplot(table(hdb_sample$town), main = "Sample - Towns")
```

#### 4.3.2 Tran/Test split

``` r
# 80/20 split
set.seed(456)
train_idx <- sample(1:nrow(hdb_sample), 0.8 * nrow(hdb_sample))

train <- hdb_sample[train_idx, ]   # 4000 records
test <- hdb_sample[-train_idx, ]   # 1000 records

cat("Train size:", nrow(train), "
")
cat("Test size:", nrow(test), "
")
```

## 5 Multiple Linear Regression (MLR)

### 5.1 Model specification

### 5.2 Model fitting

### 5.3 Results & interpretation

### 5.4 Residual diagnostics

## 6 Geographically Weighted Regression (GWR)

### 6.1 Bandwidth selection

### 6.2 Model fitting

### 6.3 Local coefficient maps

### 6.4 Results & interpretation

## 7 Non-sptial Random Forest Model

### 7.1 Feature selection

### 7.2 Hyperparameter tuning

### 7.3 Model fitting

### 7.4 Feature importance

### 7.5 Results & interpretation

## 8 Geographical Weighted Random Forest Model (gwRF)

### 8.1 Model specification

### 8.2 Bandwidth selection

### 8.3 Model fitting

### 8.4 Local predictions & interpretation

## 9 Model Evaluation

### 9.1 Performance metrics (RMSE, MAE, R²)

### 9.2 Cross-validation results

### 9.3 Model comparison

### 9.4 Best model selection

### 9.5 Spatial patterns in predictions / Key findings from best model

## 10 Discussion & Conclusion

### 10.1 Key findings

### 10.2 Policy implications

### 10.3 Limitations

### 10.4 Future work

✅ Data preparation (MUST finish) ✅ Minimal EDA (quality checks only) ✅ All 4 models running with basic tuning ✅ Model comparison and evaluation



::::: markgoals
::: markgoals-header
**Note**
:::

::: markgoals-container
:::
:::::

```{r}
mdata <- read_rds("data/rawdata/mdata.rds")
```


```{r}
tm_shape(mpsz) +
  tm_polygons(
    fill = "REGION_N",
    fill.scale = tm_scale_categorical(
      values = c("#FFC4C4", "#87BAC3", "#8FABD4", "#FDE7B3", "#FDCFFA")
    ),
    fill.legend = tm_legend(title = "Region"),
    col = "white",
    lwd = 0.8
  ) +
  tm_layout(
    main.title = "Singapore Planning Regions",
    legend.outside = TRUE
  )
```


Plot the MRT station exits in a tmap.

```{r}
#| eval: false
#| fig-width: 12
#| fig-height: 8
#| code-fold: true
#tmap_mode("view")
tm_shape(mpsz) +
  tm_polygons(fill = "#FAF3F3", fill_alpha = 0.2) +
tm_shape(mrt_sf) +
  tm_symbols(
   fill = "#87A7B3", size = 0.5,
   col = "#0C4271",
#   popup = "STATION"
  ) +
  tm_layout(main.title = "MRT Stations",
            bg.color = "#F6E6E4",
            outer.margins = c(0.01, 0.01, 0.01, 0.01))
#tmap_mode("plot")
```

```{.r}
test <- st_read("data/rawdata/JTCBusinessParkLandGEOJSON.geojson") 
read_html(test$Description[1]) %>% 
  html_table() %>% 
  .[[1]]
```


