---
title: "03 - Predictive Modeling for Singapore HDB Resale Prices"
author: "Cathy C."
date: "Nov 09, 2025"
date-modified: "last-modified" #auto render date
format: html
execute: #how YML page work
  echo: true #will show code
  eval: true #will run code every time and show output
  warning: false #dont show warning msg
  freeze: true #controls rendering - no render if no changes
---

## 1 Overview

### 1.1 Background

### 1.2 Objective

The objective of this take-home exercise is to develop a spatially informed model to predict HDB resale prices in Singapore, making use of appropriate geospatial analytics methods. We will calibrate a predictive model to predict HDB resale prices for the period July to September 2025, using HDB resale transaction data from July 2024 to June 2025.

### 1.3 The Task

In this study, we will first determine potential factors that influence the HDB resale price, import the data of these factors, and process them for predictive modeling. The modeling we will investigate in this task include Multi-linear Regression (MLR), Geographically Weighted Regression (GWR), non-spatial Random Forest, as well as Geographical Weighted Random Forest Model (gwRF). Among these models, we will examine and compare them to derive the best performing model.

## 2 Installing and Loading R packages

One new package called [**jsonlite**]{style="color: #F08787;"} will be installed and loaded onto R environment. It will be used to parse json file into R as data frame.

```{r}
pacman::p_load(httr, tidyverse, sf, tmap, jsonlite, progress,
               spdep, GWmodel, SpatialML, rsample, Metrics, knitr,
               kableExtra, spatialRF, randomForestExplainer, rvest)
```

<package intro> progress - to monitor and track the computation.

## 3 The Data

We will make use of the following factors ...

-   Structural factors

    -   Area of the unit
    -   Floor level
    -   Remaining lease
    -   Age of the unit
    -   Main Upgrading Program (MUP) completed (optional)
    -   Average population age in the area??

-   Location factors

    -   Proxomity to CBD
    -   Proxomity to business park
    -   Proximity to eldercare
    -   Proximity to foodcourt/hawker centres
    -   Proximity to MRT
    -   Proximity to park
    -   Proximity to top primary school
    -   Proximity to shopping mall
    -   Proximity to supermarket
    -   Proximity to healthcare facilities
    -   Proximity to sports halls
    -   Numbers of kindergartens within 350m
    -   Numbers of childcare centres within 350m
    -   Numbers of bus stop within 350m
    -   Numbers of primary school within 1km

+--------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------+
| Data set                       | Link                                                                                                                                                                            | Source        |
+================================+=================================================================================================================================================================================+===============+
| CBD                            |                                                                                                                                                                                 |               |
+--------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------+
| Business Parks                 | [JTC Business Park Land](https://data.gov.sg/datasets?query=park&page=1&resultId=d_7555d52fb87e75da7397bd65d4f107dd)                                                            |               |
+--------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------+
| eldercare                      | [Eldercare Services](https://data.gov.sg/datasets?query=eldercare&page=1&resultId=d_f0fd1b3643ed8bd34bd403dedd7c1533)                                                           |               |
+--------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------+
| foodcourt/hawker centres       | [**Hawker Centres (GEOJSON)**](https://data.gov.sg/datasets?query=hawker+centre&page=1&resultId=d_4a086da0a5553be1d89383cd90d07ecd)                                             | data.gov.sg   |
+--------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------+
| MRT                            | [TrainStation_Aug2025](https://datamall.lta.gov.sg/content/datamall/en/static-data.html)                                                                                        | DataMall      |
+--------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------+
| Major interchange stations     | Derive from MRT based on [LTAguru](https://landtransportguru.net/bus-interchanges-and-terminals/)                                                                               |               |
+--------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------+
| Upcoming bus interchange       | Derive from MRT based on [LTAguru](https://landtransportguru.net/bus-interchanges-and-terminals/)                                                                               |               |
+--------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------+
| park                           | [Park Facilities](https://data.gov.sg/datasets?query=park&page=1&resultId=d_14d807e20158338fd578c2913953516e)                                                                   |               |
+--------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------+
| shopping mall                  | wiki + Kaggle                                                                                                                                                                   | wiki + Kaggle |
+--------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------+
| supermarket                    | Supermarkets (GEOJSON)                                                                                                                                                          | data.gov.sg   |
+--------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------+
| kindergartens within 350m      | [Kindergartens](https://data.gov.sg/datasets?query=kindergarten&resultId=d_7fe9a72b1afff18e48111772c8d0fd39&page=1)                                                             |               |
+--------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------+
| childcare centres within 350m  | [Pre-Schools Location](https://data.gov.sg/datasets?query=preschool&resultId=d_61eefab99958fd70e6aab17320a71f1c&page=1)                                                         |               |
|                                |                                                                                                                                                                                 |               |
|                                | [Listing of Centres](https://data.gov.sg/datasets?query=kindergarten&resultId=d_696c994c50745b079b3684f0e90ffc53&page=1)                                                        |               |
+--------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------+
| bus stop within 350m           |                                                                                                                                                                                 | DataMall      |
+--------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------+
| primary school within 1km      | [School Directory and Information](https://data.gov.sg/datasets?query=school+information&resultId=457&page=1)                                                                   |               |
+--------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------+
| good primary school within 1km | Primary school with data from [sophia education](https://sophiaeducation.sg/2025-singapore-psle-school-rankings/#%F0%9F%8F%AB_Top_20_Primary_Schools_in_Singapore_2025_Edition) |               |
+--------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------+
| healthcare facilities          | [CHAS Clinics](https://data.gov.sg/datasets?query=clinic&resultId=d_548c33ea2d99e29ec63a7cc9edcccedc&page=1)                                                                    |               |
+--------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------+
| Sports hall                    | [SportsFields\@SG](https://data.gov.sg/datasets?query=sports+hub&page=1&resultId=d_f71b449b4b43a69b5ecfe411b440d249)                                                            | data.gov.sg   |
+--------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------+

### 2.1 Import Data

#### 2.1.0 Singapore Maps

Below we will import *Master Plan 2019 Subzone Boundary (No Sea).*

```{r}
#| eval: false
#| code-fold: true
mpsz <- st_read("data/rawdata/MasterPlan2019SubzoneBoundaryNoSeaKML.kml") %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 3414)

# extract values from the HTML description
extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)
  rows <- page %>% html_elements("tr")
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>%
    html_element("td") %>%
    html_text2()
  
  if (length(value) == 0) NA_character_ else value
}

# extract info
mpsz <- mpsz %>%
  mutate(
    REGION_N = map_chr(Description, extract_kml_field, "REGION_N"),
    PLN_AREA_N = map_chr(Description, extract_kml_field, "PLN_AREA_N"),
    SUBZONE_N = map_chr(Description, extract_kml_field, "SUBZONE_N"),
    SUBZONE_C = map_chr(Description, extract_kml_field, "SUBZONE_C")
  ) %>%
  select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())

write_rds(mpsz, "data/rds/mpsz.rds")
```


```{r}
#| echo: false
mpsz <- read_rds("data/rds/mpsz.rds")
st_crs(mpsz)
```


 #### 2.1.1 HDB resale data Preparation
-   **Load HDB Resale Data**
This data set comprises of HDB resale transactions from July 2017 onwards. Below we will extract a sub-set of transaction records on 4-Room flat transacted during July 2024 to June 2025.

```{r}
#| eval: false
#| code-fold: true

HDBresale <- read_csv(
  "data/rawdata/ResaleflatpricesJan2017.csv") %>%
  filter(flat_type == "4 ROOM") %>%
  filter(month >= "2024-07" & month < "2025-06") # filter the sample data for the exercise.

write_rds(HDBresale, "data/rds/HDBresale.rds")
```

-   **Further processing for street names**
We do further processing for the data as the street names with Saint are usually written as St.

``` {r}
#| eval: false
#| code-fold: true
# street names is ST, but it's actually Sait.
HDBresale$street_name <- gsub(
  "ST\\.", "SAINT",
  HDBresale$street_name
)
```

```{r}
#| echo: false
HDBresale <- read_rds("data/rds/HDBresale.rds")
glimpse(HDBresale)
```

-   **Geocoding HDB Resale Data**
We use OneMap Geocode API: - <https://www.onemap.gov.sg/apidocs/.> - [ttps://www.onemap.gov.sg/apidocs/reverseGeocode](ttps://www.onemap.gov.sg/apidocs/reverseGeocode)

Below we will prepare a geocoding function. It aims to perform the following tasks:

-   Combining the block and street name into an address. The addresses will be used to geocode instead of postal code
-   Passing the address as the **searchVal** in our query
-   Sending the query to **OneMapSG search**. Note: Since we don’t need all the address details, we can set getAddrDetails as ‘N’
-   Converting response (JSON object) to text
-   Saving response in text form as a dataframe
-   Retain the latitude and longitude for our output.

``` {r}
#| eval: false
# for multiple queries
geocode <- function(block, streetname) {
  base_url <- "https://onemap.gov.sg/api/common/elastic/search"
  address <- paste(block, streetname, sep = " ")
  query <- list("searchVal" = address,
                "returnGeom" = "Y",
                "getAddrDetails" = "N", #No. can change to Y for comparison
                "pageNum" = "1") 
  res <- GET(base_url, query = query) #get URL query
  restext <- content(res, as = "text")
  
  output <- fromJSON(restext) %>%
    as.data.frame %>%
    select(results.LATITUDE, results.LONGITUDE)
  
  return(output)
}
```

-   **Run functions**

``` {r}
#| eval: false
HDBresale$LATITUDE <- 0
HDBresale$LONGITUDE <- 0

for(i in 1:nrow(HDBresale)){
  temp_output <- geocode(HDBresale[i, 4], HDBresale[i, 5])
  
  HDBresale$LATITUDE[i] <- temp_output$results.LATITUDE
  HDBresale$LONGITUDE[i] <- temp_output$results.LONGITUDE
}
```

-   **Transform to sf data frame**
We want to ensure that:
-   both input data sets must be in sf objects
-   they must be in similar projected coordinates systems

```{r}
#| eval: false
#| code-fold: true

st_crs(HDBresale)
HDBresale_sf <- HDBresale %>%
  st_as_sf(
    coords = c("LONGITUDE", "LATITUDE"),
    crs = 4326) %>% #if decimal degree
  st_transform(crs = 3414)

write_rds(HDBresale_sf, "data/rds/HDBresale_sf.rds")
```
```{r}
#| echo: false
HDBresale_sf <- read_rds("data/rds/HDBresale_sf.rds")
```


#### 2.1.2 Load raw data

::: panel-tabset

## Education

-   **Childcare centers**

```{r}
#| eval: false
#| code-fold: true
preschool_data <- st_read("data/rawdata/PreschoolsLocation.kml") %>%
  st_transform(crs = 3414)
write_rds(preschool_data, "data/rds/preschool_data.rds")
```

```{r}
#| echo: false
preschool_data <- read_rds("data/rds/preschool_data.rds")
```


Make sure the childcare coordinate system is SVY21.
```{r}
#| code-fold: true
st_crs(preschool_data)
```

-   **Kindergartens**
```{r}
#| eval: false
#| code-fold: true
kindergarten_data <- st_read("data/rawdata/Kindergartens.geojson") %>%
  st_transform(crs = 3414)

write_rds(kindergarten_data, "data/rds/kindergarten_data.rds")
```
```{r}
#| echo: false
#| code-fold: true
kindergarten_data <- read_rds("data/rds/kindergarten_data.rds")
```

Make sure the kindergarten coordinate system is SVY21.

```{r}
#| code-fold: true
st_crs(kindergarten_data)
```

-   **Primary schools data**

```{r}
#| code-fold: true
#| eval: false
primschool_data <- read_csv("data/rawdata/General Information of schools.csv") %>%
  select(school_name, postal_code, address)

write_rds(primschool_data, "data/rds/primschool_data.rds") 
```

```{r}
#| echo: false
primschool_data <- read_rds("data/rds/primschool_data.rds")
```

We will geocode for primary school data in later section.

## Business

-   **CBD**

For CBD, we will extract the area for CBD from `mpsz`.
```{r}
#| code-fold: true
#| eval: false
cbd_data <- mpsz %>%
  filter(SUBZONE_N %in% c("RAFFLES PLACE", "MARINA CENTRE", "TANJONG PAGAR", "CITY HALL", "CENTRAL SUBZONE")) %>%
  st_union()

write_rds(cbd_data, "data/rds/cbd_data.rds")
```

```{r}
#| echo: false
cbd_data <- read_rds("data/rds/cbd_data.rds")
```

```{r}
glimpse(cbd_data)
```
cbd_data is now a single merged polygon representing the entire CBD area. We will use this to calculate distance in later section.

-   **Business park**

Business parks are typically smaller than planning areas, and we have multiple business parks to consider. Therefore, we will calculate centroid for measuring proximity, and transform polygon data to centroid point data.

```{r}
#| code-fold: true
#| eval: false

bizpark_data <- st_read("data/rawdata/JTCBusinessParkLandGEOJSON.geojson") %>% 
# calculate centroids
  mutate(centroid = st_centroid(geometry)) %>%
  select(OBJECTID_1, centroid)
```

## Recreation

-   **Shopping malls**

```{r}
#| code-fold: true
#| eval: false

sgmalls_rawdata <- read_csv("data/rawdata/singapore_malls.csv") %>%
  filter(!Mall_Name %in% c('JCube', 'JCube (closed)', 'Liang Court', 'Lot 1', 
                           'Nex', 'Novena Square Shopping Mall'))

sgmalls_coordinates <- read_csv("data/rawdata/shopping_mall_coordinates.csv") %>%
  rename(Mall_Name = 'Mall Name')

merged_malls <- full_join(sgmalls_rawdata, sgmalls_coordinates, by = "Mall_Name")

```
After the merge, we examine if any shopping malls turn out without coordinates data. 
```{r}
#| code-fold: true
#| eval: false
malls_without_coords <- merged_malls %>%
  filter(is.na(LONGITUDE) | is.na(LATITUDE)) %>%
  pull(Mall_Name)

malls_without_coords
```

Amber Mansions - demolished in 1984
Bedok Point
Capitol Centre
Ellenborough Market
Funan DigitaLife Mall - duplicate
Great World - duplicate
GV Yishun - cinema
Jurong Entertainment Centre - demolished 2010
Shaw Tower - demolished
Specialists' Shopping Centre - redeveloped as Orchard Gateway
Tanglin Shopping Centre - CLOSED
The Golden Mile - CLOSED
The Verge - closed
Woodland Mall - change to Woods square mall

For this case, we manually check with Google maps
```{r}
#| code-fold: true
#| eval: false

coords_to_fill <- tibble(
  Mall_Name = c("Alexandra Retail Centre", "Capitol Singapore", "Change Alley",
                "Chinatown Point", "Clementi Mall", "Dawson Place", 
                "Esplanade Mall", "Forum The Shopping Mall", "Fu Lu Shou Complex", 
                "HarbourFront Centre", "Holland Road Shopping Centre", 
                "Holland Village Shopping Mall", 
                "Mandarin Gallery", "Mustafa Centre", "Novena Square", 
                "Orchard Towers", "Paya Lebar Quarter", "Peninsula Plaza",
                "Punggol Coast Mall", "Scotts Shopping Centre", "Seletar Mall",
                "Shaw House and Centre", "Singpost Center",
                "South Beach", "Suntec City Mall",
                "Tang Plaza", "The Cathay", 
                "The Majestic", "The Rail Mall", 
                "VivoCity", "Woodland Mall"),
  
  LONGITUDE = c(103.8015, 103.8511, 103.8518, 
                103.8446, 103.7642, 103.8107,
                103.8561, 103.8284, 103.8543,
                103.8201, 103.7952, 
                103.7937, 
                103.8363, 103.8554, 103.8439,
                103.8292, 103.8930, 103.8508,
                103.9104, 103.8329, 103.8758,
                103.8318, 103.8940, 
                103.8557, 103.8581, 
                103.8329, 103.8477,
                103.8432, 103.7678,
                103.8231, 103.7859
                ),
  LATITUDE = c(1.2741176, 1.293231, 1.284643, 
               1.285434, 1.315111, 1.293505, 
               1.290017, 1.306391, 1.301906,
               1.265355, 1.310419, 1.319192,
               
               1.311708, 
               1.317859, 1.292544, 1.302237, 
               1.310192, 1.320230, 1.307743, 
               1.415818, 1.306093, 1.392464,
               1.306507, 
               1.294485, 1.295778, 
               1.305221, 1.300409, 
               1.284871, 1.358212, 
               1.265357, 1.438082
               )
) 

merged_malls <- merged_malls %>%
  rows_update(coords_to_fill, by = "Mall_Name", unmatched = "ignore")

merged_malls <-  merged_malls %>%
  mutate(
    Mall_Name = case_when(
      Mall_Name == "Woods Square Mall" ~ "Woodland Mall",
      TRUE ~ Mall_Name
    )
  )

# add new mall data
new_malls <- tibble(
  Mall_Name = "Woodleigh Mall",
  LONGITUDE = 103.8718,
  LATITUDE = 1.339560
)  

merged_malls <- merged_malls %>%
  bind_rows(new_malls)
```

Check the malls with missing coordinates, meaning that we did not find the coordinates for these mall as they're closed.
```{r}
#| code-fold: true
#| eval: false
malls_with_na <- merged_malls %>%
  filter(is.na(LONGITUDE) | is.na(LATITUDE)) 

malls_with_na
```
These malls are the ones duplicate or closed. We will go ahead to remove them. We will also remove 2 duplicate malls observed from *merged_malls*. 
```{r}
#| code-fold: true
#| eval: false
merged_malls <- merged_malls %>%
  filter(!is.na(LONGITUDE) & !is.na(LATITUDE)) %>%
  distinct(Mall_Name, .keep_all = TRUE) %>%
  filter(!Mall_Name %in% c("Suntec City Mall", "Capitol Piazza"))

# Check how many remain
nrow(merged_malls)
write_rds(merged_malls, "data/rds/malls_data.rds")
```

```{r}
#| echo: false

malls_data <- read_rds("data/rds/malls_data.rds")
```


-   **Sports halls**


```{.r}
read_html(cc_data$Description[1]) %>% 
  html_table() %>% 
  .[[1]]
```


```{r}
#| code-fold: true
#| eval: false
sports_data <- st_read("data/rawdata/SportsFieldsSG.geojson") %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 3414)

extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)
  rows <- page %>% html_elements("tr")
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>%
    html_element("td") %>%
    html_text2()
  
  if (length(value) == 0) NA_character_ else value
}

# extract info
sports_data <- sports_data %>%
  mutate(
    name = map_chr(Description, extract_kml_field, "NAME"),
    description = map_chr(Description, extract_kml_field, "DESCRIPTION")
  ) %>%
  select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())

write_rds(mpsz, "data/rds/sports_data.rds")
```
-   **Parks**

```{r}
#| code-fold: true
#| eval: false
park_data <- st_read("data/rawdata/MasterPlan2019SDCPParkandOpenSpacelayer.kml") %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_make_valid() %>%
  st_transform(crs = 3414)

# Calculate centroids 
park_data <- park_data %>%
  mutate(centroid = st_centroid(geometry))

write_rds(park_data, "data/rds/park_data.rds")
```
```{r}
#| echo: false
park_data <- read_rds("data/rds/park_data.rds")
```

-   **Supermarkets**

```{r}
#| code-fold: true
#| eval: false


```

-   **Community clubs**
```{r}
#| code-fold: true
#| eval: false
cc_data <- st_read("data/rawdata/CommunityClubs.geojson") %>%
  st_transform(crs = 3414) %>%
  mutate(centroid = st_centroid(geometry))

extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)
  rows <- page %>% html_elements("tr")
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>%
    html_element("td") %>%
    html_text2()
  
  if (length(value) == 0) NA_character_ else value
}

# extract info
cc_data <- cc_data %>%
  mutate(
    name = map_chr(Description, extract_kml_field, "NAME")
  ) %>%
  select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())


write_rds(cc_data, "data/rds/cc_data.rds")
```

```{r}
#| echo: false
cc_data <- read_rds("data/rds/cc_data.rds")
```

## Transportation

-   **MRT**

Below we will import MRT station exits, and compute the centroid of these exit for these stations as the poin for proximity calculation to HDB units.

```{r}
#| code-fold: true
#| eval: false
mrt <- st_read("data/LTAMRTStationExitKML.kml") %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 3414)
```

Below code is to write MRT station data into a rsd file, so we can directly call it without processing the data set again.

```{r}
#| eval: false
extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)
  rows <- page %>% html_elements("tr")
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>%
    html_element("td") %>%
    html_text2()
  
  if (length(value) == 0) NA_character_ else value
}

mrt_all <- mrt %>%
  mutate(
    STATION = map_chr(Description, extract_kml_field, "STATION_NA"),
    EXIT_CODE = map_chr(Description, extract_kml_field, "EXIT_CODE")
  ) %>%         
  select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())

mrt_sf <- mrt_all %>%
  # Get just one exit per station 
  distinct(STATION, .keep_all = TRUE) 
  
#  select(-Name, -Description) %>%
#  relocate(geometry, .after = last_col())

write_rds(mrt_all, "data/rds/mrt_all.rds")
write_rds(mrt_sf, "data/rds/mrt_sf.rds")
```

```{r}
#| eval: false
#| echo: false
mrt_all <- read_rds("data/rds/mrt_all.rds")
mrt_sf <- read_rds("data/rds/mrt_sf.rds")
```

Plot the MRT station exits in a tmap.

```{r}
#| eval: false
#| fig-width: 12
#| fig-height: 8
#| code-fold: true
#tmap_mode("view")
tm_shape(mpsz) +
  tm_polygons(fill = "#FAF3F3", fill_alpha = 0.2) +
tm_shape(mrt_sf) +
  tm_symbols(
   fill = "#87A7B3", size = 0.5,
   col = "#0C4271",
#   popup = "STATION"
  ) +
  tm_layout(main.title = "MRT Stations",
            bg.color = "#F6E6E4",
            outer.margins = c(0.01, 0.01, 0.01, 0.01))
#tmap_mode("plot")
```

-   **Bus stops**

Below we will import *Bus Stop Location* shapfile from LTA DataMall.

```{r}
#| code-fold: true
#| eval: false
BusStop = st_read(dsn = "data/LTADataMall",
                  layer = "BusStop")
BusStop <- BusStop %>%
  st_transform(crs = 3414)

write_rds(BusStop, "data/rds/BusStop.rds")
```

Check again on BusStop:

```{r}
#| eval: false
BusStop <- read_rds("data/rds/BusStop.rds")
st_crs(BusStop)
```

## Healthcare
-   **Clinics**
```{r}
#| code-fold: true
#| eval: false


```

-   **Eldercare**

```{r}
#| code-fold: true
#| eval: false


```
:::

#### 2.1.3 Initial data inspection


### 2.2 Structural factors

flat type, floor level, town, planning area, missing values, area of unit, age of unit, main upgrading program, flat model

### 2.3 Location factors Preparation

#### 2.3.1 Geocode amenity locations

#### 2.3.2 Validate coordinates

check for outliers / errors Verify coordinates are in Singapore (plot tmaps)

### 2.4 Proximity Analysis
##-------


::::: markgoals
::: markgoals-header
**Note**
:::

::: markgoals-container
-   `st_is_within_distance()` of [**sf**]{style="color: #F08787;"} package is used to determines whether two spatial objects (HDB property and preschool) are within 350m distance of each other. It returns a logical matrix (or a sparse list of indices if sparse = TRUE) indicating for each geometry in HDB property which geometries in preschool are within 350m distance.
-   `map_int()` of [**purr**]{style="color: #F08787;"} package is used to the number of preschools found in the newly created WITHIN_350M_PRISCHOOL field.
:::
:::::

### 2.8 Computing shortest distance

Instead of counting number of preschools located within 350m of each HDB resale property, code chunk below is used to compute the distance between the nearest preschool from each HDB resale property.

``` r
HDBresale_sf <- HDBresale_sf %>%
  mutate(
    PROX_PRESCHOOL = as.numeric(
      st_distance(geometry,
                  preschool[st_nearest_feature(
                    geometry, preschool), ], 
                  by_element = TRUE)
      )
    )
```



##--------

#### 2.4.1 Calculate distance to nearby amenities

::: panel-tabset
## Future interchange

## Shopping mall

## CBD/town centre
below code is a sample.
```{r}
#| eval: false
resale <- HDBresale %>%
  mutate(
    dist_to_downtown_km = as.numeric(st_distance(geometry, downtown_core)) / 1000,
    log_dist_downtown = log1p(dist_to_downtown_km)
  )

# Save
saveRDS(resale, "data/resale_with_features.rds")

```

## Business park

## eldercare

## Park

## Supermarket
:::

#### 2.4.2 Create binary proximity indicators

Next we will derive a binary data to find out which schools are within 350 meters.

``` {r}
#| eval: false
within_350m <- st_is_within_distance(
  HDBresale_sf, preschool, dist = 350)

HDBresale_sf <- HDBresale_sf %>%
  mutate(prischool_350m = map_int(within_350m, length)
)
```


::: panel-tabset
## Top primary school

Within 1 km

## MRT Within 500 m

## Hawker centre Within 500m

:::

### 2.5 Amenity Density Analysis

Computing number of location factors

::: panel-tabset
## kindergarten

within 350m

## childcare

within 350m

## bus stop

within 350m

## primary school

within 350m
:::

### 2.6 Final Dataset Assembly

#### 2.6.1 Merge all features into master dataset

#### 2.6.2 Handle remaining missing values

#### 2.6.3 Save processed dataset

## 3 Data Quality & Spatial Overview

### 3.1 Data Quality Check

-   Summary statistics
-   Check for outliers for price
-   Coordinate validation

### 3.2 Spatial Visualization of Amenities

#### 3.2.1 Target Variable Analysis

HDB resale price distribution

#### 3.2.2 Interactive map of all amenities

::: panel-tabset
## HDB Units

-   All resale units (color by price)

## Education

-   Primary schools (top 20 highlighted)
-   Kindergartens
-   Childcare centres

## Transport

-   MRT stations (current + future)
-   Major bus interchanges

## Commercial

-   Shopping malls
-   Hawker centres
-   Supermarkets

## Business

-   CBD
-   BUsiness parks

## Recreation & Others

-   Parks
-   Sports facilities
-   Healthcare
:::

## 4 Pre-modeling Preparation

### 4.1 Multicollinearity Check

-   Correlation matrix of predictors
-   VIF scores
-   Identify multicollinearity issues Remove highly correlated features

### 4.2 Feature Selection

-   Final feature list for modeling

### 4.3 Train/Test Split

#### 4.3.1 Stratified Spatial Sampling

``` r
library(dplyr)
library(sf)

set.seed(123)

# Stratify by town AND year (if multi-year)
hdb_sample <- hdb_data %>%
  # Ensure temporal coverage
  group_by(year, town) %>%
  # Sample proportionally
  slice_sample(prop = 5000/nrow(hdb_data)) %>%
  ungroup()

# If you get slightly more/less than 5000, adjust:
if(nrow(hdb_sample) > 5000) {
  hdb_sample <- hdb_sample %>% sample_n(5000)
} else if(nrow(hdb_sample) < 5000) {
  # Add more randomly to reach 5000
  additional <- hdb_data %>%
    anti_join(hdb_sample) %>%
    sample_n(5000 - nrow(hdb_sample))
  hdb_sample <- bind_rows(hdb_sample, additional)
}

# Verify distribution
hdb_sample %>% count(town)
hdb_sample %>% count(flat_type)

# Save
write_csv(hdb_sample, "data/processed/hdb_sample_5000.csv")
```

``` r
cat("Original dataset:", nrow(hdb_data), "transactions
")
cat("Sample size:", nrow(hdb_sample), "transactions
")
cat("Sampling rate:", round(5000/nrow(hdb_data)*100, 2), "%
")
```

-   Total original dataset size
-   Sampling method used
-   Why 5000 was chosen
-   How representativeness was ensured

``` r
# Compare distributions
par(mfrow = c(2,2))

# Price distribution
plot(density(hdb_data$resale_price), main = "Full Dataset", col = "blue")
plot(density(hdb_sample$resale_price), main = "Sample", col = "red")

# Town distribution
barplot(table(hdb_data$town), main = "Full Dataset - Towns")
barplot(table(hdb_sample$town), main = "Sample - Towns")
```

#### 4.3.2 Tran/Test split

``` r
# 80/20 split
set.seed(456)
train_idx <- sample(1:nrow(hdb_sample), 0.8 * nrow(hdb_sample))

train <- hdb_sample[train_idx, ]   # 4000 records
test <- hdb_sample[-train_idx, ]   # 1000 records

cat("Train size:", nrow(train), "
")
cat("Test size:", nrow(test), "
")
```

## 5 Multiple Linear Regression (MLR)

### 5.1 Model specification

### 5.2 Model fitting

### 5.3 Results & interpretation

### 5.4 Residual diagnostics

## 6 Geographically Weighted Regression (GWR)

### 6.1 Bandwidth selection

### 6.2 Model fitting

### 6.3 Local coefficient maps

### 6.4 Results & interpretation

## 7 Non-sptial Random Forest Model

### 7.1 Feature selection

### 7.2 Hyperparameter tuning

### 7.3 Model fitting

### 7.4 Feature importance

### 7.5 Results & interpretation

## 8 Geographical Weighted Random Forest Model (gwRF)

### 8.1 Model specification

### 8.2 Bandwidth selection

### 8.3 Model fitting

### 8.4 Local predictions & interpretation

## 9 Model Evaluation

### 9.1 Performance metrics (RMSE, MAE, R²)

### 9.2 Cross-validation results

### 9.3 Model comparison

### 9.4 Best model selection

### 9.5 Spatial patterns in predictions / Key findings from best model

## 10 Discussion & Conclusion

### 10.1 Key findings

### 10.2 Policy implications

### 10.3 Limitations

### 10.4 Future work

✅ Data preparation (MUST finish) ✅ Minimal EDA (quality checks only) ✅ All 4 models running with basic tuning ✅ Model comparison and evaluation



::::: markgoals
::: markgoals-header
**Note**
:::

::: markgoals-container
:::
:::::

```{r}
mdata <- read_rds("data/rawdata/mdata.rds")
```

```{r}
```{r}
tm_shape(mpsz) +
  tm_polygons(
    fill = "REGION_N",
    fill.scale = tm_scale_categorical(
      values = c("#FFC4C4", "#87BAC3", "#8FABD4", "#FDE7B3", "#FDCFFA")
    ),
    fill.legend = tm_legend(title = "Region"),
    col = "white",
    lwd = 0.8
  ) +
  tm_layout(
    main.title = "Singapore Planning Regions",
    legend.outside = TRUE
  )
```
```

