---
title: "Take-home Exercise: Public Good"
author: "Cathy C."
date: "Sep 17, 2025"
date-modified: "last-modified" #auto render date
format: html
execute: #how YML page work
  echo: true #will show code
  eval: true #will run code every time and show output
  warning: false #dont show warning msg
  freeze: true #controls rendering - no render if no changes
---

In this exercise, we will go through the steps for preparing data for Take-home exercise one where we will investigate the spatial and spatio-temporal patterns of new businesses and firms established in Singapore.

## 1 Loading the R package

We will use **tidyverse, sf, tmap, tmap,** and **httr** packages for this exercise to process data sets.

```{r}
pacman::p_load(tidyverse, sf, tmap, httr, knitr)
```

## 2 Data Wrangling
### 2.1 Importing ACRA data

We will import the ACRA data and transform them as tibble frame

```{r}

folder_path <- "data/aspatial/acra"
file_list <- list.files(path = folder_path,
                        pattern = "^ACRA*.*\\.csv$",
                        full.names = TRUE)
acra_data <- file_list %>%
  map_dfr(read_csv) #import and append all of them together!!
  
```

Then, we will write the data into .rds to store it in the data folder. This is so no need to process data every time.

```{r}

write_rds(acra_data, "data/rds/acra_data.rds")
```

### 2.2 Tidying ACRA data

With this step, we will extract the business codes 9601, 9602, and 86204, then rename the columns and transform the data type necessarily.

The business types chosen are:

| Business Code | Types / subtypes |
|------------------|------------------------------------------------------|
| **9601** | **Washing and (Dry-) Cleaning of Textile and Fur Products** |
| 96011 | Laundry and dry cleaning services except self-operated laundries |
| 96012 | Self-operated laundries |
| **9602** | **Hairdressing, Beauty and Other Personal Care Services** |
| 96021 | Hairdressing salons/shops (including barber shops) |
| 96022 | Beauty salons and spas (including slimming, skin care and hair care centres) |
| 96023 | Manicure and pedicure services |
| **86024** | **Dental services** |

: Business Code & Types


We use below code to retrieve the data between Jan 1, 2024 to June 30, 2025 for the 3 different major business types. 

::: panel-tabset
## 9601
**Washing and (Dry-) Cleaning of Textile and Fur Products**
Business Code 9601 emcompasses Laundry cleaning and self-operated laundries (96011 & 96012)

```{r}
biz_9601 <- acra_data %>%
  select(1:24) %>%
  filter(primary_ssic_code == 96011 | primary_ssic_code == 96012) %>%
  rename(date = registration_incorporation_date) %>%
  mutate(date = as.Date(date),
         YEAR = year(date),
         MONTH_NUM = month(date),
         MONTH_ABBR = month(date, label = TRUE, abbr = TRUE)) %>% #this gives Jan, Feb, Mar...
  mutate(postal_code = str_pad(postal_code,
                               width = 6, side = "left", pad = "0")) %>% #if missing digits, will replace with zero.
filter(YEAR == 2024 | YEAR == 2025 & MONTH_NUM <= 6)

glimpse(biz_9601)
```

There are 150 observations we have retrieved for this analysis.

## 9602
**Hairdressing, Beauty and Other Personal Care Services**
Business Code 9602 includes hair dressing, beauty salons and manicure & pedicure services.
```{r}
biz_9602 <- acra_data %>%
  select(1:24) %>%
  filter(primary_ssic_code == 96021 | primary_ssic_code == 96022 | primary_ssic_code == 96023) %>%
  rename(date = registration_incorporation_date) %>%
  mutate(date = as.Date(date),
         YEAR = year(date),
         MONTH_NUM = month(date),
         MONTH_ABBR = month(date, label = TRUE, abbr = TRUE)) %>% #this gives Jan, Feb, Mar...
  mutate(postal_code = str_pad(postal_code,
                               width = 6, side = "left", pad = "0")) %>% #if missing digits, will replace with zero.
filter(YEAR == 2024 | YEAR == 2025 & MONTH_NUM <= 6)

glimpse(biz_9602)
```

There are 1762 observations we have retrieved for this analysis.

## 86204
**Dental services**
```{r}
biz_86204 <- acra_data %>%
  select(1:24) %>%
  filter(primary_ssic_code == 86204) %>%
  rename(date = registration_incorporation_date) %>%
  mutate(date = as.Date(date),
         YEAR = year(date),
         MONTH_NUM = month(date),
         MONTH_ABBR = month(date, label = TRUE, abbr = TRUE)) %>% #this gives Jan, Feb, Mar...
  mutate(postal_code = str_pad(postal_code,
                               width = 6, side = "left", pad = "0")) %>% #if missing digits, will replace with zero.
filter(YEAR == 2024 | YEAR == 2025 & MONTH_NUM <= 6)

glimpse(biz_86204)
```

There are 81 observations we have retrieved for this analysis.
:::

### 2.3 Geocoding

(Avoid doing so for \> 5K datasets) We will perform the following in this step:

-   Extract unique postal code from the poastal_code field of the tidied tibble data frame.
-   Perform geocode by submitting the postal code to SLA OneMap Reverse Geocode API.
-   Save the return results into 2 data frame namely *found* and *not_found*.

::: panel-tabset
## 9601

```{r}

postcodes_9601 <- unique(biz_9601$postal_code)

url <- "https://onemap.gov.sg/api/common/elastic/search"

found_9601 <- data.frame()
not_found_9601 <- data.frame(postcode_9601 = character())

for (pc_9601 in postcodes_9601) {
  query_9601 <- list(
    searchVal = pc_9601,
    returnGeom = "Y",
    getAddrDetails = "Y",
    pageNum = "1"
  )
  
  res_9601 <- GET(url, query = query_9601)
  json_9601 <- content(res_9601)
  
  if(json_9601$found != 0) {
    df_9601 <- as.data.frame(json_9601$results, stringsAsFactors = FALSE)
    df_9601$input_postcode_9601 <- pc_9601
    found_9601 <- bind_rows(found_9601, df_9601)
  } else{
    not_found_9601 <- bind_rows(not_found_9601, data.frame(postcode_9601 = pc_9601))
  }
}
```

## 9602

```{r}

postcodes_9602 <- unique(biz_9602$postal_code)

url <- "https://onemap.gov.sg/api/common/elastic/search"

found_9602 <- data.frame()
not_found_9602 <- data.frame(postcode_9602 = character())

for (pc_9602 in postcodes_9602) {
  query_9602 <- list(
    searchVal = pc_9602,
    returnGeom = "Y",
    getAddrDetails = "Y",
    pageNum = "1"
  )
  
  res_9602 <- GET(url, query = query_9602)
  json_9602 <- content(res_9602)
  
  if(json_9602$found != 0) {
    df_9602 <- as.data.frame(json_9602$results, stringsAsFactors = FALSE)
    df_9602$input_postcode_9602 <- pc_9602
    found_9602 <- bind_rows(found_9602, df_9602)
  } else{
    not_found_9602 <- bind_rows(not_found_9602, data.frame(postcode_9602 = pc_9602))
  }
}
```

## 86204

```{r}

postcodes_86204 <- unique(biz_86204$postal_code)

url <- "https://onemap.gov.sg/api/common/elastic/search"

found_86204 <- data.frame()
not_found_86204 <- data.frame(postcode_86204 = character())

for (pc_86204 in postcodes_86204) {
  query_86204 <- list(
    searchVal = pc_86204,
    returnGeom = "Y",
    getAddrDetails = "Y",
    pageNum = "1"
  )
  
  res_86204 <- GET(url, query = query_86204)
  json_86204 <- content(res_86204)
  
  if(json_86204$found != 0) {
    df_86204 <- as.data.frame(json_86204$results, stringsAsFactors = FALSE)
    df_86204$input_postcode_86204 <- pc_86204
    found_86204 <- bind_rows(found_86204, df_86204)
  } else{
    not_found_86204 <- bind_rows(not_found_86204, data.frame(postcode_86204 = pc_86204))
  }
}
```

:::
### 2.4 Appending the location information
::: panel-tabset

## 9601
```{r}

biz_9601 = biz_9601 %>%
  left_join(found_9601,
            by = c('postal_code' = 'POSTAL'))
```

## 9602
```{r}

biz_9602 = biz_9602 %>%
  left_join(found_9602,
            by = c('postal_code' = 'POSTAL'))
```


## 86204
```{r}

biz_86204 = biz_86204 %>%
  left_join(found_86204,
            by = c('postal_code' = 'POSTAL'))
```

:::
### 2.5 Converting into SF data frame
::: panel-tabset
## 9601

```{r}

biz_9601_sf <- st_as_sf(biz_9601,
                         coords = c("X", "Y"),
                         crs = 3414)
```


## 9602
```{r}

biz_9602_sf <- st_as_sf(biz_9602,
                         coords = c("X", "Y"),
                         crs = 3414)
```

## 86204
```{r}

biz_86204_sf <- st_as_sf(biz_86204,
                         coords = c("X", "Y"),
                         crs = 3414)
```

::: 

```{r}

kable(head(biz_9601_sf))
```

\*\* planning subzone area \> speed up computation.

\*\* narrow down to study area.

\*\*
