---
title: "Take-home Exercise: Public Good"
author: "Cathy C."
date: "Sep 17, 2025"
date-modified: "last-modified" #auto render date
format: html
execute: #how YML page work
  echo: true #will show code
  eval: true #will run code every time and show output
  warning: false #dont show warning msg
  freeze: true #controls rendering - no render if no changes
---
## Introduction
In this exercise, we will go through the steps for preparing data for Take-home exercise one where we will investigate the spatial and spatio-temporal patterns of new businesses and firms established in Singapore.

## 1 Loading the R package

We will use **tidyverse, sf, tmap, tmap,** and **httr** packages for this exercise to process data sets.

```{r}
pacman::p_load(tidyverse, sf, tmap, httr, knitr, rvest, spatstat, terra)
```

## 2 Data Wrangling
### 2.1 Importing ACRA data

We will import the ACRA data and transform them as tibble frame

```{r}

folder_path <- "data/aspatial/acra"
file_list <- list.files(path = folder_path,
                        pattern = "^ACRA*.*\\.csv$",
                        full.names = TRUE)
acra_data <- file_list %>%
  map_dfr(read_csv) #import and append all of them together!!
  
```

Then, we will write the data into .rds to store it in the data folder. This is so no need to process data every time.

```{r}

write_rds(acra_data, "data/rds/acra_data.rds")
```

### 2.2 Tidying ACRA data

With this step, we will extract the business codes 9601, 9602, and 86204, then rename the columns and transform the data type necessarily.

The business types chosen are:

| Business Code | Types / subtypes |
|------------------|------------------------------------------------------|
| **9601** | **Washing and (Dry-) Cleaning of Textile and Fur Products** |
| 96011 | Laundry and dry cleaning services except self-operated laundries |
| 96012 | Self-operated laundries |
| **9602** | **Hairdressing, Beauty and Other Personal Care Services** |
| 96021 | Hairdressing salons/shops (including barber shops) |
| 96022 | Beauty salons and spas (including slimming, skin care and hair care centres) |
| 96023 | Manicure and pedicure services |
| **86024** | **Dental services** |

: Business Code & Types


We use below code to retrieve the data between Jan 1, 2024 to June 30, 2025 for the 3 different major business types. 

::: panel-tabset
## 9601
**Washing and (Dry-) Cleaning of Textile and Fur Products**
Business Code 9601 emcompasses Laundry cleaning and self-operated laundries (96011 & 96012)

```{r}
biz_9601 <- acra_data %>%
  select(1:24) %>%
  filter(primary_ssic_code == 96011 | primary_ssic_code == 96012) %>%
  rename(date = registration_incorporation_date) %>%
  mutate(date = as.Date(date),
         YEAR = year(date),
         MONTH_NUM = month(date),
         MONTH_ABBR = month(date, label = TRUE, abbr = TRUE)) %>% #this gives Jan, Feb, Mar...
  mutate(postal_code = str_pad(postal_code,
                               width = 6, side = "left", pad = "0")) %>% #if missing digits, will replace with zero.
filter(YEAR == 2024 | YEAR == 2025 & MONTH_NUM <= 6)

glimpse(biz_9601)
```

There are 150 observations we have retrieved for this analysis.

## 9602
**Hairdressing, Beauty and Other Personal Care Services**
Business Code 9602 includes hair dressing, beauty salons and manicure & pedicure services.
```{r}
biz_9602 <- acra_data %>%
  select(1:24) %>%
  filter(primary_ssic_code == 96021 | primary_ssic_code == 96022 | primary_ssic_code == 96023) %>%
  rename(date = registration_incorporation_date) %>%
  mutate(date = as.Date(date),
         YEAR = year(date),
         MONTH_NUM = month(date),
         MONTH_ABBR = month(date, label = TRUE, abbr = TRUE)) %>% #this gives Jan, Feb, Mar...
  mutate(postal_code = str_pad(postal_code,
                               width = 6, side = "left", pad = "0")) %>% #if missing digits, will replace with zero.
filter(YEAR == 2024 | YEAR == 2025 & MONTH_NUM <= 6)

glimpse(biz_9602)
```

There are 1762 observations we have retrieved for this analysis.

## 86204
**Dental services**
```{r}
biz_86204 <- acra_data %>%
  select(1:24) %>%
  filter(primary_ssic_code == 86204) %>%
  rename(date = registration_incorporation_date) %>%
  mutate(date = as.Date(date),
         YEAR = year(date),
         MONTH_NUM = month(date),
         MONTH_ABBR = month(date, label = TRUE, abbr = TRUE)) %>% #this gives Jan, Feb, Mar...
  mutate(postal_code = str_pad(postal_code,
                               width = 6, side = "left", pad = "0")) %>% #if missing digits, will replace with zero.
filter(YEAR == 2024 | YEAR == 2025 & MONTH_NUM <= 6)

glimpse(biz_86204)
```

There are 81 observations we have retrieved for this analysis.
:::

### 2.3 Geocoding

(Avoid doing so for \> 5K datasets) We will perform the following in this step:

-   Extract unique postal code from the poastal_code field of the tidied tibble data frame.
-   Perform geocode by submitting the postal code to SLA OneMap Reverse Geocode API.
-   Save the return results into 2 data frame namely *found* and *not_found*.

::: panel-tabset
## 9601

```{r}

postcodes_9601 <- unique(biz_9601$postal_code)

url <- "https://onemap.gov.sg/api/common/elastic/search"

found_9601 <- data.frame()
not_found_9601 <- data.frame(postcode_9601 = character())

for (pc_9601 in postcodes_9601) {
  query_9601 <- list(
    searchVal = pc_9601,
    returnGeom = "Y",
    getAddrDetails = "Y",
    pageNum = "1"
  )
  
  res_9601 <- GET(url, query = query_9601)
  json_9601 <- content(res_9601)
  
  if(json_9601$found != 0) {
    df_9601 <- as.data.frame(json_9601$results, stringsAsFactors = FALSE)
    df_9601$input_postcode_9601 <- pc_9601
    found_9601 <- bind_rows(found_9601, df_9601)
  } else{
    not_found_9601 <- bind_rows(not_found_9601, data.frame(postcode_9601 = pc_9601))
  }
}
```

## 9602

```{r}

postcodes_9602 <- unique(biz_9602$postal_code)

url <- "https://onemap.gov.sg/api/common/elastic/search"

found_9602 <- data.frame()
not_found_9602 <- data.frame(postcode_9602 = character())

for (pc_9602 in postcodes_9602) {
  query_9602 <- list(
    searchVal = pc_9602,
    returnGeom = "Y",
    getAddrDetails = "Y",
    pageNum = "1"
  )
  
  res_9602 <- GET(url, query = query_9602)
  json_9602 <- content(res_9602)
  
  if(json_9602$found != 0) {
    df_9602 <- as.data.frame(json_9602$results, stringsAsFactors = FALSE)
    df_9602$input_postcode_9602 <- pc_9602
    found_9602 <- bind_rows(found_9602, df_9602)
  } else{
    not_found_9602 <- bind_rows(not_found_9602, data.frame(postcode_9602 = pc_9602))
  }
}
```

## 86204

```{r}

postcodes_86204 <- unique(biz_86204$postal_code)

url <- "https://onemap.gov.sg/api/common/elastic/search"

found_86204 <- data.frame()
not_found_86204 <- data.frame(postcode_86204 = character())

for (pc_86204 in postcodes_86204) {
  query_86204 <- list(
    searchVal = pc_86204,
    returnGeom = "Y",
    getAddrDetails = "Y",
    pageNum = "1"
  )
  
  res_86204 <- GET(url, query = query_86204)
  json_86204 <- content(res_86204)
  
  if(json_86204$found != 0) {
    df_86204 <- as.data.frame(json_86204$results, stringsAsFactors = FALSE)
    df_86204$input_postcode_86204 <- pc_86204
    found_86204 <- bind_rows(found_86204, df_86204)
  } else{
    not_found_86204 <- bind_rows(not_found_86204, data.frame(postcode_86204 = pc_86204))
  }
}
```

:::
### 2.4 Appending the location information
::: panel-tabset

## 9601
```{r}

biz_9601 = biz_9601 %>%
  left_join(found_9601,
            by = c('postal_code' = 'POSTAL'))
```

## 9602
```{r}

biz_9602 = biz_9602 %>%
  left_join(found_9602,
            by = c('postal_code' = 'POSTAL'))
```


## 86204
```{r}

biz_86204 = biz_86204 %>%
  left_join(found_86204,
            by = c('postal_code' = 'POSTAL'))
```

:::
### 2.5 Converting into SF data frame
::: panel-tabset
## 9601

```{r}

biz_9601_sf <- st_as_sf(biz_9601,
                         coords = c("X", "Y"),
                         crs = 3414)
```


## 9602
```{r}

biz_9602_sf <- st_as_sf(biz_9602,
                         coords = c("X", "Y"),
                         crs = 3414)
```

## 86204
```{r}

biz_86204_sf <- st_as_sf(biz_86204,
                         coords = c("X", "Y"),
                         crs = 3414)
```

::: 

```{r}

kable(head(biz_9601_sf))
```

### 2.6 Import Geospatial data into R
The Master Plan 2019 Boundary (No Sea) (KML) was downloaded from data.gov.sg that consists of the geographical boundary of Singapore at the planning subzone level. The data is based on URA Master Plan 2019.
We will use this to plot the boundaries of the subzones.

:::panel-tabset
## Import data
```{r}
mpsz <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaKML.kml") %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 3414)
```

## Tidy data
```{r}
extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)
  rows <- page %>% html_elements("tr")
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>%
    html_element("td") %>%
    html_text2()
  
  if (length(value) == 0) NA_character_ else value
}

mpsz <- mpsz %>%
  mutate(
    REGION_N = map_chr(Description, extract_kml_field, "REGION_N"),
    PLN_AREA_N = map_chr(Description, extract_kml_field, "PLN_AREA_N"),
    SUBZONE_N = map_chr(Description, extract_kml_field, "SUBZONE_N"),
    SUBZONE_C = map_chr(Description, extract_kml_field, "SUBZONE_C")
  ) %>%
  select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())

mpsz_cl <- mpsz %>%
  filter(SUBZONE_N != "SOUTHERN GROUP",
         PLN_AREA_N != "WESTERN ISLANDS",
         PLN_AREA_N != "NORTH-EASTERN ISLANDS")
```

## Check data
```{r}
mpsz
```
```{r}
write_rds(mpsz_cl, "data/rds/mpsz_cl.rds")
```

:::
## 3 First-order Analysis
<summary for first-order analysis>

<visualisation of the spatial pattern>

:::panel-tabset
## Laundry Services
```{r}
#| fig-width: 12
#| fig-height: 8
par(bg = "#EEEEEE")
plot(st_geometry(mpsz_cl))
plot(st_geometry(biz_9601_sf), add = TRUE, col = "#476EAE", pch = 16, cex = 0.8)
```


## Beauty Services

```{r}
#| fig-width: 12
#| fig-height: 8
par(bg = "#EEEEEE")
plot(st_geometry(mpsz_cl))
plot(st_geometry(biz_9602_sf), add = TRUE, col = "#BA487F", pch = 16, cex = 0.5)
```


## Dental Services
```{r}
#| fig-width: 12
#| fig-height: 8
par(bg = "#EEEEEE")
plot(st_geometry(mpsz_cl))
plot(st_geometry(biz_86204_sf), add = TRUE, col = "#EF7722", pch = 16, cex = 0.8)
```

:::
### 3.1 Spatial KDE
<summary of spatial KDE>

#### 3.1.1 Geospatial data wrangling
In this section, we will convert the sf objects to spatstat ppp and owin object.

:::panel-tabset
## Laundry Services

**Convert sf data frames to ppp class**

```{r}
biz_9601_ppp <- as.ppp(biz_9601_sf)
```

```{r}
summary(biz_9601_ppp)
```


**Create Owin object**
```{r}
sg_owin <- as.owin(mpsz_cl)
```

**Combine point events object and owin object**
```{r}
biz_9601_ppp = biz_9601_ppp[sg_owin]
biz_9601_ppp
```

## Beauty Services
**Convert sf data frames to ppp class**
```{r}
biz_9602_ppp <- as.ppp(biz_9602_sf)
```


**Combine point events object and owin object**
```{r}
biz_9602_ppp = biz_9602_ppp[sg_owin]
biz_9602_ppp
```

## Dental Services
**Convert sf data frames to ppp class**
```{r}
biz_86204_ppp <- as.ppp(biz_86204_sf)
```


**Combine point events object and owin object**
```{r}
biz_86204_ppp = biz_86204_ppp[sg_owin]
biz_86204_ppp
```

:::
#### 3.1.2 Clark-Evans test
In this section, we will perform Clark-Evans tests for the 3 business entities separately, to quantify whether a point pattern is clustered, random, or uniformly spaced. The result will help us understand the underlying spatial patterns of each business type. This will inform our interpretation of the subsequent KDE analysis.

:::panel-tabset
## Laundry Services
The test hypotheses are:
Ho = The distribution of laundry services are randomly distributed.
H1 = The distribution of laundry services are not randomly distributed.
The 95% confident interval will be used.

```{r}
clarkevans.test(biz_9601_ppp,
                correction="none",
                clipregion="sg_owin",
                alternative=c("clustered"),
                method="MonteCarlo",
                nsim=999)
```
Laundry service R = 0.59189 shows there are some clustering in residential areas. It is more dispersed to serve neighbourhood needs, but it is not random.

## Beauty Services
The test hypotheses are:
Ho = The distribution of beauty services are randomly distributed.
H1 = The distribution of beauty services are not randomly distributed.
The 95% confident interval will be used.

```{r}
clarkevans.test(biz_9602_ppp,
                correction="none",
                clipregion="sg_owin",
                alternative=c("clustered"),
                method="MonteCarlo",
                nsim=999)
```
Beauty services (R = 0.36) indicates that they are most clustered. 

## Dental Services
The test hypotheses are:
Ho = The distribution of dental services are randomly distributed.
H1 = The distribution of dental services are not randomly distributed.
The 95% confident interval will be used.

```{r}
clarkevans.test(biz_86204_ppp,
                correction="none",
                clipregion="sg_owin",
                alternative=c("clustered"),
                method="MonteCarlo",
                nsim=999)
```
Dental service R = 0.49043 shows moderate clustering. This is possibily due to medical professional buildigs or proximity to hospitals. 
:::
#### 3.1.3 Compute KDE with adaptive/fixed bandwith

(maybe)The spatial patterns we observed in the plots above show that in the central areas there are higher density. 


Laundry Services
Beauty Services 
Dental Services 

### 3.2 Spatio-Temporal KDE
Laundry Services 
Beauty Services 
Dental Services 

#### 3.21 Visualise business entity points
Laundry Services 
Beauty Services 
Dental Services 

#### 3.2.2 Compute Spatio-Temporal STKDE by month
Laundry Services 
Beauty Services 
Dental Services 

## 4 Second-order Analysis
Laundry Services - F-function
Beauty Services - G-function, K-function
Dental Services - F-function

### 4.1 Spatial analysis
Laundry Services - F-function
Beauty Services - G-function, K-function
Dental Services - F-function

### 4.2 Spatio-Temporal analysis
Laundry Services 
H12024, H22024, H12025

Beauty Services - G-function, K-function
H12024, H22024, H12025

Dental Services - F-function
H12024, H22024, H12025



