---
title: "Take-home Exercise: Public Good"
author: "Cathy C."
date: "Sep 17, 2025"
date-modified: "last-modified" #auto render date
format: html
execute: #how YML page work
  echo: true #will show code
  eval: true #will run code every time and show output
  warning: false #dont show warning msg
  freeze: true #controls rendering - no render if no changes
  cache: true
---

## Introduction

In this study, we are interested in understanding the spatial and spatio-temporal patterns of new businesses and firms established in Singapore. Based on the Singapore Standard Industrial Classification 2020, there are over 1,500 business types. Among these, we will focus on **Laundry services** (including Laundry and dry cleaning services and self-operated laundries), **Beauty services** (including Hairdressing salons/shops, beauty salons and spas, as well as manicure and pedicure services, and **Dental services**.

The 3 chosen businesses are essential for working adults / professionals to take care of self for outlook and health. We would like to observe how these activities evolve and potentially to enhance quality of life and well-being.

We will start from processing the data, perform Kernel Density Estimation, Clarke-Evans test, and then conduct analysis with visualisation and test results. First-order and second-order spatial and spatio-temporal analyses are conducted to understand the overall new businesses and their temporal patterns during the period of time Jan 2024 to June 2025.

Due to data set sizes, we will analyse spatial patterns by quarter for Laundry Services and Dental Services, and by month for Beauty Services. Different spatial point pattern testing menthods are used to analyse the 3 business types: G-function for Dental Services, L-function for Beauty Services, and F-function for Laundry Services.

## 1 Loading the R package

We will use **tidyverse, sf, tmap, tmap,** and **httr** packages to process data sets, along with other packages for computing KDE and plotting visualisations.

| R Package | Purpose |
|------------------|------------------------------------------------------|
| **tidyverse** | the **tidyverse** package is designed to make it easy to install and load core packages from the tidyverse in a single command. There are ggplot2, dplyr, tidyr, readr, purrr, and tibble, etc to manipulate, tidy, import, visualise data. |
| **sf** | provides support for working with simple features, a standardized way to represent geographic vector data. It serves as a comprehensive tool for handling spatial data within the R environment. |
| **tmap** | serves the purpose of creating thematic maps, offering a flexible and intuitive way to visualize spatial data. It is designed to produce both static and interactive maps using a consistent, layer-based syntax similar to `ggplot2`. |
| **httr** | serves as a tool for interacting with web APIs and performing HTTP requests within the R environment. |
| **knitr** | seamlessly integrate R code, its output (including text, tables, and graphics), and narrative text within a single document. |
| **rvest** | simplify the process of scraping and manipulating web data. It allows for the extraction of information from websites and transforms it into a format suitable for analysis in R. |
| **spatstat** | provides functions for spatial statistics with a strong focus on analysing spatial point patterns. |
| **terra** | provides methods for manipulating geographic (spatial) data in both raster and vector formats. |
| **raster** | provides functions which reads, writes, manipulates, analyses and model of gridded spatial data (i.e. raster). In this hands-on exercise, it will be used to convert image output generate by spatstat into raster format. |
| **sparr** | provides a suite of functions for analyzing spatial and spatiotemporal point patterns, particularly focusing on the estimation and inference of relative risk surfaces. |

: R Packages

```{r}
pacman::p_load(tidyverse, sf, tmap, httr, knitr, rvest, spatstat, terra, raster, sparr)
```

## 2 The Data

There are mainly two raw public datasets used in this study.

-   **ACRA (Accounting and Corporate Regulatory Authority) Information on Corporate Entities datasets**: contains the business entities information, including their names, registration, status, and location.
-   **The Master Plan 2019 Boundary**: consists of the geographical boundary of Singapore at the planning subzone level. The data is based on URA Master Plan 2019.
-   **Singapore Residents by Planning Area / Subzone, Age Group, Sex and Type of Dwelling, June 2024**: *respopagesextod2024.csv* is an aspatial data fie. It can be downloaded at Department of Statistics, Singapore.

### 2.1 Importing ACRA data

We downloaded the ACRA (Accounting and Corporate Regulatory Authority) Information on Corporate Entities datasets from data.gov.sg, and will import the ACRA data and transform them as tibble frame.

```{r}
folder_path <- "data/aspatial/acra"
file_list <- list.files(path = folder_path,
                        pattern = "^ACRA*.*\\.csv$",
                        full.names = TRUE)
acra_data <- file_list %>%
  map_dfr(read_csv) #import and append all of them together!!
  
```

Then, we will write the data into .rds to store it in the data folder. This is so no need to process data every time.

```{r}
#| eval: false
write_rds(acra_data, "data/rds/acra_data.rds")
```

```{r}
acra_data <- read_rds("data/rds/acra_data.rds")
```

### 2.2 Tidying ACRA data

With this step, we will extract the business codes 9601, 9602, and 86204, then rename the columns and transform the data type necessarily.

The business types chosen are:

| Business Code | Types / subtypes |
|----------------------|--------------------------------------------------|
| **9601** | **Washing and (Dry-) Cleaning of Textile and Fur Products** |
| 96011 | Laundry and dry cleaning services except self-operated laundries |
| 96012 | Self-operated laundries |
| **9602** | **Hairdressing, Beauty and Other Personal Care Services** |
| 96021 | Hairdressing salons/shops (including barber shops) |
| 96022 | Beauty salons and spas (including slimming, skin care and hair care centres) |
| 96023 | Manicure and pedicure services |
| **86024** | **Dental services** |

: Business Code & Types

We use below code to retrieve the data between Jan 1, 2024 to June 30, 2025 for the 3 different major business types.

::: panel-tabset
## 9601

**Washing and (Dry-) Cleaning of Textile and Fur Products** Business Code 9601 encompasses Laundry cleaning and self-operated laundries (96011 & 96012)

```{r}
biz_9601 <- acra_data %>%
  dplyr::select(1:24) %>%
  filter(primary_ssic_code == 96011 | primary_ssic_code == 96012) %>%
  rename(date = registration_incorporation_date) %>%
  mutate(date = as.Date(date),
         YEAR = year(date),
         MONTH_NUM = month(date),
         MONTH_ABBR = month(date, label = TRUE, abbr = TRUE)) %>% #this gives Jan, Feb, Mar...
  mutate(postal_code = str_pad(postal_code,
                               width = 6, side = "left", pad = "0")) %>% #if missing digits, will replace with zero.
filter(YEAR == 2024 | YEAR == 2025 & MONTH_NUM <= 6)

glimpse(biz_9601)
```

**There are 150 observations we have retrieved for this analysis.**

## 9602

**Hairdressing, Beauty and Other Personal Care Services** Business Code 9602 includes hair dressing, beauty salons and manicure & pedicure services.

```{r}
biz_9602 <- acra_data %>%
  dplyr::select(1:24) %>%
  filter(primary_ssic_code == 96021 | primary_ssic_code == 96022 | primary_ssic_code == 96023) %>%
  rename(date = registration_incorporation_date) %>%
  mutate(date = as.Date(date),
         YEAR = year(date),
         MONTH_NUM = month(date),
         MONTH_ABBR = month(date, label = TRUE, abbr = TRUE)) %>% #this gives Jan, Feb, Mar...
  mutate(postal_code = str_pad(postal_code,
                               width = 6, side = "left", pad = "0")) %>% #if missing digits, will replace with zero.
filter(YEAR == 2024 | YEAR == 2025 & MONTH_NUM <= 6)

glimpse(biz_9602)
```

**There are 1762 observations we have retrieved for this analysis.**

## 86204

**Dental services**

```{r}
biz_86204 <- acra_data %>%
  dplyr::select(1:24) %>%
  filter(primary_ssic_code == 86204) %>%
  rename(date = registration_incorporation_date) %>%
  mutate(date = as.Date(date),
         YEAR = year(date),
         MONTH_NUM = month(date),
         MONTH_ABBR = month(date, label = TRUE, abbr = TRUE)) %>% #this gives Jan, Feb, Mar...
  mutate(postal_code = str_pad(postal_code,
                               width = 6, side = "left", pad = "0")) %>% #if missing digits, will replace with zero.
filter(YEAR == 2024 | YEAR == 2025 & MONTH_NUM <= 6)

glimpse(biz_86204)
```

**There are 81 observations we have retrieved for this analysis.**
:::

### 2.3 Geocoding

We will perform the following in this step: (Avoid doing so for \> 5K datasets)

-   Extract unique postal code from the poastal_code field of the tidied tibble data frame.
-   Perform geocode by submitting the postal code to SLA OneMap Reverse Geocode API.
-   Save the return results into 2 data frame namely *found* and *not_found*.

::: panel-tabset
## 9601

Below is to perform geocoding for the Laundry Services type.

```{r}

postcodes_9601 <- unique(biz_9601$postal_code)

url <- "https://onemap.gov.sg/api/common/elastic/search"

found_9601 <- data.frame()
not_found_9601 <- data.frame(postcode_9601 = character())

for (pc_9601 in postcodes_9601) {
  query_9601 <- list(
    searchVal = pc_9601,
    returnGeom = "Y",
    getAddrDetails = "Y",
    pageNum = "1"
  )
  
  res_9601 <- GET(url, query = query_9601)
  json_9601 <- content(res_9601)
  
  if(json_9601$found != 0) {
    df_9601 <- as.data.frame(json_9601$results, stringsAsFactors = FALSE)
    df_9601$input_postcode_9601 <- pc_9601
    found_9601 <- bind_rows(found_9601, df_9601)
  } else{
    not_found_9601 <- bind_rows(not_found_9601, data.frame(postcode_9601 = pc_9601))
  }
}
```

## 9602

Below is to perform geocoding for the Beauty Services type.

```{r}

postcodes_9602 <- unique(biz_9602$postal_code)

url <- "https://onemap.gov.sg/api/common/elastic/search"

found_9602 <- data.frame()
not_found_9602 <- data.frame(postcode_9602 = character())

for (pc_9602 in postcodes_9602) {
  query_9602 <- list(
    searchVal = pc_9602,
    returnGeom = "Y",
    getAddrDetails = "Y",
    pageNum = "1"
  )
  res_9602 <- GET(url, query = query_9602)
  json_9602 <- content(res_9602)
  
  if(json_9602$found != 0) {
    df_9602 <- as.data.frame(json_9602$results, stringsAsFactors = FALSE)
    df_9602$input_postcode_9602 <- pc_9602
    found_9602 <- bind_rows(found_9602, df_9602)
  } else{
    not_found_9602 <- bind_rows(not_found_9602, data.frame(postcode_9602 = pc_9602))
  }
}
```

## 86204

Below is to perform geocoding for the Dental Services type.

```{r}

postcodes_86204 <- unique(biz_86204$postal_code)

url <- "https://onemap.gov.sg/api/common/elastic/search"

found_86204 <- data.frame()
not_found_86204 <- data.frame(postcode_86204 = character())

for (pc_86204 in postcodes_86204) {
  query_86204 <- list(
    searchVal = pc_86204,
    returnGeom = "Y",
    getAddrDetails = "Y",
    pageNum = "1"
  )
  
  res_86204 <- GET(url, query = query_86204)
  json_86204 <- content(res_86204)
  
  if(json_86204$found != 0) {
    df_86204 <- as.data.frame(json_86204$results, stringsAsFactors = FALSE)
    df_86204$input_postcode_86204 <- pc_86204
    found_86204 <- bind_rows(found_86204, df_86204)
  } else{
    not_found_86204 <- bind_rows(not_found_86204, data.frame(postcode_86204 = pc_86204))
  }
}
```
:::

### 2.4 Appending the location information
In this step, we will append the location information to the business entities based on their postal code.

::: panel-tabset
## 9601
Below is to perform location appendix for the Laundry Services type.
```{r}

biz_9601 = biz_9601 %>%
  left_join(found_9601,
            by = c('postal_code' = 'POSTAL'))
```

## 9602
Below is to perform location appendix for the Beauty Services type.
```{r}

biz_9602 = biz_9602 %>%
  left_join(found_9602,
            by = c('postal_code' = 'POSTAL'))
```

## 86204
Below is to perform location appendix for the Dental Services type.
```{r}

biz_86204 = biz_86204 %>%
  left_join(found_86204,
            by = c('postal_code' = 'POSTAL'))
```
:::


### 2.5 Converting into SF data frame
Now we will convert the tibble data frame to a `sf` object. 
::: panel-tabset
## 9601
Below is to convert data type for the Laundry Services type.
```{r}

biz_9601_sf <- st_as_sf(biz_9601,
                         coords = c("X", "Y"),
                         crs = 3414)
```

## 9602
Below is to convert data type for the Beauty Services type.
```{r}

biz_9602_sf <- st_as_sf(biz_9602,
                         coords = c("X", "Y"),
                         crs = 3414)
```

## 86204
Below is to convert data type for the Dental Services type.
```{r}

biz_86204_sf <- st_as_sf(biz_86204,
                         coords = c("X", "Y"),
                         crs = 3414)
```
:::

**Take a look at the simple feature data (Laundry service).**

```{r}

kable(head(biz_9601_sf))
```

### 2.6 Import geospatial data into R

*The Master Plan 2019 Boundary (No Sea) (KML)* was downloaded from data.gov.sg that consists of the geographical boundary of Singapore at the planning subzone level. The data is based on URA Master Plan 2019. We will use this to plot the boundaries of the subzones.

::: panel-tabset
## Import data

```{r}
mpsz <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaKML.kml") %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 3414)
```

## Tidy data

```{r}
extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)
  rows <- page %>% html_elements("tr")
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>%
    html_element("td") %>%
    html_text2()
  
  if (length(value) == 0) NA_character_ else value
}

mpsz <- mpsz %>%
  mutate(
    REGION_N = map_chr(Description, extract_kml_field, "REGION_N"),
    PLN_AREA_N = map_chr(Description, extract_kml_field, "PLN_AREA_N"),
    SUBZONE_N = map_chr(Description, extract_kml_field, "SUBZONE_N"),
    SUBZONE_C = map_chr(Description, extract_kml_field, "SUBZONE_C")
  ) %>%
  dplyr::select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())

mpsz_cl <- mpsz %>%
  filter(SUBZONE_N != "SOUTHERN GROUP",
         PLN_AREA_N != "WESTERN ISLANDS",
         PLN_AREA_N != "NORTH-EASTERN ISLANDS")
```

## Check data

```{r}
mpsz_cl
```

```{r}
#| eval: false
write_rds(mpsz_cl, "data/rds/mpsz_cl.rds")
```
:::


### 2.7 Prepare population data

:::panel-tabset
## Import data
```{r}
popdata <- read_csv("data/aspatial/respopagesextod2024.csv")
kable(head(popdata))
```

## Data wrangling

```{r}
popdata2024 <- popdata %>%
  group_by(PA, SZ) %>%
  summarise(TOTAL_POP = sum(`Pop`)) %>%
  ungroup() 

kable(head(popdata2024))

```


## Join population and geospatial data

Use left_join to join the population data and geospatial data.
```{r}
popdata2024 <- popdata2024 %>%
  mutate_at(.vars = vars(PA, SZ),
            .funs = list(toupper)) %>%
  filter(`TOTAL_POP` > 0)
mpsz_pop2024 <- left_join(mpsz_cl, popdata2024,
                          by = c("SUBZONE_N" = "SZ"))
```

Then write it into a rds file.
```{r}
#| eval: false
write_rds(mpsz_pop2024, "data/rds/mpsz_pop2024.rds")
```


We will call it from rds file to save computation.
```{r}
mpsz_pop2024 <- read_rds("data/rds/mpsz_pop2024.rds")
```


:::

## 3 First-order Analysis

<summary for first-order analysis>


::: panel-tabset
## Laundry Services
### Mapping geospatial data sets

```{r}
#| fig-width: 12
#| fig-height: 8
#| fig-align: center
par(bg = "#EEEEEE")
plot(st_geometry(mpsz_cl), border = "grey70")
plot(st_geometry(biz_9601_sf), add = TRUE, col = "#FF9800", pch = 16, cex = 0.8)
```

### Mapping with population
```{r}
#| fig-width: 12
#| fig-height: 6
#| fig-align: center

par(bg = "#EEEEEE")
tmap_mode("plot")

tm_shape(mpsz_pop2024) +
  tm_polygons(col = "TOTAL_POP", 
          border.col = "#7077A1", lwd = 0.3) +
tm_shape(biz_9601_sf) +
  tm_dots(col = "#FF9800", size = 0.3, alpha = 0.9) +
tm_layout(bg.color =  "#EEEEEE",
          outer.bg.color = NA,
          frame = FALSE,
          legend.bg.color = "transparent",      
          legend.frame = FALSE)

```
::::: notegoals
::: noterkgoals-header
**Insight**
:::
::: notegoals-container
New laundry businesses show a clear spatial pattern that aligns with both residential density and commercial activity. 
**Residential Area Patterns**: Laundry businesses concentrate in high-density residential areas. Areas with lower population density show correspondingly fewer or no new laundry establishments, which suggests a strong relationship between residential demand and business location decisions.

**Commercial District Concentration**: The Central Business District (CBD) exhibits high concentrations of new laundry businesses. It likely serves the large working population who require convenient dry cleaning services near their workplaces.

**Spatial Distribution**: This dual pattern - clustering in both high-density residential areas and the CBD, indicates that laundry services strategically locate to serve both residential neighborhoods and working professionals.


:::
:::::


## Beauty Services
### Mapping geospatial data sets
```{r}
#| fig-width: 12
#| fig-height: 8
#| fig-align: center
par(bg = "#EEEEEE")
plot(st_geometry(mpsz_cl), border = "grey70")
plot(st_geometry(biz_9602_sf), add = TRUE, col = "#BA487F", pch = 16, cex = 0.5)
```


### Mapping with population
```{r}
#| fig-width: 12
#| fig-height: 6
#| fig-align: center

par(bg = "#EEEEEE")
tmap_mode("plot")

tm_shape(mpsz_pop2024) +
  tm_polygons(col = "TOTAL_POP", 
          border.col = "#7077A1", lwd = 0.3) +
tm_shape(biz_9602_sf) +
  tm_dots(col = "#BA487F", size = 0.2, alpha = 0.9) +
tm_layout(bg.color =  "#EEEEEE",
          outer.bg.color = NA,
          frame = FALSE,
          legend.bg.color = "transparent",      
          legend.frame = FALSE)

```

::::: notegoals
::: noterkgoals-header
**Insight**
:::

::: notegoals-container

New beauty service establishments show a spatial pattern similar to laundry services, but with significantly higher overall density concentrations.

**Residential Area Distribution**: In residential neighborhoods, beauty services show a more dispersed pattern compared to the central areas. While the density of new establishments correlates positively with residential population levels, areas with low population still have few business providers. Even for non-residential areas in the west, there are also new businesses there, likely to serve working professional or representing locations of chain store offices.

**Commercial Center Concentration**: Beauty services demonstrate particularly strong density in central south areas, likely targeting areas with high concentrations of commercial buildings and shopping centers. This concentration is even more evident than that observed for laundry services.
 
:::
:::::

## Dental Services
### Mapping geospatial data sets
```{r}
#| fig-width: 12
#| fig-height: 8
#| fig-align: center
par(bg = "#EEEEEE")
plot(st_geometry(mpsz_cl), border = "grey70")
plot(st_geometry(biz_86204_sf), add = TRUE, col = "#18978F", pch = 16, cex = 0.8)
```


### Mapping with population
```{r}
#| fig-width: 12
#| fig-height: 6
#| fig-align: center

par(bg = "#EEEEEE")
tmap_mode("plot")

tm_shape(mpsz_pop2024) +
  tm_polygons(col = "TOTAL_POP", 
          border.col = "#7077A1", lwd = 0.3) +
tm_shape(biz_86204_sf) +
  tm_dots(col = "#18978F             4xddddddddddddddddddddddddddddd", size = 0.3, alpha = 0.9) +
tm_layout(bg.color =  "#EEEEEE",
          outer.bg.color = NA,
          frame = FALSE,
          legend.bg.color = "transparent",      
          legend.frame = FALSE)

```

::::: notegoals
::: noterkgoals-header
**Insight**
:::

::: notegoals-container
Dental service establishments display a distinctly different spatial pattern compared to laundry and beauty services and show a more selective location strategy.

**Selective Geographic Distribution**: Unlike the other 2 business types, dental services do not establish in every subzone. Instead, they cluster in specific subzones. They also have a clear preference for central locations over residential areas.

**Central Location Dominance**: The majority of new dental services concentrate in central Singapore, significantly outnumbering those in residential areas. This pattern suggests that dental practices strategically locate in medical or commercial buildingswhere they can serve a broader catchment area.

**Limited Residential Presence**: While some new dental services do establish in residential neighborhoods, these are relatively few and appear in select subzones only. This selective distribution indicates that residential areas are not the preferred location type for new dental businesses. This is likely due to different service requirements and patient accessibility needs compared to laundry and beauty services.
:::
:::::

:::

### 3.1 Spatial KDE

<summary of spatial KDE>

#### 3.1.1 Geospatial data wrangling

In this section, we will convert the sf objects to spatstat ppp and owin object.

::: panel-tabset
## Laundry Services

**Convert sf data frames to ppp class**

```{r}
biz_9601_ppp <- as.ppp(biz_9601_sf)
```

```{r}
summary(biz_9601_ppp)
```

**Create Owin object**

```{r}
sg_owin <- as.owin(mpsz_cl)
```

**Combine point events object and owin object**

```{r}
biz_9601_ppp = biz_9601_ppp[sg_owin]
biz_9601_ppp
```

## Beauty Services

**Convert sf data frames to ppp class**

```{r}
biz_9602_ppp <- as.ppp(biz_9602_sf)
```

**Combine point events object and owin object**

```{r}
biz_9602_ppp = biz_9602_ppp[sg_owin]
biz_9602_ppp
```

## Dental Services

**Convert sf data frames to ppp class**

```{r}
biz_86204_ppp <- as.ppp(biz_86204_sf)
```

**Combine point events object and owin object**

```{r}
biz_86204_ppp = biz_86204_ppp[sg_owin]
biz_86204_ppp
```
:::

#### 3.1.2 Clark-Evans test

In this section, we will perform Clark-Evans tests for the 3 business entities separately, to quantify whether a point pattern is clustered, random, or uniformly spaced. The result will help us understand the underlying spatial patterns of each business type. This will inform our interpretation of the subsequent KDE analysis.

::: panel-tabset
## Laundry Services

The test hypotheses are: 
Ho = The distribution of laundry services are randomly distributed. 
H1 = The distribution of laundry services are not randomly distributed. 
The 95% confident interval will be used.

```{r}
clarkevans.test(biz_9601_ppp,
                correction="none",
                clipregion="sg_owin",
                alternative=c("clustered"),
                method="MonteCarlo",
                nsim=99)
```

::::: notegoals
::: noterkgoals-header
**Insight**
:::

::: notegoals-container
Laundry service R = 0.59189 shows there are some clustering in residential areas. It is more dispersed to serve neighbourhood needs, but it is not random.
:::
:::::

## Beauty Services

The test hypotheses are: Ho = The distribution of beauty services are randomly distributed. H1 = The distribution of beauty services are not randomly distributed. The 95% confident interval will be used.

```{r}
clarkevans.test(biz_9602_ppp,
                correction="none",
                clipregion="sg_owin",
                alternative=c("clustered"),
                method="MonteCarlo",
                nsim=99)
```

::::: notegoals
::: noterkgoals-header
**Insight**
:::

::: notegoals-container
Beauty services (R = 0.36) indicates that they are most clustered.
:::
:::::

## Dental Services

The test hypotheses are:

Ho = The distribution of dental services are randomly distributed.

H1 = The distribution of dental services are not randomly distributed. The 95% confident interval will be used.

```{r}
clarkevans.test(biz_86204_ppp,
                correction="none",
                clipregion="sg_owin",
                alternative=c("clustered"),
                method="MonteCarlo",
                nsim=99)
```

::::: notegoals
::: noterkgoals-header
**Insight**
:::

::: notegoals-container
Dental service R = 0.49043 shows moderate clustering. This is possibily due to medical professional buildigs or proximity to hospitals.
:::
:::::
:::

#### 3.1.3 Compute KDE with adaptive bandwidth

We will compute Kernel Density Estimation (KDE) to visualise and analyse first-order spatial point patterns. We use adaptive bandwidth because all 3 business types show significant clustering with varying intensities, and adaptive approach handles multi-scale density variations beter than fixed bandwidth by using finer resolution in high-density areas and broader smoothing in sparse regions.

::: panel-tabset
## Laundry Services

```{r}
par(bg = "#EEEEEE")
biz_9601_ppp_km <- rescale.ppp(biz_9601_ppp, 1000, "km")
kde_biz_9601_ab <- adaptive.density(
  biz_9601_ppp_km, method = "kernel"
)
plot(kde_biz_9601_ab,
     col = terrain.colors(256)) 
```

## Beauty Services

```{r}
par(bg = "#EEEEEE")
biz_9602_ppp_km <- rescale.ppp(biz_9602_ppp, 1000, "km")
kde_biz_9602_ab <- adaptive.density(
  biz_9602_ppp_km, method = "kernel"
)
plot(kde_biz_9602_ab,
     col = terrain.colors(256))
```

## Dental Services

```{r}
par(bg = "#EEEEEE")
biz_86204_ppp_km <- rescale.ppp(biz_86204_ppp, 1000, "km")
kde_biz_86204_ab <- adaptive.density(
  biz_86204_ppp_km, method = "kernel"
)
plot(kde_biz_86204_ab,
     col = terrain.colors(256))
```
:::

### 3.2 Spatio-Temporal KDE

Jaundry, Beauty, and Dental business types have 150, 1762 and 81 points respectively during Jan 2024 - June 2025. In this study, we will slice the data compute quarterly spatio-temporal KDE for analysis to see seasonal patterns.

#### 3.2.1 Visualise business entity points by month/quarter

::: panel-tabset
## Laundry Services

```{r}
#| fig-width: 12
#| fig-height: 12

biz_9601_quarter <- biz_9601_sf %>%
  mutate(
    quarter = case_when(
      MONTH_NUM %in% c(1, 2, 3) ~ "Q1",
      MONTH_NUM %in% c(4, 5, 6) ~ "Q2",
      MONTH_NUM %in% c(7, 8, 9) ~ "Q3",
      MONTH_NUM %in% c(10, 11, 12) ~ "Q4"
    ),
    year_quarter = paste(YEAR, quarter, sep = "-")
  )

tmap_options(bg.color = "#EEEEEE")
tm_shape(mpsz_cl) + 
  tm_polygons(fill = "#FFF2F2", col = "grey80") +
tm_shape(biz_9601_quarter) +
  tm_dots(fill = "#476EAE", size = 0.4) +
  tm_facets(by = "year_quarter", free.coords = FALSE, drop.units = TRUE,
            ncol = 2) +
tm_layout(
    main.title = "Laundry Services - Quarterly Density",
    bg.color = "#EED3D9")
```

::::: notegoals
::: noterkgoals-header
**Insight**
:::

::: notegoals-container

:::
:::::

## Beauty Services

```{r}
#| fig-width: 12
#| fig-height: 32

biz_9602_my <- biz_9602_sf %>%
  mutate(
    date = as.Date(paste(YEAR, MONTH_NUM, "01", sep = "-")),
    month_year = format(date, "%Y-%m"),
    month_year_label = format(date, "%b %Y")
  )

tmap_options(bg.color = "#EEEEEE")
tm_shape(mpsz_cl) + 
  tm_polygons(fill = "#FFF2F2", col = "grey80") +
  tm_shape(biz_9602_my) +
  tm_dots(fill = "#BA487F", size = 0.2) +
  tm_facets(by = "month_year", free.coords = FALSE, drop.units = TRUE,
            ncol = 2) +
tm_layout(
    main.title = "Beauty Services - Monthly Density",
    bg.color = "#CADABF")
```

::::: notegoals
::: noterkgoals-header
**Insight**
:::

::: notegoals-container

:::
:::::

## Dental Services

```{r}
#| fig-width: 12
#| fig-height: 12

biz_86204_quarter <- biz_86204_sf %>%
  mutate(
    quarter = case_when(
      MONTH_NUM %in% c(1, 2, 3) ~ "Q1",
      MONTH_NUM %in% c(4, 5, 6) ~ "Q2",
      MONTH_NUM %in% c(7, 8, 9) ~ "Q3",
      MONTH_NUM %in% c(10, 11, 12) ~ "Q4"
    ),
    year_quarter = paste(YEAR, quarter, sep = "-")
  )

tmap_options(bg.color = "#EEEEEE")
tm_shape(mpsz_cl) + 
  tm_polygons(fill = "#FFF2F2", col = "grey80") +
  tm_shape(biz_86204_quarter) +
  tm_dots(fill = "#EF7722", size = 0.4) +
  tm_facets(by = "year_quarter", free.coords = FALSE, drop.units = TRUE,
            ncol = 2) +
tm_layout(
    main.title = "Dental Services - Quarterly Density",
    bg.color = "#B5C0D0")
```

::::: notegoals
::: noterkgoals-header
**Insight**
:::

::: notegoals-container

:::
:::::

:::

#### 3.2.2 Compute Spatio-Temporal STKDE by month/quarter

::: panel-tabset
## Laundry Services

```{r}
# extract laundry data by quarter
kde_biz_9601_quarter <- biz_9601_quarter %>%
  dplyr::select(year_quarter)

# create ppp
biz_9601_quarter_ppp <- as.ppp(kde_biz_9601_quarter)
  

# include owin object
biz_9601_quarter_owin <- biz_9601_quarter_ppp[sg_owin]

```

```{r}
par(bg = "#EEEEEE")
plot(biz_9601_quarter_owin)
```

### Compute STKDE

```{r}
#convert year_quarter to numeric
marks(biz_9601_quarter_owin) <- case_when(
  marks(biz_9601_quarter_owin) == "2024-Q1" ~ 1,
  marks(biz_9601_quarter_owin) == "2024-Q2" ~ 2,
  marks(biz_9601_quarter_owin) == "2024-Q3" ~ 3,
  marks(biz_9601_quarter_owin) == "2024-Q4" ~ 4,
  marks(biz_9601_quarter_owin) == "2025-Q1" ~ 5,
  marks(biz_9601_quarter_owin) == "2025-Q2" ~ 6
)

st_kde_9601 <- spattemp.density(biz_9601_quarter_owin)
summary(st_kde_9601)
```

### Plot STKDE object

```{r}
#| fig-width: 12
#| fig-height: 8
tims <- c(1,2,3,4,5,6)
par(mfrow = c(2,3), bg = "#EEEEEE")
for(i in tims){
  plot(st_kde_9601, i,
       override.par = FALSE,
       fix.range = TRUE, main = paste("Laundry Services KDE at Quarter", i
     ), col = terrain.colors(256))
}

```

::::: notegoals
::: noterkgoals-header
**Insight**
:::

::: notegoals-container

:::
:::::

## Beauty Services

```{r}
# extract laundry data by quarter
kde_biz_9602_month <- biz_9602_my %>%
  dplyr::select(month_year)

# create ppp
biz_9602_month_ppp <- as.ppp(kde_biz_9602_month)
  

# include owin object
biz_9602_month_owin <- biz_9602_month_ppp[sg_owin]

```

```{r}
par(bg = "#EEEEEE")
plot(biz_9602_month_owin)
```

### Compute STKDE

```{r}

#convert year_quarter to numeric
marks(biz_9602_month_owin) <- case_when(
  marks(biz_9602_month_owin) == "2024-01" ~ 1,
  marks(biz_9602_month_owin) == "2024-02" ~ 2,
  marks(biz_9602_month_owin) == "2024-03" ~ 3,
  marks(biz_9602_month_owin) == "2024-04" ~ 4,
  marks(biz_9602_month_owin) == "2024-05" ~ 5,
  marks(biz_9602_month_owin) == "2024-06" ~ 6,
  marks(biz_9602_month_owin) == "2024-07" ~ 7,
  marks(biz_9602_month_owin) == "2024-08" ~ 8,
  marks(biz_9602_month_owin) == "2024-09" ~ 9,
  marks(biz_9602_month_owin) == "2024-10" ~ 10,
  marks(biz_9602_month_owin) == "2024-11" ~ 11,
  marks(biz_9602_month_owin) == "2024-12" ~ 12,  
  marks(biz_9602_month_owin) == "2025-01" ~ 13,
  marks(biz_9602_month_owin) == "2025-02" ~ 14,
  marks(biz_9602_month_owin) == "2025-03" ~ 15,
  marks(biz_9602_month_owin) == "2025-04" ~ 16,
  marks(biz_9602_month_owin) == "2025-05" ~ 17,
  marks(biz_9602_month_owin) == "2025-06" ~ 18
  )

st_kde_9602 <- spattemp.density(biz_9602_month_owin)
summary(st_kde_9602)
```

### Plot STKDE object

```{r}
#| fig-width: 12
#| fig-height: 36

tims <- c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18)
par(mfrow = c(9,2), bg = "#EEEEEE")
for(i in tims){
  plot(st_kde_9602, i,
       override.par = FALSE,
       fix.range = TRUE, main = paste("Beauty Services KDE at Month", i),
       col = terrain.colors(256))
}

```

::::: notegoals
::: noterkgoals-header
**Insight**
:::

::: notegoals-container

:::
:::::

## Dental Services

```{r}
# extract laundry data by quarter
kde_biz_86204_quarter <- biz_86204_quarter %>%
  dplyr::select(year_quarter)

# create ppp
biz_86204_quarter_ppp <- as.ppp(kde_biz_86204_quarter)
  

# include owin object
biz_86204_quarter_owin <- biz_86204_quarter_ppp[sg_owin]

```

```{r}
par(bg = "#EEEEEE")
plot(biz_86204_quarter_owin)
```

### Compute STKDE

```{r}
#convert year_quarter to numeric
marks(biz_86204_quarter_owin) <- case_when(
  marks(biz_86204_quarter_owin) == "2024-Q1" ~ 1,
  marks(biz_86204_quarter_owin) == "2024-Q2" ~ 2,
  marks(biz_86204_quarter_owin) == "2024-Q3" ~ 3,
  marks(biz_86204_quarter_owin) == "2024-Q4" ~ 4,
  marks(biz_86204_quarter_owin) == "2025-Q1" ~ 5,
  marks(biz_86204_quarter_owin) == "2025-Q2" ~ 6
)

st_kde_86204 <- spattemp.density(biz_86204_quarter_owin)
summary(st_kde_86204)
```

### Plot STKDE object

```{r}
#| fig-width: 12
#| fig-height: 8
tims <- c(1,2,3,4,5,6)
par(mfrow = c(2,3), bg = "#EEEEEE")
for(i in tims){
  plot(st_kde_86204, i,
       override.par = FALSE,
       fix.range = TRUE, main = paste("Dental Services KDE at Quarter", i),
       col = terrain.colors(256))
}

```
::::: notegoals
::: noterkgoals-header
**Insight**
:::

::: notegoals-container

:::
:::::

:::

## 4 Second-order Analysis

<overview>

### 4.1 Spatial analysis

In this section, we will compute spatial point patterns

::: panel-tabset
## Laundry Services

For laundry services, we will use F-function to analyse spatial point process as this function is best for random locations to nearest business. Laundry services may be important to residents, be it dry cleaning or self-operated. We want to observe its distribution whether the placement is random or follows specific spatial strategy.

```{r}
set.seed(1222)
```

### F-function estimation

```{r}
par(bg = "#EEEEEE")
F_9601 = Fest(biz_9601_ppp)
plot(F_9601)
```

### Complete spatial randomness test

We use the hypothesis test below to confirm the spatial patterns above. Ho = The distribution of laundry services are randomly distributed. H1= The distribution of laundry services are not randomly distributed. The null hypothesis will be rejected is p-value is smaller than alpha value of 0.05.

```{r}
F_9601.csr <- envelope(biz_9601_ppp, Fest, correction = "all", nsim = 99)
```

### Plot simulation

```{r}
par(bg = "#EEEEEE")
plot(F_9601.csr)
```

::::: notegoals
::: noterkgoals-header
**Insight**
:::

::: notegoals-container

:::
:::::
<under envelope>

## Beauty Services

For beauty services such as hair salons, manicure and pedicure, personal care services, we will use K-function to analyze clustering patterns by examining the expected number of neighboring businesses within given distances. The intention is to identify and quantify clustering at multiple spatial scales.

### L-function estimate

```{r}
L_9602 = Lest(biz_9602_ppp, correction = "Ripley")
plot(L_9602, . -r ~ r, ylab = "L(d)-r", xlab = "d(m)")
```

### Spatial randomness test

The hypothesis and test are as follows:

Ho = The distribution of Beauty services are randomly distributed.

H1= The distribution of Beauty services are not randomly distributed.

The null hypothesis will be rejected if p-value is smaller than alpha value of 0.05.

```{r}
L_9602.csr <- envelope(biz_9602_ppp, Lest, nsim = 39, rank = 1)
```

### Plot L-function test

```{r}
plot(L_9602.csr, . - r ~ r, xlab = "d", ylab = "L(d)-r")
```

::::: notegoals
::: noterkgoals-header
**Insight**
:::

::: notegoals-container

:::
:::::

## Dental Services - G-function

### G-function estimate

```{r}
G_86204 = Lest(biz_86204_ppp, correction = "border")
plot(G_86204, xlim = c(0, 500))
```

### Spatial randomness test

The hypothesis and test are as follows:

Ho = The distribution of Beauty services are randomly distributed. H1= The distribution of Beauty services are not randomly distributed. The null hypothesis will be rejected if p-value is smaller than alpha value of 0.05.

```{r}
G_86204.csr <- envelope(biz_86204_ppp, Gest, nsim = 99)
```

### Plot G-function test

```{r}
plot(G_86204.csr)
```

::::: notegoals
::: noterkgoals-header
**Insight**
:::

::: notegoals-container

:::
:::::
:::

### 4.2 Spatio-Temporal analysis

::: panel-tabset
## Laundry Services

### Perform F-function estimation

```{r}
#| fig-width: 12
#| fig-height: 12
biz_9601_quarter_split <- split(biz_9601_quarter_owin, f = as.factor(marks(biz_9601_quarter_owin)))

F_9601_quarter <- lapply(biz_9601_quarter_split, Fest)

par(mfrow = c(3,2), bg = "#EEEEEE")
biz_9601_quarter_names <- c("2024-Q1", "2024-Q2", "2024-Q3", "2024-Q4", "2025-Q1", "2025-Q2")

for (i in 1:6) {
  plot(F_9601_quarter[[i]],
       main = paste("Laundry Services F-function -", biz_9601_quarter_names[i]))
}


```

### Perform spatial randomness test for each quarter

Hypothesis test: H₀ = The distribution of laundry services is randomly distributed. H₁= The distribution of laundry services is not randomly distributed.

Alpha value = 0.05.

```{r}
F_9601_quarter.csr <- lapply(biz_9601_quarter_split, function(x) envelope(x, Fest, nsim = 99))
```

### Plot result of test

```{r}
#| fig-width: 12
#| fig-height: 12
par(mfrow = c(3, 2), bg = "#EEEEEE")
biz_9601_quarter_labels <- c(paste0("2024-Q", 1:4), paste0("2025-Q", 1:2))

for (i in 1:6) {
  plot(F_9601_quarter.csr[[i]], main = paste("Laundry F-function:", biz_9601_quarter_labels[i]))
}
```

::::: notegoals
::: noterkgoals-header
**Insight**
:::

::: notegoals-container

:::
:::::

## Beauty Services

### Perform L-function estimation by month

```{r}
#| fig-width: 12
#| fig-height: 32
biz_9602_month_split <- split(biz_9602_month_owin, f = as.factor(marks(biz_9602_month_owin)))

L_9602_month <- lapply(biz_9602_month_split, Lest)

par(mfcol = c(9,2), bg = "#EEEEEE")
biz_9602_month_names <- c("2024-1", "2024-2", "2024-3", "2024-4", "2024-5","2024-6","2024-7","2024-8","2024-9","2024-10","2024-11","2024-12", "2025-1", "2025-2", "2025-3", "2025-4", "2025-5", "2025-6")

for (i in 1:18) {
  plot(L_9602_month[[i]],
       main = paste("Beauty Services L-function -", biz_9602_month_names[i]))
}


```

### Perform spatial randomness test for each quarter

Hypothesis test: H₀ = The distribution of laundry services is randomly distributed. H₁= The distribution of laundry services is not randomly distributed.

Alpha value = 0.05.

```{r}
L_9602_month.csr <- lapply(biz_9602_month_split, function(x) envelope(x, Fest, nsim = 99))
```

### Plot result of test

```{r}
#| fig-width: 12
#| fig-height: 36

par(mfrow = c(9, 2), bg = "#EEEEEE")
biz_9602_month_labels <- c(paste0("2024-", 1:12), paste0("2025-", 1:6))

for (i in 1:18) {
  plot(L_9602_month.csr[[i]], main = paste("Beauty L-function:", biz_9602_month_labels[i]))
}
```

::::: notegoals
::: noterkgoals-header
**Insight**
:::

::: notegoals-container

:::
:::::

## Dental Services

### Perform G-function estimation by quarter

```{r}
#| fig-width: 12
#| fig-height: 12
biz_86204_quarter_split <- split(biz_86204_quarter_owin, f = as.factor(marks(biz_86204_quarter_owin)))

G_86204_quarter <- lapply(biz_86204_quarter_split, Gest)

par(mfrow = c(3,2), bg = "#EEEEEE")
biz_86204_quarter_names <- c("2024-Q1", "2024-Q2", "2024-Q3", "2024-Q4", "2025-Q1", "2025-Q2", "2025-Q3", "2025-Q4")

for (i in 1:6) {
  plot(G_86204_quarter[[i]], xlim = c(0, 3000),
       main = paste("Dental Services G-function -", biz_86204_quarter_names[i]))
}


```

### Perform spatial randomness test for each quarter

Hypothesis test: H₀ = The distribution of laundry services is randomly distributed. H₁= The distribution of laundry services is not randomly distributed.

Alpha value = 0.05.

```{r}
G_86204_quarter.csr <- lapply(biz_86204_quarter_split, function(x) envelope(x, Gest, nsim = 99))
```

### Plot result of test

```{r}
#| fig-width: 12
#| fig-height: 12
par(mfrow = c(3, 2), bg = "#EEEEEE")
biz_86204_quarter_labels <- c(paste0("2024-Q", 1:4), paste0("2025-Q", 1:2))

for (i in 1:6) {
  plot(G_86204_quarter.csr[[i]], xlim = c(0,3000), main = paste("Dental G-function:", biz_86204_quarter_labels[i]))
}
```

::::: notegoals
::: noterkgoals-header
**Insight**
:::

::: notegoals-container

:::
:::::

:::

## 5 Conclusion

discuss how these findings can be used to support urban land use planning and management. (Not more than 500 words)

### 5.1 Key findings

### 5.2 Limitation

### 5.3 Future works

## 6 References
